---
title: "Insula laterality (INS_HE merged v1+v2)"
subtitle: "Full v2 pipeline on corrected INS table — same notebook as laterality_ver2.Rmd, standalone"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    fig_width: 12
    fig_height: 8
---

This is the **full** laterality analysis in one file (no child document). It matches **`laterality_ver2.Rmd`**: SQ1–SQ6 structure from **`laterality_by_questions.Rmd`**, plus adjusted PERMANOVA, balanced Monte Carlo PERMANOVA, and `lm(log1p(total) ~ Neuron_Type + Soma_Side)`.

**Default input** is the corrected workbook **`251637_INS_HE_inferred.xlsx`**. Edit **`PATH_INS_XLSX`** in `config` if you use `251637_INS_HE.xlsx` or `251637_INS.xlsx` instead.

**Prelude (table changes in inferred INS_HE):** verified soma locations; rule-based fill for previously unknown side/type; some PrCO neurons reclassified into insular groups; `Soma_Region` uses `CL_`/`CR_` stripped in prep with side in **`Soma_Side`**.

**Maintenance:** Body matches **`laterality_ver2.Rmd`** as of the last copy. If you change v2 logic, update this file the same way (or copy `laterality_ver2.Rmd` over this file and re-apply the header + `config` chunk above).

```{r config, include=TRUE}
# --- Paths (relative to this Rmd: R_analysis/scripts/) ---
PATH_INS_XLSX   <- "../tables/251637_INS_HE_inferred.xlsx"
PATH_CLUSTER_RDS <- "data_output/clustering/clustering_results.rds"
PATH_FNT_XLSX    <- "data_output/clustering/raw_matrix_神经元矩阵.xlsx"
FIGURE_DIR       <- "figures/LR_analysis"
stopifnot(file.exists(PATH_INS_XLSX))
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

PERM_N <- 999L
BALANCE_REPS           <- 299L
BALANCE_MIN_PER_SIDE   <- 3L
RUN_BALANCE_PERMANOVA  <- TRUE
RUN_BALANCE_BRAY_PROP  <- FALSE

set.seed(42)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = TRUE,
  fig.align = "center",
  fig.path = paste0(FIGURE_DIR, "/laterality_")
)
library(readxl)
library(dplyr)
library(ggplot2)
library(vegan)
library(ComplexHeatmap)
library(circlize)
library(ggpubr)
library(scales)
library(patchwork)

save_plot <- function(plot_obj, filename, width, height, dpi = 300) {
  out_path <- file.path(FIGURE_DIR, filename)
  ggplot2::ggsave(
    filename = out_path,
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    units = "in"
  )
  message("Saved figure: ", out_path)
  invisible(out_path)
}

cleanup_vars <- function(vars) {
  to_drop <- intersect(vars, ls(envir = .GlobalEnv))
  if (length(to_drop)) {
    rm(list = to_drop, envir = .GlobalEnv)
  }
  invisible(gc())
}

# Automatic significance labeling used across SQ1/SQ2 tables
p_to_sig <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}
```

------------------------------------------------------------------------

# Preparation: align projection matrices with neuron metadata

```{r prep}
cluster_info <- readRDS(PATH_CLUSTER_RDS)
proj_ipsi    <- read_excel(PATH_INS_XLSX, sheet = "Projection_Strength_L3_ipsi")
proj_contra  <- read_excel(PATH_INS_XLSX, sheet = "Projection_Strength_L3_contra")
summary_sh   <- read_excel(PATH_INS_XLSX, sheet = "Summary")

# Normalize region labels to hemisphere-neutral form.
# Handles prefixes like CL_/CR_, L_/R_, L-/R-; side remains in Soma_Side.
.strip_soma_region_prefix <- function(x) {
  v <- trimws(as.character(x))
  w <- sub("^(CL|CR)[_\\-]", "", v, perl = TRUE)
  w <- sub("^[LR][_\\-]", "", w, perl = TRUE)
  ifelse(nzchar(w), w, v)
}

.pick_first_col <- function(df, candidates) {
  hit <- candidates[candidates %in% colnames(df)][1]
  if (is.na(hit)) NA_character_ else hit
}

.ap_col_meta <- .pick_first_col(summary_sh, c(
  "Soma_NII_Y", "Soma_Phys_Y",
  "AP", "ap", "Soma_AP", "soma_ap", "Y", "y", "Soma_Y", "soma_y"
))
.ml_col_meta <- .pick_first_col(summary_sh, c(
  "Soma_NII_X", "Soma_Phys_X",
  "ML", "ml", "Soma_ML", "soma_ml", "X", "x", "Soma_X", "soma_x"
))

summary_coords <- summary_sh %>%
  dplyr::select(NeuronID, everything()) %>%
  transmute(
    NeuronID = NeuronID,
    AP_coord = if (!is.na(.ap_col_meta)) as.numeric(.data[[.ap_col_meta]]) else NA_real_,
    ML_coord = if (!is.na(.ml_col_meta)) as.numeric(.data[[.ml_col_meta]]) else NA_real_
  )

meta_cols <- 6L

proj_ipsi <- proj_ipsi %>%
  merge(cluster_info$cluster_df %>% dplyr::select(-BioType), by = "NeuronID") %>%
  merge(summary_sh %>% dplyr::select(NeuronID, Soma_Region, Soma_Side), by = "NeuronID") %>%
  relocate(NeuronID, Soma_Region, Soma_Side, Neuron_Type, Cluster, DendroOrder) %>%
  mutate(Soma_Region = factor(.strip_soma_region_prefix(Soma_Region)))

proj_contra <- proj_contra %>%
  merge(cluster_info$cluster_df %>% dplyr::select(-BioType), by = "NeuronID") %>%
  merge(summary_sh %>% dplyr::select(NeuronID, Soma_Region, Soma_Side), by = "NeuronID") %>%
  relocate(NeuronID, Soma_Region, Soma_Side, Neuron_Type, Cluster, DendroOrder) %>%
  mutate(Soma_Region = factor(.strip_soma_region_prefix(Soma_Region)))

mat_ipsi <- t(as.matrix(proj_ipsi %>% dplyr::select(-(1:meta_cols))))
colnames(mat_ipsi) <- proj_ipsi$NeuronID

mat_contra <- t(as.matrix(proj_contra %>% dplyr::select(-(1:meta_cols))))
colnames(mat_contra) <- proj_contra$NeuronID

stopifnot(all.equal(proj_ipsi$NeuronID, proj_contra$NeuronID))

meta <- proj_ipsi %>%
  transmute(
    NeuronID = NeuronID,
    Soma_Region = droplevels(Soma_Region),
    Soma_Side = factor(Soma_Side, levels = c("L", "R")),
    Neuron_Type = factor(Neuron_Type),
    Cluster = Cluster
  ) %>%
  left_join(summary_coords, by = "NeuronID")

cat(sprintf("Neurons: n=%d | L=%d R=%d\n",
            nrow(meta), sum(meta$Soma_Side == "L"), sum(meta$Soma_Side == "R")))
cat(sprintf("Soma regions (hemisphere prefixes stripped in prep): %d levels\n", nlevels(meta$Soma_Region)))
```

### Matrix sparsity diagnostics

```{r matrix_sparsity_report}
matrix_sparsity <- function(m, name) {
  nz_per_neuron <- colSums(m > 0, na.rm = TRUE)
  nz_per_region <- rowSums(m > 0, na.rm = TRUE)
  data.frame(
    matrix = name,
    n_regions = nrow(m),
    n_neurons = ncol(m),
    zero_fraction = mean(m == 0, na.rm = TRUE),
    nonzero_fraction = mean(m > 0, na.rm = TRUE),
    mean_nonzero_targets_per_neuron = mean(nz_per_neuron, na.rm = TRUE),
    median_nonzero_targets_per_neuron = median(nz_per_neuron, na.rm = TRUE),
    mean_neurons_per_region = mean(nz_per_region, na.rm = TRUE),
    median_neurons_per_region = median(nz_per_region, na.rm = TRUE)
  )
}

sparsity_report <- bind_rows(
  matrix_sparsity(mat_ipsi, "ipsi_strength"),
  matrix_sparsity(mat_contra, "contra_strength"),
  {
    # Align by shared region and neuron IDs before summing.
    common_regions <- intersect(rownames(mat_ipsi), rownames(mat_contra))
    common_neurons <- intersect(colnames(mat_ipsi), colnames(mat_contra))
    if (length(common_regions) == 0L || length(common_neurons) == 0L) {
      warning("No overlap between ipsi and contra matrices; total_strength sparsity uses ipsi only.")
      matrix_sparsity(mat_ipsi, "total_strength_fallback_ipsi")
    } else {
      mat_ipsi_al <- mat_ipsi[common_regions, common_neurons, drop = FALSE]
      mat_contra_al <- mat_contra[common_regions, common_neurons, drop = FALSE]
      matrix_sparsity(mat_ipsi_al + mat_contra_al, "total_strength")
    }
  }
)

print(sparsity_report)
cat(sprintf("ipsi matrix sparsity: %.1f%% zeros\n", 100 * sparsity_report$zero_fraction[sparsity_report$matrix == "ipsi_strength"]))
cat(sprintf("contra matrix sparsity: %.1f%% zeros\n", 100 * sparsity_report$zero_fraction[sparsity_report$matrix == "contra_strength"]))
```

Whether **FNT (morphology / feature-neuron distance)** is available controls whether chunks SQ4–SQ5 evaluate (`has_fnt`):

```{r prep_fnt}
has_fnt <- file.exists(PATH_FNT_XLSX)
if (has_fnt) {
  fnt_full <- readxl::read_excel(PATH_FNT_XLSX)
  fnt_mat <- as.matrix(fnt_full)
  rownames(fnt_mat) <- colnames(fnt_mat) <- colnames(fnt_full)
  common_ids <- sort(intersect(colnames(mat_ipsi), colnames(fnt_mat)))
  fnt_sub <- fnt_mat[common_ids, common_ids, drop = FALSE]
  mat_ipsi_fnt <- mat_ipsi[, common_ids, drop = FALSE]
  proj_dist <- as.matrix(vegdist(t(mat_ipsi_fnt), method = "bray"))
  rownames(proj_dist) <- colnames(proj_dist) <- common_ids
  meta_fnt <- meta %>% filter(NeuronID %in% common_ids)
  cat(sprintf("FNT ok: |common|=%d (of %d projection neurons)\n",
              length(common_ids), ncol(mat_ipsi)))
} else {
  common_ids <- character(0)
  fnt_sub <- proj_dist <- NULL
  meta_fnt <- meta
  cat("FNT file not found: skip SQ4–SQ5 or set PATH_FNT_XLSX in config.\n")
}
```

------------------------------------------------------------------------

# Part A: Count-based Laterality Analyses

This section is specialized for **count / composition statistics** (contingency tables, proportions, chi-square/Fisher).

## SQ1: Count-based composition tests block (SQ1a + SQ1b)

This unified SQ1 block includes both count-statistics components:
- **SQ1a (L/R):** `Neuron_Type × Soma_Side`
- **SQ1b (Region):** `Neuron_Type × Soma_Region`
- **SQ1c (Region vs Side count):** `Soma_Region × Soma_Side`

For both tables, the primary inference is Pearson $\chi^2$; Fisher's exact test is used as a fallback when expected counts are small and the table size is feasible.

**Composition display**: Plots use **proportions within each side** ($n_{\text{type,side}} / N_{\text{side}}$), so unequal total neuron counts between L and R do not distort the visual comparison of cell-type mix. **Atlas `Soma_Region` strings are stripped of `CL_` / `CR_` in prep** so labels match across hemispheres; side comes from **`Soma_Side`**. SQ1 includes **raw counts per `Soma_Region` dodged by L/R** for region-wise sampling clarity.

```{r sq1_composition, fig.width=14, fig.height=12.5}
tab <- xtabs(~ Neuron_Type + Soma_Side, data = meta)
print(tab)
# Proportion of each type within each hemisphere (columns sum to 1)
prop_side <- prop.table(tab, margin = 2)
print(round(prop_side, 4))

sq1_by_side <- meta %>%
  dplyr::count(Soma_Side, Neuron_Type, name = "n") %>%
  group_by(Soma_Side) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# Pairwise L vs R test helper for bar pairs (difference in proportions)
.pair_prop_test <- function(xL, xR, nL, nR) {
  tab2 <- matrix(
    c(xL, nL - xL,
      xR, nR - xR),
    nrow = 2,
    byrow = TRUE
  )
  # Use exact test for 2x2 count comparisons to avoid chi-square approximation warnings.
  p <- tryCatch(
    fisher.test(tab2)$p.value,
    error = function(e) NA_real_
  )
  if (is.na(p)) {
    p <- tryCatch(
      suppressWarnings(prop.test(x = c(xL, xR), n = c(nL, nR), correct = FALSE)$p.value),
      error = function(e) NA_real_
    )
  }
  p
}

chi <- suppressWarnings(chisq.test(tab))
print(chi)

# Rule of thumb: expected frequencies < 5 undermine chi-square asymptotics (Cochran's guideline).
exp_ok <- all(chi$expected >= 5, na.rm = TRUE)
if (!exp_ok) {
  message("SQ1: Some expected counts < 5; chi-square p-value may be inaccurate; running Fisher's exact test.")
}
fish_tab <- NULL
if (!exp_ok && is.matrix(tab) && prod(dim(tab)) <= 1e6) {
  fish_tab <- tryCatch(fisher.test(tab), error = function(e) NULL)
  if (!is.null(fish_tab)) print(fish_tab)
}

sq1_lr_summary <- data.frame(
  table_name = "Neuron_Type_x_Soma_Side",
  preferred_test = if (exp_ok) "chi-square" else if (!is.null(fish_tab)) "fisher" else "chi-square_with_small_expected",
  p_value = if (exp_ok || is.null(fish_tab)) chi$p.value else fish_tab$p.value
) %>%
  mutate(
    significant = p_value < 0.05,
    sig_label = p_to_sig(p_value)
  )
print(sq1_lr_summary)
sq1_lr_sig <- ifelse(sq1_lr_summary$significant[1], "sig", "ns")
sq1_lr_test_txt <- sprintf(
  "Type × Side count test (%s): %s, p=%.3g",
  sq1_lr_summary$preferred_test[1],
  sq1_lr_sig,
  sq1_lr_summary$p_value[1]
)
sq1_caption_lr <- paste0(
  "Hypothesis 1: neuron-type composition differs between L and R soma sides. ",
  "Outcome: ", sq1_lr_test_txt, "."
)

side_cols <- c(L = "#E74C3C", R = "#3498DB")

# Region-level count tables used by SQ1b/SQ1c statistics and plots
meta_sq1_region <- meta %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region)))) %>%
  mutate(
    Soma_Region = droplevels(Soma_Region),
    Neuron_Type = droplevels(Neuron_Type)
  )

# Region pick strategy for SQ1:
# keep only bilateral soma regions (both L and R present) to avoid side-encoded labels.
meta_sq1_region <- meta_sq1_region %>%
  group_by(Soma_Region) %>%
  filter(n_distinct(Soma_Side) == 2L) %>%
  ungroup() %>%
  mutate(Soma_Region = droplevels(Soma_Region))

if (nrow(meta_sq1_region) >= 4L && nlevels(meta_sq1_region$Soma_Region) >= 2L) {
  # SQ1c: Region vs Side (count table)
  sq1_region_side_tab <- xtabs(~ Soma_Region + Soma_Side, data = meta_sq1_region)
  print(sq1_region_side_tab)
  sq1_region_side_chi <- suppressWarnings(chisq.test(sq1_region_side_tab))
  print(sq1_region_side_chi)
  sq1_region_side_exp_ok <- all(sq1_region_side_chi$expected >= 5, na.rm = TRUE)
  if (!sq1_region_side_exp_ok && prod(dim(sq1_region_side_tab)) <= 1e6) {
    sq1_region_side_fisher <- tryCatch(fisher.test(sq1_region_side_tab), error = function(e) NULL)
    if (!is.null(sq1_region_side_fisher)) print(sq1_region_side_fisher)
  }
  sq1_region_side_summary <- data.frame(
    table_name = "Soma_Region_x_Soma_Side",
    preferred_test = if (sq1_region_side_exp_ok) "chi-square" else if (!is.null(sq1_region_side_fisher)) "fisher" else "chi-square_with_small_expected",
    p_value = if (sq1_region_side_exp_ok || is.null(sq1_region_side_fisher)) sq1_region_side_chi$p.value else sq1_region_side_fisher$p.value
  ) %>%
    mutate(
      significant = p_value < 0.05,
      sig_label = p_to_sig(p_value)
    )
  print(sq1_region_side_summary)
  sq1_region_side_sig <- ifelse(sq1_region_side_summary$significant[1], "sig", "ns")
  sq1_region_side_test_txt <- sprintf(
    "Region × Side count test (%s): %s, p=%.3g",
    sq1_region_side_summary$preferred_test[1],
    sq1_region_side_sig,
    sq1_region_side_summary$p_value[1]
  )

  # SQ1b: Region vs Type (count table)
  if (nlevels(meta_sq1_region$Neuron_Type) >= 2L) {
    sq1b_tab <- xtabs(~ Neuron_Type + Soma_Region, data = meta_sq1_region)
    print(sq1b_tab)
    print(addmargins(sq1b_tab))
    sq1b_prop_region <- prop.table(sq1b_tab, margin = 2)
    print(round(sq1b_prop_region, 4))

    sq1b_chi <- suppressWarnings(chisq.test(sq1b_tab))
    print(sq1b_chi)
    sq1b_exp_ok <- all(sq1b_chi$expected >= 5, na.rm = TRUE)
    if (!sq1b_exp_ok && prod(dim(sq1b_tab)) <= 1e6) {
      sq1b_fish_tab <- tryCatch(fisher.test(sq1b_tab), error = function(e) NULL)
      if (!is.null(sq1b_fish_tab)) print(sq1b_fish_tab)
    }
    sq1_region_type_summary <- data.frame(
      table_name = "Neuron_Type_x_Soma_Region",
      preferred_test = if (sq1b_exp_ok) "chi-square" else if (!is.null(sq1b_fish_tab)) "fisher" else "chi-square_with_small_expected",
      p_value = if (sq1b_exp_ok || is.null(sq1b_fish_tab)) sq1b_chi$p.value else sq1b_fish_tab$p.value
    ) %>%
      mutate(
        significant = p_value < 0.05,
        sig_label = p_to_sig(p_value)
      )
    print(sq1_region_type_summary)
  }
}
if (!exists("sq1_region_side_test_txt")) {
  sq1_region_side_test_txt <- "Region × Side count test: insufficient bilateral regions"
}
sq1_caption_region <- paste0(
  "Hypothesis 2: soma-region distribution differs between L and R. ",
  "Outcome: ", sq1_region_side_test_txt, "."
)

n_L <- sum(meta$Soma_Side == "L")
n_R <- sum(meta$Soma_Side == "R")

# Per-type L vs R tests for SQ1 paired bars
sq1_type_pair_test <- meta %>%
  group_by(Neuron_Type) %>%
  summarise(
    n_L_type = sum(Soma_Side == "L"),
    n_R_type = sum(Soma_Side == "R"),
    .groups = "drop"
  ) %>%
  mutate(
    p = mapply(.pair_prop_test, n_L_type, n_R_type, MoreArgs = list(nL = n_L, nR = n_R)),
    sig = p_to_sig(p)
  ) %>%
  left_join(
    sq1_by_side %>%
      group_by(Neuron_Type) %>%
      summarise(y = max(prop, na.rm = TRUE) + 0.06, .groups = "drop"),
    by = "Neuron_Type"
  )

sq1_p1 <- ggplot(meta, aes(x = Soma_Side, fill = Soma_Side)) +
  geom_bar(width = 0.6) +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.35,
    size = 4
  ) +
  scale_fill_manual(values = side_cols) +
  labs(
    title = "SQ1: Total neurons per hemisphere (raw N)",
    subtitle = sprintf("N_L = %d, N_R = %d", n_L, n_R),
    x = "Soma side",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.margin = margin(8, 18, 8, 8)
  )

# Within each side, proportion per type (sums to 100% for L and separately for R)
sq1_p2 <- ggplot(sq1_by_side, aes(x = Neuron_Type, y = prop, fill = Soma_Side)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.75) +
  geom_text(
    aes(label = sprintf("%d\n(%s)", n, percent(prop, accuracy = 0.1))),
    position = position_dodge(width = 0.85),
    vjust = -0.1,
    size = 2.85,
  ) +
  geom_text(
    data = sq1_type_pair_test,
    aes(x = Neuron_Type, y = y, label = sig),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 4
  ) +
  scale_fill_manual(values = side_cols, name = "Side") +
  scale_y_continuous(labels = label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "SQ1: Cell-type mix standardized by side (proportion of N_side)",
    subtitle = "Each bar = n_type,side / N_hemisphere; compare L vs R at equal total weight",
    x = "Neuron type",
    y = "Proportion within hemisphere"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    plot.margin = margin(8, 18, 8, 8)
  )

# 100% stacked: full composition per hemisphere on one axis
sq1_p3 <- ggplot(meta, aes(x = Soma_Side, fill = Neuron_Type)) +
  geom_bar(position = "fill", width = 0.55) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "SQ1: Composition (100% per side)",
    subtitle = "Composition view only; inferential test is reported in SQ1 main panel",
    x = "Soma side",
    y = "Proportion within hemisphere"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    plot.margin = margin(8, 18, 8, 8)
  )

# Soma_Region hemisphere-neutral in prep.
# For display coverage, include both bilateral and unilateral non-blank regions.
meta_sq1_region_plot <- meta %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region)))) %>%
  mutate(Soma_Region = droplevels(Soma_Region))

n_drop_region <- nrow(meta) - nrow(meta_sq1_region_plot)
sq1_p4_sub <- if (n_drop_region > 0L) {
  sprintf("Excluded %d neuron(s) with missing/blank Soma_Region.", n_drop_region)
} else {
  "All neurons have a non-blank Soma_Region."
}

sq1_p3b <- NULL
if (nrow(meta_sq1_region_plot)) {
  sq1_region_pair_test <- meta_sq1_region_plot %>%
    group_by(Soma_Region) %>%
    summarise(
      n_L_region = sum(Soma_Side == "L"),
      n_R_region = sum(Soma_Side == "R"),
      .groups = "drop"
    ) %>%
    mutate(
      p = mapply(.pair_prop_test, n_L_region, n_R_region, MoreArgs = list(nL = n_L, nR = n_R)),
      sig = p_to_sig(p)
    ) %>%
    left_join(
      meta_sq1_region_plot %>%
        count(Soma_Region, Soma_Side, name = "n") %>%
        group_by(Soma_Region) %>%
        summarise(y = max(n, na.rm = TRUE) + 1.5, .groups = "drop"),
      by = "Soma_Region"
    )

  sq1_p3b <- ggplot(meta_sq1_region_plot, aes(x = Soma_Region, fill = Soma_Side)) +
    geom_bar(position = position_dodge(width = 0.85), width = 0.78) +
    geom_text(
      stat = "count",
      aes(label = after_stat(count), group = Soma_Side),
      position = position_dodge(width = 0.85),
      vjust = -0.25,
      size = 3.2
    ) +
    geom_text(
      data = sq1_region_pair_test,
      aes(x = Soma_Region, y = y, label = sig),
      inherit.aes = FALSE,
      fontface = "bold",
      size = 4
    ) +
    scale_fill_manual(values = side_cols, name = "Soma side") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.14))) +
    labs(
      title = "SQ1: Neuron count per soma region, by hemisphere",
      subtitle = paste0(
        "Dodged bars: L vs R within each Soma_Region (includes unilateral regions). ",
        "Star labels are per-region L vs R pairwise tests. ",
        sq1_p4_sub
      ),
      x = "Soma_Region",
      y = "Count"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      plot.margin = margin(8, 18, 8, 8)
    )

  # Additional composition view restricted to bilateral regions only
  sq1_region_bilateral_prop <- meta_sq1_region %>%
    count(Soma_Region, Soma_Side, name = "n") %>%
    group_by(Soma_Region) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup()

  sq1_p3c <- ggplot(sq1_region_bilateral_prop, aes(x = Soma_Region, y = prop, fill = Soma_Side)) +
    geom_col(position = "fill", width = 0.78) +
    scale_fill_manual(values = side_cols, name = "Soma side") +
    scale_y_continuous(labels = label_percent(accuracy = 1)) +
    labs(
      title = "SQ1: Bilateral-region composition (100% per region)",
      subtitle = "Only regions with both L and R are included",
      x = "Soma_Region (bilateral only)",
      y = "Proportion within region"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      plot.margin = margin(8, 18, 8, 8)
    )
}

sq1_top <- (sq1_p1 | sq1_p2) + plot_layout(widths = c(1, 2.25), guides = "collect")

# Show each hypothesis caption once at the figure level
sq1_fig <- if (!is.null(sq1_p3b)) {
  patchwork::wrap_plots(
    if (exists("sq1_p3c", inherits = FALSE) && !is.null(sq1_p3c)) {
      list(sq1_top, sq1_p3, sq1_p3b, sq1_p3c)
    } else {
      list(sq1_top, sq1_p3, sq1_p3b)
    },
    ncol = 1
  ) +
    plot_annotation(caption = paste(sq1_caption_lr, sq1_caption_region, sep = "\n"))
} else {
  patchwork::wrap_plots(list(sq1_top, sq1_p3), ncol = 1) +
    plot_annotation(caption = sq1_caption_lr)
}

sq1_fig <- sq1_fig & theme(plot.title = element_text(size = 11.5))
print(sq1_fig)
save_plot(sq1_fig, "sq1_composition.png", width = 14, height = 14)

# SQ1 extra: AP-axis count density by type and by region
sq1_ap_df <- meta %>%
  dplyr::select(NeuronID, Soma_Side, Neuron_Type, Soma_Region, dplyr::any_of("AP_coord"))

if (!("AP_coord" %in% colnames(sq1_ap_df))) {
  ap_col_sq1 <- .pick_first_col(summary_sh, c(
    "Soma_NII_Y", "Soma_Phys_Y",
    "AP", "ap", "Soma_AP", "soma_ap", "Y", "y", "Soma_Y", "soma_y"
  ))
  if (!is.na(ap_col_sq1)) {
    sq1_ap_df <- sq1_ap_df %>%
      dplyr::select(-dplyr::any_of("AP_coord")) %>%
      left_join(
        summary_sh %>%
          transmute(NeuronID = NeuronID, AP_coord = as.numeric(.data[[ap_col_sq1]])),
        by = "NeuronID"
      )
  }
}

sq1_ap_df <- sq1_ap_df %>%
  filter(is.finite(AP_coord), !is.na(Neuron_Type), !is.na(Soma_Region))

if (nrow(sq1_ap_df) >= 10L) {
  sq1_ap_type_plot <- ggplot(sq1_ap_df, aes(x = AP_coord, color = Neuron_Type, fill = Neuron_Type)) +
    geom_density(alpha = 0.16, linewidth = 0.55, adjust = 1.0) +
    labs(
      title = "SQ1 extra: AP distribution of neuron counts by type",
      subtitle = "Density curves reflect count distribution along AP axis (verified NMT RAS: AP = +Y; e.g. Soma_NII_Y)",
      x = "AP coordinate (+Y = anterior)",
      y = "Count density"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "right")
  print(sq1_ap_type_plot)
  save_plot(sq1_ap_type_plot, "sq1_ap_count_density_by_type.png", width = 12, height = 5.8)

  sq1_ap_region_plot <- ggplot(sq1_ap_df, aes(x = AP_coord, color = Soma_Region, fill = Soma_Region)) +
    geom_density(alpha = 0.14, linewidth = 0.45, adjust = 1.0) +
    labs(
      title = "SQ1 extra: AP distribution of neuron counts by region",
      subtitle = "Density curves reflect count distribution along AP axis (verified NMT RAS: AP = +Y; e.g. Soma_NII_Y)",
      x = "AP coordinate (+Y = anterior)",
      y = "Count density"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "right")
  print(sq1_ap_region_plot)
  save_plot(sq1_ap_region_plot, "sq1_ap_count_density_by_region.png", width = 12, height = 5.8)
} else {
  message("SQ1 extra: too few valid AP coordinates for AP count-density plots.")
}

cleanup_vars(c(
  "sq1_p1", "sq1_p2", "sq1_p3", "sq1_p3b", "sq1_p3c",
  "sq1_top", "sq1_fig",
  "sq1_type_pair_test", "sq1_region_pair_test", "sq1_region_bilateral_prop",
  "sq1_ap_df", "sq1_ap_type_plot", "sq1_ap_region_plot"
))
```

------------------------------------------------------------------------

# Part B: Projection-Strength Laterality Analyses

This section is specialized for **projection strength** (continuous outcomes; Wilcoxon / distributional comparisons / multivariate distance models).

## SQ2: Projection-strength counterparts to SQ1 (L/R and Soma_Region)

**SQ2a (L/R counterpart):** Within each `Neuron_Type`, compare total projection strength between L and R (`wilcox.test(total ~ Soma_Side)` with BH correction across types).\
**SQ2b (Region counterpart):** In **bilateral** regions (both L and R present), compare total strength by hemisphere overall within each region and within `Soma_Region × Neuron_Type` strata (BH across strata).

**Note**: `total` is a matrix-derived projection-strength summary, not necessarily physical axon length unless source columns are length-based.

```{r sq2_univariate_strength, fig.width=18, fig.height=10}
strength <- data.frame(
  NeuronID = meta$NeuronID,
  Soma_Side = meta$Soma_Side,
  Neuron_Type = meta$Neuron_Type,
  Soma_Region = meta$Soma_Region,
  ipsi = colSums(mat_ipsi),
  contra = colSums(mat_contra),
  total = colSums(mat_ipsi) + colSums(mat_contra),
  stringsAsFactors = FALSE
)

side_cols <- c(L = "#E74C3C", R = "#3498DB")

# ---- New SQ2 views: region-level density cloud + AP/ML spatial density ----
strength_region_mean <- strength %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region)))) %>%
  group_by(Soma_Region, Soma_Side) %>%
  summarise(
    mean_total = mean(total, na.rm = TRUE),
    median_total = median(total, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

sq2_region_density_with_mean <- strength %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region))))

sq2_region_mean_plot <- ggplot(
  sq2_region_density_with_mean,
  aes(x = total, color = Soma_Region, fill = Soma_Region)
) +
  geom_density(alpha = 0.18, linewidth = 0.35) +
  labs(
    title = "SQ2 extra: Region-wise strength density cloud (all regions stacked)",
    subtitle = "Each colored cloud is one Soma_Region; all regions shown in one panel",
    x = "Total projection strength",
    y = "Density"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

# Coordinate source is now meta dataframe (AP_coord / ML_coord prepared in prep)
strength_spatial <- strength %>%
  left_join(meta %>% dplyr::select(NeuronID, dplyr::any_of(c("AP_coord", "ML_coord"))), by = "NeuronID")

# Fallback: if current session `meta` has no coords (e.g. prep chunk not re-run),
# recover AP/ML directly from `summary_sh` so SQ2 spatial plot still works.
if (!all(c("AP_coord", "ML_coord") %in% colnames(strength_spatial))) {
  ap_col_sq2 <- .pick_first_col(summary_sh, c(
    "Soma_NII_Y", "Soma_Phys_Y",
    "AP", "ap", "Soma_AP", "soma_ap", "Y", "y", "Soma_Y", "soma_y"
  ))
  ml_col_sq2 <- .pick_first_col(summary_sh, c(
    "Soma_NII_X", "Soma_Phys_X",
    "ML", "ml", "Soma_ML", "soma_ml", "X", "x", "Soma_X", "soma_x"
  ))

  if (!is.na(ap_col_sq2) && !is.na(ml_col_sq2)) {
    coord_sq2 <- summary_sh %>%
      transmute(
        NeuronID = NeuronID,
        AP_coord = as.numeric(.data[[ap_col_sq2]]),
        ML_coord = as.numeric(.data[[ml_col_sq2]])
      )
    strength_spatial <- strength %>%
      left_join(coord_sq2, by = "NeuronID")
    message("SQ2 extra: using fallback AP/ML coords from Summary sheet in current chunk.")
  }
}

sq2_spatial_plot <- NULL
if (all(c("AP_coord", "ML_coord") %in% colnames(strength_spatial))) {
  strength_spatial <- strength_spatial %>%
    filter(is.finite(AP_coord), is.finite(ML_coord), is.finite(total))

  if (nrow(strength_spatial) >= 10L) {
    spatial_long <- bind_rows(
      strength_spatial %>%
        transmute(Soma_Side, total, Axis = "AP", Coord = AP_coord),
      strength_spatial %>%
        transmute(Soma_Side, total, Axis = "ML", Coord = ML_coord)
    )

    sq2_spatial_plot <- ggplot(
      spatial_long,
      aes(x = Coord, color = Soma_Side, fill = Soma_Side, weight = total)
    ) +
      geom_density(alpha = 0.20, linewidth = 0.55, adjust = 1.0) +
      facet_wrap(~ Axis, ncol = 1, scales = "free_x") +
      scale_color_manual(values = side_cols, name = "Side") +
      scale_fill_manual(values = side_cols, name = "Side") +
      labs(
        title = "SQ2 extra: Projection-strength-weighted spatial density",
        subtitle = "Density-cloud style along AP/ML axes (weights = total projection strength). Orientation verified from NMT affine (RAS): AP = +Y, ML = +X.",
        x = "Coordinate value (AP: +Y anterior; ML: +X rightward)",
        y = "Weighted density"
      ) +
      theme_minimal(base_size = 11) +
      theme(legend.position = "right")
  } else {
    message("SQ2 extra: no neurons with valid AP/ML coordinates after filtering.")
  }
} else {
  message("SQ2 extra: AP/ML columns not found in meta; spatial plot skipped.")
}

strength_cmp <- strength %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Soma_Side) == 2L) %>%
  ungroup()

# Left panel: only types with both L and R neurons (drop types present on one side only, e.g. all-ITc-on-one-side)
sq2_density <- ggplot(strength_cmp, aes(x = total, fill = Soma_Side)) +
  geom_density(alpha = 0.45, linewidth = 0.3) +
  facet_wrap(~Neuron_Type, scales = "free") +
  scale_fill_manual(values = side_cols, name = "Side") +
  labs(
    title = "SQ2: Distribution of total strength (ipsi + contra)",
    x = "Total projection strength",
    y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 11, face = "plain"),
    axis.text = element_text(size = 10)
  )

sq2_n_lab <- strength_cmp %>%
  group_by(Neuron_Type, Soma_Side) %>%
  summarise(
    n = n(),
    y = max(total, na.rm = TRUE) * 1.08,
    .groups = "drop"
  )

# Right panel: matches laterality_hypothesis_testing.Rmd style — one figure, x = neuron type
sq2_box <- ggplot(strength_cmp, aes(x = Neuron_Type, y = total, fill = Soma_Side)) +
  geom_boxplot(
    position = position_dodge(width = 0.85),
    width = 0.62,
    alpha = 0.88,
    outlier.size = 1
  ) +
  geom_text(
    data = sq2_n_lab,
    aes(x = Neuron_Type, y = y, label = paste0("n=", n), group = Soma_Side),
    position = position_dodge(width = 0.85),
    size = 3.2, 
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = side_cols, name = "Side") +
    # Two-sample Wilcoxon (Mann–Whitney); uncorrected p per type — see table for BH-adjusted q
    stat_compare_means(
    aes(group = Soma_Side),
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE,
    tip.length = 0.02,
    bracket.size = 0.22,
    step.increase = 0.06
  ) +
  labs(
    title = "SQ2: Total projection by type & side (Wilcoxon L vs R within each type; plot p uncorrected)",
    x = "Neuron type",
    y = "Total projection strength"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sq2_row1 <- (sq2_density | sq2_box) +
  plot_layout(widths = c(1, 1), guides = "collect") &
  theme(plot.title = element_text(size = 12))

strength_reg <- strength %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region)))) %>%
  mutate(Soma_Region = droplevels(factor(Soma_Region)))

all_region_levels <- strength_reg %>%
  distinct(Soma_Region) %>%
  pull(Soma_Region)

bilateral_region_levels <- strength_reg %>%
  group_by(Soma_Region) %>%
  filter(n_distinct(Soma_Side) == 2L) %>%
  group_keys() %>%
  pull(Soma_Region)

strength_bilateral_regions <- strength_reg %>%
  filter(Soma_Region %in% bilateral_region_levels) %>%
  mutate(Soma_Region = droplevels(Soma_Region))

sq2_region_density <- NULL
wt_region_overall <- NULL
wt_region_type <- NULL

if (nrow(strength_reg) >= 2L && length(all_region_levels)) {
  sq2_region_density <- ggplot(strength_reg, aes(x = total, fill = Soma_Side)) +
    geom_density(alpha = 0.45, linewidth = 0.3) +
    facet_wrap(~ Soma_Region, nrow = 1L, scales = "free") +
    scale_fill_manual(values = side_cols, name = "Side") +
    labs(
      title = "SQ2 (regions): Total strength by hemisphere across all soma regions",
      subtitle = "All neuron types pooled; unilateral regions show one-side density only",
      x = "Total projection strength",
      y = "Density"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text = element_text(size = 10, face = "plain"),
      legend.position = "bottom"
    )
}

if (nrow(strength_bilateral_regions) >= 4L && length(bilateral_region_levels)) {
  wt_region_overall <- strength_bilateral_regions %>%
    group_by(Soma_Region) %>%
    filter(n_distinct(Soma_Side) == 2L, n() >= 4L) %>%
    summarise(
      p_wilcox_total = tryCatch(
        wilcox.test(total ~ Soma_Side)$p.value,
        error = function(e) NA_real_
      ),
      n = n(),
      n_L = sum(Soma_Side == "L"),
      n_R = sum(Soma_Side == "R"),
      .groups = "drop"
    ) %>%
    mutate(
      p_wilcox_BH = p.adjust(p_wilcox_total, method = "BH"),
      significant = p_wilcox_BH < 0.05,
      sig_label = p_to_sig(p_wilcox_BH)
    )

  wt_region_type <- strength_bilateral_regions %>%
    group_by(Soma_Region, Neuron_Type) %>%
    filter(n_distinct(Soma_Side) == 2L, n() >= 4L) %>%
    summarise(
      p_wilcox_total = tryCatch(
        wilcox.test(total ~ Soma_Side)$p.value,
        error = function(e) NA_real_
      ),
      n = n(),
      n_L = sum(Soma_Side == "L"),
      n_R = sum(Soma_Side == "R"),
      .groups = "drop"
    ) %>%
    mutate(
      p_wilcox_BH = p.adjust(p_wilcox_total, method = "BH"),
      significant = p_wilcox_BH < 0.05,
      sig_label = p_to_sig(p_wilcox_BH)
    )

  cat("SQ2: Wilcoxon total ~ Side, bilateral regions (all types pooled)\n")
  print(wt_region_overall)
  cat("SQ2: Wilcoxon total ~ Side within Soma_Region × Neuron_Type (BH across strata)\n")
  print(wt_region_type)
}
sq2_region_outcome_txt <- if (!is.null(wt_region_overall) && nrow(wt_region_overall)) {
  n_sig <- sum(wt_region_overall$significant, na.rm = TRUE)
  sprintf(
    "SQ2b outcome (bilateral-only Wilcoxon + BH): %d/%d bilateral regions significant.",
    n_sig,
    nrow(wt_region_overall)
  )
} else {
  "SQ2b outcome: insufficient bilateral regions for inferential region-side test."
}

sq2a_eval <- strength_cmp %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Soma_Side) == 2L, n() >= 4L) %>%
  summarise(
    p_raw = tryCatch(wilcox.test(total ~ Soma_Side)$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    p_BH = p.adjust(p_raw, method = "BH"),
    significant = p_BH < 0.05
  )
sq2a_outcome_txt <- if (nrow(sq2a_eval)) {
  sprintf(
    "SQ2a outcome (type-wise Wilcoxon + BH): %d/%d neuron types significant.",
    sum(sq2a_eval$significant, na.rm = TRUE),
    nrow(sq2a_eval)
  )
} else {
  "SQ2a outcome: insufficient type strata for L/R inferential test."
}

sq2_caption <- paste(
  "Hypothesis SQ2a: total projection strength differs between L and R within neuron type (Wilcoxon, BH across types).",
  sq2a_outcome_txt,
  "Hypothesis SQ2b: total projection strength differs between L and R within soma region.",
  sq2_region_outcome_txt,
  sep = "\n"
)
sq2_fig <- if (!is.null(sq2_region_density)) {
  (sq2_row1) / sq2_region_density +
    plot_layout(heights = c(1, 0.68), guides = "collect")
} else {
  message("SQ2: No valid Soma_Region entries; skipping region-level strength plots.")
  sq2_row1
}
sq2_fig <- sq2_fig +
  plot_annotation(caption = sq2_caption) &
  theme(plot.title = element_text(size = 11), legend.position = "bottom")
print(sq2_fig)
save_plot(sq2_fig, "sq2_univariate_strength.png", width = 18, height = 10)

print(sq2_region_mean_plot)
save_plot(sq2_region_mean_plot, "sq2_region_mean_strength.png", width = 15, height = 7)

if (!is.null(sq2_spatial_plot)) {
  print(sq2_spatial_plot)
  save_plot(sq2_spatial_plot, "sq2_ap_ml_strength_density.png", width = 12, height = 5.8)
}
cleanup_vars(c(
  "sq2_density", "sq2_box", "sq2_row1", "sq2_region_density", "sq2_fig",
  "sq2_n_lab", "strength_reg", "strength_bilateral_regions", "all_region_levels", "bilateral_region_levels",
  "strength_region_mean", "sq2_region_density_with_mean", "sq2_region_mean_plot",
  "strength_spatial", "sq2_spatial_plot"
))

# Per-type Wilcoxon on total strength (no region stratification); BH across types
wt <- strength %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Soma_Side) == 2, n() >= 4) %>%
  summarise(
    p_wilcox_total = tryCatch(
      wilcox.test(total ~ Soma_Side)$p.value,
      error = function(e) NA_real_
    ),
    n = n(),
    .groups = "drop"
  )
wt <- wt %>%
  mutate(
    p_wilcox_BH = p.adjust(p_wilcox_total, method = "BH"),
    significant = p_wilcox_BH < 0.05,
    sig_label = p_to_sig(p_wilcox_BH)
  )
print(wt)
```

### SQ2c (v2): `lm` on log total strength (side after neuron type)

**Interpretation:** Type I **ANOVA** tests `Soma_Side` after `Neuron_Type`, addressing uneven L/R *composition* across types. **`drop1(..., test = "F")`** gives marginal deletion tests (each term given others). Report both briefly in the HTML; the workbook stores **`SQ2_lm_anova`** and **`SQ2_lm_drop1`**.

```{r sq2_lm_adjusted}
fit_total <- tryCatch(
  lm(log1p(total) ~ Neuron_Type + Soma_Side, data = strength),
  error = function(e) NULL
)
if (is.null(fit_total)) {
  message("SQ2 v2: lm(log1p(total) ~ ...) failed (e.g. rank deficiency).")
  sq2_lm_anova <- NULL
  sq2_lm_drop1 <- NULL
  sq2_lm_side <- NULL
} else {
  sq2_lm_anova <- as.data.frame(anova(fit_total))
  sq2_lm_anova$term <- rownames(sq2_lm_anova)
  rownames(sq2_lm_anova) <- NULL
  sq2_lm_drop1 <- as.data.frame(drop1(fit_total, test = "F"))
  sq2_lm_drop1$term <- rownames(sq2_lm_drop1)
  rownames(sq2_lm_drop1) <- NULL
  sq2_lm_side <- sq2_lm_drop1 %>% filter(term == "Soma_Side")
  cat("SQ2 v2: ANOVA (Type I; order = Neuron_Type, then Soma_Side)\n")
  print(sq2_lm_anova)
  cat("\nSQ2 v2: drop1 marginal F-tests\n")
  print(sq2_lm_drop1)
}
```

------------------------------------------------------------------------

## SQ3: Do left and right differ in **multivariate projection pattern** (target-region vectors)?

**Hypothesis**: In the high-dimensional space of region × strength, the L vs R clouds differ in location and/or spread.\
**Why ipsi-only here?** We use `mat_ipsi` to isolate hemisphere-specific targeting and avoid a trivial side-coupling that can appear when ipsi + contra are pooled (because contra definitions are side-referenced). This keeps the distance model focused on within-hemisphere projection organization.\
**Steps**: (1) neuron × region matrix (`mat_ipsi`) → Bray–Curtis dissimilarities between neurons; (2) `adonis2` (PERMANOVA) on `Side`; (3) `betadisper` + `permutest` to test dispersion homogeneity (with euclidification for Bray–Curtis in this step: `sqrt.dist=TRUE`, `add=\"lingoes\"`); (4) PCoA for ordination. **Caveat**: PERMANOVA assumes **exchangeability** of residuals under permutation; heterogeneity of dispersion can inflate false positives, hence the explicit `betadisper` check. $R^2$ is **partial variance explained** by `Side` in the pseudo-F framework, not OLS $R^2$.

```{r sq3_permanova, fig.width=14, fig.height=13}
ipsi_nr <- t(mat_ipsi) # neuron x region
side <- meta$Soma_Side[match(rownames(ipsi_nr), meta$NeuronID)]

# Profile-normalized matrix (composition-like view)
ipsi_tot <- rowSums(ipsi_nr)
ipsi_prop <- ipsi_nr
ok_tot <- ipsi_tot > 0
ipsi_prop[ok_tot, ] <- ipsi_nr[ok_tot, , drop = FALSE] / ipsi_tot[ok_tot]

# Presence/absence matrix for Jaccard
ipsi_pa <- ipsi_nr > 0

# Aitchison via CLR + Euclidean with a small pseudocount
pos_vals <- ipsi_nr[ipsi_nr > 0]
pseudocount <- if (length(pos_vals)) min(pos_vals) / 2 else 1e-6
ipsi_clr <- log(ipsi_nr + pseudocount) - rowMeans(log(ipsi_nr + pseudocount))

dist_list <- list(
  bray_raw = vegdist(ipsi_nr, method = "bray"),
  bray_prop = vegdist(ipsi_prop, method = "bray"),
  jaccard_pa = vegdist(ipsi_pa, method = "jaccard", binary = TRUE),
  aitchison_clr = dist(ipsi_clr)
)

sq3_multi_tab <- lapply(names(dist_list), function(nm) {
  d <- dist_list[[nm]]
  side_d <- side[match(labels(d), rownames(ipsi_nr))]
  ado_i <- adonis2(d ~ side_d, permutations = PERM_N, by = "margin")

  # Euclidify for dispersion check when distance is non-Euclidean
  bd_i <- betadisper(d, side_d, type = "centroid", sqrt.dist = TRUE, add = "lingoes")
  bd_perm_i <- permutest(bd_i, permutations = PERM_N)
  p_bd <- suppressWarnings(as.numeric(bd_perm_i$tab[1, "Pr(>F)"]))

  data.frame(
    distance = nm,
    R2 = as.numeric(ado_i$R2[1]),
    p_permanova = as.numeric(ado_i$`Pr(>F)`[1]),
    p_betadisper = p_bd
  )
}) %>%
  bind_rows() %>%
  mutate(
    significant_permanova = p_permanova < 0.05,
    sig_permanova = p_to_sig(p_permanova),
    significant_betadisper = p_betadisper < 0.05,
    sig_betadisper = p_to_sig(p_betadisper)
  )
print(sq3_multi_tab)

# --- v2: covariate-adjusted PERMANOVA (rows match `ipsi_nr` / dist_list) ---
df_sq3 <- meta[match(rownames(ipsi_nr), meta$NeuronID), , drop = FALSE]
df_sq3 <- transmute(
  df_sq3,
  NeuronID = NeuronID,
  Neuron_Type = factor(Neuron_Type),
  Soma_Side = factor(Soma_Side, levels = c("L", "R")),
  Soma_Region = droplevels(factor(Soma_Region))
)
stopifnot(all.equal(as.character(df_sq3$NeuronID), rownames(ipsi_nr)))

sq3_adjusted_list <- list()
for (nm in names(dist_list)) {
  d <- dist_list[[nm]]
  a_ts <- tryCatch(
    adonis2(d ~ Neuron_Type + Soma_Side, data = df_sq3, permutations = PERM_N, by = "margin"),
    error = function(e) NULL
  )
  if (!is.null(a_ts)) {
    tt <- as.data.frame(a_ts)
    tt$term <- rownames(tt)
    rownames(tt) <- NULL
    tt$distance <- nm
    tt$model <- "~ Neuron_Type + Soma_Side"
    sq3_adjusted_list[[length(sq3_adjusted_list) + 1L]] <- tt
  }
  a_trs <- tryCatch(
    adonis2(
      d ~ Neuron_Type + Soma_Region + Soma_Side,
      data = df_sq3,
      permutations = PERM_N,
      by = "margin"
    ),
    error = function(e) NULL
  )
  if (!is.null(a_trs)) {
    tt <- as.data.frame(a_trs)
    tt$term <- rownames(tt)
    rownames(tt) <- NULL
    tt$distance <- nm
    tt$model <- "~ Neuron_Type + Soma_Region + Soma_Side"
    sq3_adjusted_list[[length(sq3_adjusted_list) + 1L]] <- tt
  }
}
sq3_adjusted_tab <- if (length(sq3_adjusted_list)) {
  res <- bind_rows(sq3_adjusted_list)
  pcol <- names(res)[grepl("^Pr\\(", names(res))]
  if (length(pcol) == 1L) {
    res$sig_label <- p_to_sig(as.numeric(res[[pcol]]))
  } else {
    res$sig_label <- NA_character_
  }
  res
} else {
  message("SQ3 v2: adjusted PERMANOVA tables could not be built.")
  NULL
}
if (!is.null(sq3_adjusted_tab) && nrow(sq3_adjusted_tab)) {
  cat("\nSQ3 v2: Adjusted PERMANOVA (by = margin). Soma_Side row = partial effect after preceding terms in formula.\n")
  print(sq3_adjusted_tab)
}

# SQ3-specific sparsity diagnostics on the matrices used in this section.
sq3_sparsity <- bind_rows(
  matrix_sparsity(t(ipsi_nr), "sq3_ipsi_raw_region_by_neuron"),
  matrix_sparsity(t(ipsi_prop), "sq3_ipsi_profile_normalized"),
  matrix_sparsity(t(ipsi_pa), "sq3_ipsi_presence_absence")
) %>%
  mutate(zero_percent = 100 * zero_fraction)
print(sq3_sparsity)

# Keep legacy export variables based on primary metric (bray_raw)
sq3_summary <- sq3_multi_tab %>%
  filter(distance == "bray_raw") %>%
  transmute(
    term = "side",
    R2 = R2,
    p = p_permanova,
    significant = significant_permanova,
    sig_label = sig_permanova
  )
sq3_betadisper <- sq3_multi_tab %>%
  filter(distance == "bray_raw") %>%
  transmute(
    p = p_betadisper,
    significant = significant_betadisper,
    sig_label = sig_betadisper
  )

ord_plot <- function(d, title, subtitle = NULL) {
  md <- cmdscale(d, k = 2, eig = TRUE)
  eig_pos <- pmax(0, md$eig)
  pct <- if (sum(eig_pos) > 0) eig_pos / sum(eig_pos) * 100 else c(NA_real_, NA_real_)
  odf <- data.frame(
    NeuronID = labels(d),
    Axis1 = md$points[, 1],
    Axis2 = md$points[, 2]
  ) %>%
    left_join(meta, by = "NeuronID")
  ggplot(odf, aes(.data$Axis1, .data$Axis2, color = .data$Soma_Side)) +
    geom_point(alpha = 0.8) +
    stat_ellipse(level = 0.95, linewidth = 0.35) +
    scale_color_manual(values = c(L = "#E74C3C", R = "#3498DB")) +
    labs(
      title = title,
      subtitle = subtitle,
      x = sprintf("Axis 1 (%.1f%%)", pct[1]),
      y = sprintf("Axis 2 (%.1f%%)", pct[2])
    ) +
    theme_minimal(base_size = 10)
}

row_br <- sq3_multi_tab %>% filter(distance == "bray_raw")
p1 <- ord_plot(
  dist_list$bray_raw,
  "SQ3: PCoA (Bray raw)",
  sprintf("PERMANOVA p=%.4f, betadisper p=%.4f", row_br$p_permanova[1], row_br$p_betadisper[1])
)
p2 <- ord_plot(dist_list$bray_prop, "SQ3: PCoA (Bray on normalized profile)")
p3 <- ord_plot(dist_list$jaccard_pa, "SQ3: PCoA (Jaccard presence/absence)")
p4 <- ord_plot(dist_list$aitchison_clr, "SQ3: PCoA (Aitchison CLR-Euclidean)")
(p1 | p2) / (p3 | p4)
```

**v2 note:** Adjusted PERMANOVA tables are printed above. **SQ6** still stratifies by neuron type. If SQ1 shows strong type × side association, prioritize **adjusted** and **balance MC** summaries alongside the unadjusted multi-distance table.

------------------------------------------------------------------------

## SQ3b (v2): Stratified equal-n Monte Carlo PERMANOVA (Bray–Curtis)

Within each `Neuron_Type`, sample `min(n_L, n_R)` neurons per side (without replacement), pool IDs across types, then run `adonis2(~ Soma_Side)` on **Bray–Curtis** of the subset. Repeat **`BALANCE_REPS`** times. Summaries describe how stable unadjusted side separation is when L/R counts are matched *within type*. Set `RUN_BALANCE_PERMANOVA <- FALSE` in `config` to skip (faster knit).

```{r sq3_balance_mc}
pick_balanced_neuron_ids <- function(meta_rows) {
  out <- character(0)
  for (ty in unique(as.character(meta_rows$Neuron_Type))) {
    m <- meta_rows[as.character(meta_rows$Neuron_Type) == ty, , drop = FALSE]
    if (n_distinct(m$Soma_Side) < 2L) next
    nL <- sum(m$Soma_Side == "L")
    nR <- sum(m$Soma_Side == "R")
    nstar <- min(nL, nR)
    if (nstar < BALANCE_MIN_PER_SIDE) next
    idL <- sample(m$NeuronID[m$Soma_Side == "L"], nstar)
    idR <- sample(m$NeuronID[m$Soma_Side == "R"], nstar)
    out <- c(out, idL, idR)
  }
  unique(out)
}

meta_ipsi <- meta[match(rownames(ipsi_nr), meta$NeuronID), , drop = FALSE]

sq3_balance_mc_runs <- data.frame(
  rep = integer(),
  distance = character(),
  R2 = numeric(),
  p = numeric()
)
sq3_balance_mc_runs_prop <- data.frame(
  rep = integer(),
  distance = character(),
  R2 = numeric(),
  p = numeric()
)

if (isTRUE(RUN_BALANCE_PERMANOVA)) {
  bruns <- list()
  for (r in seq_len(BALANCE_REPS)) {
    ids <- pick_balanced_neuron_ids(meta_ipsi)
    if (length(ids) < 6L) next
    sub_nr <- ipsi_nr[ids, , drop = FALSE]
    d <- tryCatch(vegdist(sub_nr, method = "bray"), error = function(e) NULL)
    if (is.null(d)) next
    sd <- meta$Soma_Side[match(labels(d), meta$NeuronID)]
    if (length(unique(sd)) < 2L) next
    a <- tryCatch(
      adonis2(d ~ sd, permutations = PERM_N),
      error = function(e) NULL
    )
    if (is.null(a)) next
    bruns[[length(bruns) + 1L]] <- data.frame(
      rep = r,
      distance = "bray_raw",
      R2 = as.numeric(a$R2[1L]),
      p = as.numeric(a$`Pr(>F)`[1L])
    )
  }
  sq3_balance_mc_runs <- bind_rows(bruns)

  if (isTRUE(RUN_BALANCE_BRAY_PROP)) {
    bruns_p <- list()
    for (r in seq_len(BALANCE_REPS)) {
      ids <- pick_balanced_neuron_ids(meta_ipsi)
      if (length(ids) < 6L) next
      sub_nr <- ipsi_nr[ids, , drop = FALSE]
      rt <- rowSums(sub_nr)
      ok <- rt > 0
      sub_prop <- sub_nr
      sub_prop[ok, ] <- sub_nr[ok, , drop = FALSE] / rt[ok]
      d <- tryCatch(vegdist(sub_prop, method = "bray"), error = function(e) NULL)
      if (is.null(d)) next
      sd <- meta$Soma_Side[match(labels(d), meta$NeuronID)]
      if (length(unique(sd)) < 2L) next
      a <- tryCatch(
        adonis2(d ~ sd, permutations = PERM_N),
        error = function(e) NULL
      )
      if (is.null(a)) next
      bruns_p[[length(bruns_p) + 1L]] <- data.frame(
        rep = r,
        distance = "bray_prop",
        R2 = as.numeric(a$R2[1L]),
        p = as.numeric(a$`Pr(>F)`[1L])
      )
    }
    sq3_balance_mc_runs_prop <- bind_rows(bruns_p)
  }
} else {
  message("SQ3 v2: RUN_BALANCE_PERMANOVA is FALSE; skipping Monte Carlo balance PERMANOVA.")
}

summ_balance <- function(df) {
  if (is.null(df) || !nrow(df)) {
    return(data.frame(note = "no successful replicates", stringsAsFactors = FALSE))
  }
  data.frame(
    distance = df$distance[1L],
    n_reps = nrow(df),
    median_R2 = median(df$R2, na.rm = TRUE),
    mean_R2 = mean(df$R2, na.rm = TRUE),
    median_p = median(df$p, na.rm = TRUE),
    mean_p = mean(df$p, na.rm = TRUE),
    prop_p_lt_0.05 = mean(df$p < 0.05, na.rm = TRUE),
    q025_R2 = unname(quantile(df$R2, 0.025, na.rm = TRUE)),
    q975_R2 = unname(quantile(df$R2, 0.975, na.rm = TRUE)),
    stringsAsFactors = FALSE
  )
}

sq3_balance_mc_summary <- summ_balance(sq3_balance_mc_runs)
if (isTRUE(RUN_BALANCE_BRAY_PROP)) {
  sq3_balance_mc_summary <- bind_rows(
    sq3_balance_mc_summary,
    summ_balance(sq3_balance_mc_runs_prop)
  )
}
cat("\nSQ3 v2: Monte Carlo balance PERMANOVA summary\n")
print(sq3_balance_mc_summary)
```

------------------------------------------------------------------------

## SQ3c (v2): Region-stratified multivariate robustness (projection matrix)

This block keeps the **multivariate** framework (neuron x region projection matrix) and tests L/R separation within:

1. Coarse insula classes: `IA*` (agranular) vs `ID*` (dysgranular)  
2. Targeted regions: `IDD5` and `IDM`

All tests below use Bray-Curtis on `ipsi_nr` with PERMANOVA (`adonis2`), plus a dispersion check (`betadisper`) to flag potential spread confounds.

```{r sq3_region_stratified_mv, fig.width=12, fig.height=8}
meta_sq3 <- meta[match(rownames(ipsi_nr), meta$NeuronID), , drop = FALSE] %>%
  mutate(
    Soma_Region_chr = trimws(as.character(Soma_Region)),
    RegionClass = case_when(
      grepl("^IA", Soma_Region_chr, ignore.case = TRUE) ~ "agranular",
      grepl("^ID", Soma_Region_chr, ignore.case = TRUE) ~ "dysgranular",
      TRUE ~ NA_character_
    ),
    RegionTarget = case_when(
      Soma_Region_chr %in% c("IDD5", "IDM") ~ Soma_Region_chr,
      TRUE ~ NA_character_
    )
  )

run_mv_side_test <- function(ids, strata_label) {
  ids <- intersect(ids, rownames(ipsi_nr))
  if (length(ids) < 6L) return(NULL)

  sub_meta <- meta_sq3[match(ids, meta_sq3$NeuronID), , drop = FALSE]
  if (n_distinct(sub_meta$Soma_Side) < 2L) return(NULL)

  sub_nr <- ipsi_nr[ids, , drop = FALSE]
  d <- tryCatch(vegdist(sub_nr, method = "bray"), error = function(e) NULL)
  if (is.null(d)) return(NULL)

  side_sub <- droplevels(factor(sub_meta$Soma_Side[match(labels(d), sub_meta$NeuronID)], levels = c("L", "R")))
  if (n_distinct(side_sub) < 2L) return(NULL)

  a <- tryCatch(adonis2(d ~ side_sub, permutations = PERM_N, by = "margin"), error = function(e) NULL)
  if (is.null(a)) return(NULL)

  bd <- tryCatch(betadisper(d, side_sub, type = "centroid", sqrt.dist = TRUE, add = "lingoes"), error = function(e) NULL)
  p_bd <- if (!is.null(bd)) {
    suppressWarnings(as.numeric(permutest(bd, permutations = PERM_N)$tab[1, "Pr(>F)"]))
  } else {
    NA_real_
  }

  pcoa <- cmdscale(d, k = 2, eig = TRUE)
  ord <- data.frame(
    Axis1 = pcoa$points[, 1],
    Axis2 = pcoa$points[, 2],
    Soma_Side = side_sub
  )

  list(
    summary = data.frame(
      stratum = strata_label,
      n = length(ids),
      n_L = sum(side_sub == "L"),
      n_R = sum(side_sub == "R"),
      R2 = as.numeric(a$R2[1]),
      p_permanova = as.numeric(a$`Pr(>F)`[1]),
      p_betadisper = p_bd,
      stringsAsFactors = FALSE
    ),
    ord = ord
  )
}

# (A) Coarse classes IA* vs ID*
class_results <- lapply(c("agranular", "dysgranular"), function(cls) {
  ids <- meta_sq3$NeuronID[meta_sq3$RegionClass == cls & !is.na(meta_sq3$RegionClass)]
  run_mv_side_test(ids, paste0("class_", cls))
})
names(class_results) <- c("agranular", "dysgranular")
class_tab <- bind_rows(lapply(class_results, function(x) if (is.null(x)) NULL else x$summary))
if (nrow(class_tab)) {
  class_tab <- class_tab %>%
    mutate(
      p_BH = p.adjust(p_permanova, method = "BH"),
      significant = p_BH < 0.05,
      sig_label = p_to_sig(p_BH),
      side_ratio = pmax(n_L, n_R) / pmin(n_L, n_R)
    )
}
cat("SQ3c: IA/ID class-stratified multivariate side tests\n")
print(class_tab)

# (B) Targeted regions IDD5 and IDM
target_results <- lapply(c("IDD5", "IDM"), function(reg) {
  ids <- meta_sq3$NeuronID[meta_sq3$RegionTarget == reg & !is.na(meta_sq3$RegionTarget)]
  run_mv_side_test(ids, paste0("target_", reg))
})
names(target_results) <- c("IDD5", "IDM")
target_tab <- bind_rows(lapply(target_results, function(x) if (is.null(x)) NULL else x$summary))
if (nrow(target_tab)) {
  target_tab <- target_tab %>%
    mutate(
      p_BH = p.adjust(p_permanova, method = "BH"),
      significant = p_BH < 0.05,
      sig_label = p_to_sig(p_BH),
      side_ratio = pmax(n_L, n_R) / pmin(n_L, n_R)
    )
}
cat("SQ3c: targeted multivariate side tests (IDD5, IDM)\n")
print(target_tab)

# Simple ordination visual checks
plot_ord <- function(res, ttl) {
  if (is.null(res) || is.null(res$ord) || !nrow(res$ord)) return(NULL)
  ggplot(res$ord, aes(.data$Axis1, .data$Axis2, color = .data$Soma_Side)) +
    geom_point(size = 2.0, alpha = 0.8) +
    stat_ellipse(aes(fill = .data$Soma_Side), geom = "polygon", alpha = 0.14, color = NA) +
    scale_color_manual(values = c(L = "#E74C3C", R = "#3498DB")) +
    scale_fill_manual(values = c(L = "#E74C3C", R = "#3498DB")) +
    labs(title = ttl, x = "PCoA1", y = "PCoA2", color = "Side", fill = "Side") +
    theme_minimal(base_size = 11)
}

ord_plots <- list(
  plot_ord(class_results$agranular, "SQ3c ordination: IA* (agranular)"),
  plot_ord(class_results$dysgranular, "SQ3c ordination: ID* (dysgranular)"),
  plot_ord(target_results$IDD5, "SQ3c ordination: IDD5"),
  plot_ord(target_results$IDM, "SQ3c ordination: IDM")
)
ord_plots <- ord_plots[!vapply(ord_plots, is.null, logical(1))]
if (length(ord_plots)) {
  patchwork::wrap_plots(ord_plots, ncol = 2, guides = "collect") +
    plot_annotation(
      caption = "Hypothesis: within each stratum, L and R differ in multivariate projection pattern (Bray PERMANOVA). Report BH-adjusted p and betadisper p for dispersion caveat."
    )
}

# Outcome report (auto-text)
sq3c_class_outcome <- if (nrow(class_tab)) {
  sprintf(
    "SQ3c outcome (IA/ID classes): %d/%d significant after BH; max side-imbalance ratio = %.2f.",
    sum(class_tab$significant, na.rm = TRUE),
    nrow(class_tab),
    max(class_tab$side_ratio, na.rm = TRUE)
  )
} else {
  "SQ3c outcome (IA/ID classes): insufficient strata for testing."
}
sq3c_target_outcome <- if (nrow(target_tab)) {
  sprintf(
    "SQ3c outcome (IDD5/IDM): %d/%d significant after BH; max side-imbalance ratio = %.2f.",
    sum(target_tab$significant, na.rm = TRUE),
    nrow(target_tab),
    max(target_tab$side_ratio, na.rm = TRUE)
  )
} else {
  "SQ3c outcome (IDD5/IDM): insufficient strata for testing."
}
cat("\n", sq3c_class_outcome, "\n", sq3c_target_outcome, "\n", sep = "")

save_plot(
  plot_obj = if (length(ord_plots)) patchwork::wrap_plots(ord_plots, ncol = 2, guides = "collect") else ggplot() + theme_void(),
  filename = "sq3c_region_stratified_multivariate.png",
  width = 12,
  height = 8
)

cleanup_vars(c("meta_sq3", "class_results", "class_tab", "target_results", "target_tab", "ord_plots"))
```

------------------------------------------------------------------------

## SQ3d (v2): Projection laterality index by target brain region

Following prior single-neuron projectome work, this block summarizes **projection asymmetry** using a region-wise laterality index:

\[
LI = \frac{\mu_L - \mu_R}{\mu_L + \mu_R + \epsilon}
\]

where \(\mu_L, \mu_R\) are mean ipsilateral projection strengths onto each target brain region from L/R neuron groups.

```{r sq3_laterality_index, fig.width=13, fig.height=8}
eps_li <- 1e-8
region_names <- rownames(mat_ipsi)
mat_ipsi_prop_li <- mat_ipsi
tot_per_neuron_li <- colSums(mat_ipsi_prop_li, na.rm = TRUE)
ok_tot_li <- tot_per_neuron_li > 0
mat_ipsi_prop_li[, ok_tot_li] <- sweep(
  mat_ipsi_prop_li[, ok_tot_li, drop = FALSE],
  2,
  tot_per_neuron_li[ok_tot_li],
  "/"
)
meta_li <- meta[match(colnames(mat_ipsi), meta$NeuronID), , drop = FALSE] %>%
  mutate(
    Soma_Region_chr = trimws(as.character(Soma_Region)),
    RegionClass = case_when(
      grepl("^IA", Soma_Region_chr, ignore.case = TRUE) ~ "agranular",
      grepl("^ID", Soma_Region_chr, ignore.case = TRUE) ~ "dysgranular",
      TRUE ~ NA_character_
    )
  )

calc_li_table <- function(ids, stratum_name) {
  ids <- intersect(ids, colnames(mat_ipsi))
  if (length(ids) < 6L) return(NULL)

  sub_meta <- meta_li[match(ids, meta_li$NeuronID), , drop = FALSE]
  if (n_distinct(sub_meta$Soma_Side) < 2L) return(NULL)

  ids_L <- sub_meta$NeuronID[sub_meta$Soma_Side == "L"]
  ids_R <- sub_meta$NeuronID[sub_meta$Soma_Side == "R"]
  if (length(ids_L) < 3L || length(ids_R) < 3L) return(NULL)

  mean_L <- rowMeans(mat_ipsi_prop_li[, ids_L, drop = FALSE], na.rm = TRUE)
  mean_R <- rowMeans(mat_ipsi_prop_li[, ids_R, drop = FALSE], na.rm = TRUE)

  tab <- data.frame(
    stratum = stratum_name,
    target_region = region_names,
    mean_L = mean_L,
    mean_R = mean_R,
    n_L = length(ids_L),
    n_R = length(ids_R),
    LI = (mean_L - mean_R) / (mean_L + mean_R + eps_li),
    p_wilcox = vapply(region_names, function(rg) {
      x <- as.numeric(mat_ipsi_prop_li[rg, ids_L, drop = TRUE])
      y <- as.numeric(mat_ipsi_prop_li[rg, ids_R, drop = TRUE])
      tryCatch(
        suppressWarnings(wilcox.test(x, y, exact = FALSE)$p.value),
        error = function(e) NA_real_
      )
    }, numeric(1)),
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      abs_LI = abs(.data$LI),
      p_BH = p.adjust(.data$p_wilcox, method = "BH"),
      significant = .data$p_BH < 0.05,
      sig_label = p_to_sig(.data$p_BH)
    )
  tab
}

# Coarse LR (all neurons)
li_all <- calc_li_table(meta_li$NeuronID, "all_neurons")

# Agranular vs dysgranular
li_agran <- calc_li_table(meta_li$NeuronID[meta_li$RegionClass == "agranular"], "class_agranular")
li_dys   <- calc_li_table(meta_li$NeuronID[meta_li$RegionClass == "dysgranular"], "class_dysgranular")

# Bilateral soma regions only
bilateral_soma_regions <- meta_li %>%
  filter(!is.na(Soma_Region_chr), nzchar(Soma_Region_chr)) %>%
  group_by(Soma_Region_chr) %>%
  filter(n_distinct(Soma_Side) == 2L) %>%
  group_keys() %>%
  pull(Soma_Region_chr)

li_region_list <- lapply(bilateral_soma_regions, function(reg) {
  ids <- meta_li$NeuronID[meta_li$Soma_Region_chr == reg]
  calc_li_table(ids, paste0("region_", reg))
})
li_region <- bind_rows(li_region_list)

li_all_tab <- bind_rows(li_all, li_agran, li_dys, li_region) %>%
  filter(!is.na(LI), is.finite(LI))

cat("SQ3d: LI summary (top |LI| per stratum)\n")
li_top <- li_all_tab %>%
  group_by(stratum) %>%
  slice_max(order_by = abs_LI, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(stratum, desc(abs_LI))
print(li_top)

# Compact visualization: LI distributions by stratum family
li_plot_df <- li_all_tab %>%
  mutate(
    stratum_family = case_when(
      stratum == "all_neurons" ~ "all_neurons",
      stratum %in% c("class_agranular", "class_dysgranular") ~ "coarse_class",
      grepl("^region_", stratum) ~ "bilateral_regions",
      TRUE ~ "other"
    )
  )

ggplot(li_plot_df, aes(x = stratum_family, y = LI, fill = stratum_family)) +
  geom_violin(alpha = 0.45, trim = FALSE) +
  geom_boxplot(width = 0.18, outlier.size = 0.6, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.35) +
  labs(
    title = "SQ3d: Projection laterality index by target region",
    subtitle = "LI computed on per-neuron proportion-normalized projection profiles; strata: all, IA/ID classes, bilateral soma regions",
    x = "Stratum family",
    y = "Laterality Index (LI)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Region-level asymmetry bars (horizontal, bi-directional), ordered within stratum
li_bar_df <- li_all_tab %>%
  filter(significant) %>%
  filter(is.finite(LI), !is.na(target_region), !is.na(stratum)) %>%
  group_by(stratum) %>%
  arrange(LI, .by_group = TRUE) %>%
  mutate(
    target_region_stratum = paste0(stratum, "__", target_region),
    target_region_stratum = factor(target_region_stratum, levels = unique(target_region_stratum)),
    LI_dir = ifelse(LI >= 0, "L>R", "R>L")
  ) %>%
  ungroup()

if (nrow(li_bar_df)) {
  li_bar_plot <- ggplot(li_bar_df, aes(x = target_region_stratum, y = LI, fill = LI_dir)) +
    geom_col(width = 0.82) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.35) +
    coord_flip() +
    facet_grid(stratum ~ ., scales = "free_y", space = "free_y") +
    scale_x_discrete(labels = function(x) sub("^.*__", "", x)) +
    scale_fill_manual(values = c("L>R" = "#E74C3C", "R>L" = "#3498DB")) +
    labs(
      title = "SQ3d: Significant region-level projection asymmetry by stratum",
      subtitle = "BH-filtered only; horizontal bidirectional bars ordered by LI within each stratum",
      x = "Target brain region",
      y = "Laterality Index (LI)",
      fill = "Direction"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.text = element_text(face = "bold"),
      axis.text.y = element_text(size = 9),
      panel.spacing.y = unit(0.8, "lines")
    )
} else {
  li_bar_plot <- ggplot() +
    annotate("text", x = 0, y = 0, label = "No BH-significant LI regions in current strata") +
    theme_void() +
    labs(title = "SQ3d: Significant region-level projection asymmetry by stratum")
}

print(li_bar_plot)
save_plot(
  plot_obj = li_bar_plot,
  filename = "sq3d_projection_laterality_index_bars.png",
  width = 16,
  height = 14
)

save_plot(
  plot_obj = ggplot(li_plot_df, aes(x = stratum_family, y = LI, fill = stratum_family)) +
    geom_violin(alpha = 0.45, trim = FALSE) +
    geom_boxplot(width = 0.18, outlier.size = 0.6, alpha = 0.9) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.35) +
    labs(
      title = "SQ3d: Projection laterality index by target region",
      subtitle = "LI computed on per-neuron proportion-normalized projection profiles; strata: all, IA/ID classes, bilateral soma regions",
      x = "Stratum family",
      y = "Laterality Index (LI)"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none"),
  filename = "sq3d_projection_laterality_index.png",
  width = 11,
  height = 6.5
)

cleanup_vars(c(
  "meta_li", "region_names", "li_all", "li_agran", "li_dys", "bilateral_soma_regions",
  "li_region_list", "li_region", "li_all_tab", "li_top", "li_plot_df",
  "li_bar_df", "li_bar_plot", "eps_li", "mat_ipsi_prop_li", "tot_per_neuron_li", "ok_tot_li"
))
```

------------------------------------------------------------------------

## SQ4: Are **pairwise FNT distances** concordant with **pairwise projection distances**?

**Hypothesis**: Pairs closer in morphology tend to be closer in projection space (direction and strength of association are empirical).\
**Method**: On `common_ids`, **Mantel test** correlates the two distance matrices with a permutation null (proper inference for matrix-valued dependence). A raw **Spearman correlation** on upper-triangle pairs duplicates entries and **does not yield a valid independent-sample p-value**; we still report **rho** descriptively and use **Mantel's permuted p** for significance.

```{r sq4_mantel, eval=has_fnt}
ut <- upper.tri(fnt_sub)
fv <- fnt_sub[ut]
pv <- proj_dist[ut]

# Descriptive association only — pairs are not independent; do not interpret cor.test p-value
sp_rho <- suppressWarnings(cor(fv, pv, method = "spearman", use = "complete.obs"))
cat(sprintf("Spearman rho (upper triangle, descriptive) = %.3f\n", sp_rho))

mt <- mantel(as.dist(fnt_sub), as.dist(proj_dist), method = "spearman", permutations = PERM_N)
print(mt)
sq4_summary <- data.frame(
  statistic = unname(mt$statistic),
  p = unname(mt$signif)
) %>%
  mutate(
    significant = p < 0.05,
    sig_label = p_to_sig(p)
  )
print(sq4_summary)

ggplot(data.frame(FNT = fv, Proj = pv), aes(FNT, Proj)) +
  geom_hex(bins = 60) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  labs(title = "SQ4: Pairwise FNT vs pairwise projection distance (upper triangle)",
       subtitle = sprintf("Spearman rho=%.3f (descriptive); inference: Mantel p", sp_rho)) +
  theme_minimal()
```

------------------------------------------------------------------------

## SQ5: How to compare the two distance matrices side by side in heatmaps?

**Key idea**: Both heatmaps must use the **same row/column order** (here from a fused clustering signal). A third panel shows **\|**$z_{\text{FNT}} - z_{\text{Proj}}$\| per pair after z-scaling upper-triangle moments to highlight discrepancies.

```{r sq5_heatmaps, eval=has_fnt, fig.height=5, fig.width=14}
ut_h <- upper.tri(fnt_sub)
side_lab <- meta_fnt$Soma_Side[match(rownames(fnt_sub), meta_fnt$NeuronID)]
ha <- rowAnnotation(
  Side = side_lab,
  col = list(Side = c(L = "#E74C3C", R = "#3498DB"))
)

z_f <- (fnt_sub - mean(fnt_sub[ut_h])) / sd(fnt_sub[ut_h])
z_p <- (proj_dist - mean(proj_dist[ut_h])) / sd(proj_dist[ut_h])
diff_z <- abs(z_f - z_p)
diag(diff_z) <- NA

combo <- (fnt_sub / max(fnt_sub) + proj_dist / max(proj_dist)) / 2
ord <- hclust(as.dist(combo))$order

rcols <- function(x, utm) {
  colorRamp2(quantile(x[utm], c(0.05, 0.5, 0.95)), c("white", "orange", "red4"))
}

Heatmap(fnt_sub, name = "FNT", col = rcols(fnt_sub, ut_h),
        left_annotation = ha, cluster_rows = FALSE, cluster_columns = FALSE,
        row_order = ord, column_order = ord, show_row_names = FALSE, show_column_names = FALSE) +
Heatmap(proj_dist, name = "Proj", col = rcols(proj_dist, ut_h),
        left_annotation = ha, cluster_rows = FALSE, cluster_columns = FALSE,
        row_order = ord, column_order = ord, show_row_names = FALSE, show_column_names = FALSE) +
Heatmap(diff_z, name = "|Dz|",
        col = colorRamp2(quantile(diff_z, c(0.05, 0.5, 0.95), na.rm = TRUE), c("white", "tan", "brown4")),
        left_annotation = ha, cluster_rows = FALSE, cluster_columns = FALSE,
        row_order = ord, column_order = ord, show_row_names = FALSE, show_column_names = FALSE,
        column_title = "|z_FNT - z_Proj|")
```

------------------------------------------------------------------------

## SQ6: **Does laterality depend on neuron type?** (stratified PERMANOVA)

**Hypothesis**: Only some types show L vs R separation in multivariate projection space.

**Multiple testing**: $p$-values across types are corrected with **Benjamini–Hochberg FDR** in the printed table (`p_BH`).

```{r sq6_stratify}
res_type <- lapply(split(meta, meta$Neuron_Type), function(m) {
  if (nrow(m) < 6L || n_distinct(m$Soma_Side) < 2L) return(NULL)
  ix <- colnames(mat_ipsi) %in% m$NeuronID
  d <- vegdist(t(mat_ipsi[, ix, drop = FALSE]), method = "bray")
  s <- m$Soma_Side[match(labels(d), m$NeuronID)]
  a <- adonis2(d ~ s, permutations = PERM_N)
  data.frame(
    Neuron_Type = unique(m$Neuron_Type),
    n = nrow(m),
    R2 = a$R2[1],
    p = a$`Pr(>F)`[1]
  )
})
sq6_tab <- bind_rows(Filter(Negate(is.null), res_type))
if (nrow(sq6_tab)) {
  sq6_tab <- sq6_tab %>%
    mutate(
      p_BH = p.adjust(p, method = "BH"),
      significant_raw = p < 0.05,
      sig_label_raw = p_to_sig(p),
      significant_BH = p_BH < 0.05,
      sig_label_BH = p_to_sig(p_BH)
    )
}
print(sq6_tab)

if (nrow(sq6_tab)) {
  sq6_plot <- sq6_tab %>%
    mutate(sig_fdr = ifelse(p_BH < 0.05, "BH q < 0.05", "BH q >= 0.05"))
  ggplot(sq6_plot, aes(reorder(Neuron_Type, R2), R2, fill = sig_fdr)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
      values = c("BH q < 0.05" = "#E74C3C", "BH q >= 0.05" = "#bdc3c7"),
      name = NULL
    ) +
    labs(
      title = "SQ6: PERMANOVA R² for Side within each neuron type",
      subtitle = "Bar color: Benjamini–Hochberg FDR across neuron types",
      x = NULL
    ) +
    theme_minimal()
}
```

------------------------------------------------------------------------

# Export results

Interactive mode: final workbook export chunk removed to keep this notebook focused on chunk-by-chunk analysis and debugging.
