---
title: "渐进式假设检验：左右神经元投射差异"
subtitle: "Progressive Hypothesis Testing for Lateralization of Neuronal Projections"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    fig_width: 10
    fig_height: 8
---

<!-- Chunk-by-chunk in Cursor: install "R" (REditorSupport.r, see .vscode/extensions.json). Open repo as folder. Command Palette → "R: Create R Terminal". In an Rmd chunk: Ctrl+Shift+Enter = Run Current Chunk, Ctrl+Alt+P = run all chunks above, Ctrl+Enter = run selection/line. Full Knit needs Pandoc; integrated terminal sets RSTUDIO_PANDOC in .vscode/settings.json. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)

# Load required libraries (same as domain.Rmd)
library(readxl)
library(pheatmap)
library(ComplexHeatmap)
library(openxlsx)
library(circlize)
library(grid)
library(dplyr)
library(dendextend)
library(matrixStats)

# Additional libraries for hypothesis testing
library(vegan)       # For PERMANOVA and Mantel test
library(ggplot2)     # For additional plots
library(ggpubr)      # stat_compare_means (used before later chunks load it)
library(patchwork)   # For combining plots

# Set seed for reproducibility
set.seed(42)
```

# 数据加载 (参考 domain.Rmd)

```{r load_data}
# load data and precalculated clustering result (same as domain.Rmd)
file_results <- "../tables/251637_INS.xlsx"
cluster_info <- readRDS("../../data_output/clustering/clustering_results.rds")

dend <- as.dendrogram(cluster_info$hclust_obj)

# load sheets and combine into new one 
proj_df_ipsi <- read_excel(file_results, sheet = "Projection_Strength_L3_ipsi")
proj_df_contra <- read_excel(file_results, sheet = "Projection_Strength_L3_contra")
summary <- read_excel(file_results, sheet = "Summary")

# merge projection with cluster and soma info, _C -> cluster, _s soma
proj_df_ipsi_C <- merge(proj_df_ipsi, cluster_info$cluster_df %>% select(-BioType), by='NeuronID') %>% 
  relocate(c(Cluster, DendroOrder), .after=2) %>% arrange(DendroOrder)
proj_df_ipsi_CS <- merge(proj_df_ipsi_C, summary %>% select('NeuronID', 'Soma_Region', 'Soma_Side'), by='NeuronID') %>% 
  relocate(Soma_Region, Soma_Side, .after = 'NeuronID')

proj_df_contra_C <- merge(proj_df_contra, cluster_info$cluster_df %>% select(-BioType), by='NeuronID') %>% 
  relocate(c(Cluster, DendroOrder), .after=2) %>% arrange(DendroOrder)
proj_df_contra_CS <- merge(proj_df_contra_C, summary %>% select('NeuronID', 'Soma_Region', 'Soma_Side'), by='NeuronID') %>% 
  relocate(Soma_Region, Soma_Side, .after = 'NeuronID')

# Create indices (same as domain.Rmd)
idx_L <- proj_df_ipsi_CS$Soma_Side == "L"
idx_R <- proj_df_ipsi_CS$Soma_Side == "R"

cat(sprintf("distribution: L=%d, R=%d\n", sum(idx_L), sum(idx_R)))

# make matrices (same as domain.Rmd)
meta_cols <- 6

mat_ipsi <- t(proj_df_ipsi_CS %>%
  select(-(1:meta_cols)) %>%
  as.matrix())
colnames(mat_ipsi) <- proj_df_ipsi_CS$NeuronID
cat(sprintf("\n矩阵维度: %d 脑区 × %d 神经元\n", nrow(mat_ipsi), ncol(mat_ipsi)))

mat_contra <- t(proj_df_contra_CS %>%
  select(-(1:meta_cols)) %>%
  as.matrix())
colnames(mat_contra) <- proj_df_contra_CS$NeuronID

# Get L/R neuron column indices
ipsi_L_cols <- proj_df_ipsi_CS$Soma_Side == "L"
ipsi_R_cols <- proj_df_ipsi_CS$Soma_Side == "R"
contra_L_cols <- proj_df_contra_CS$Soma_Side == "L"
contra_R_cols <- proj_df_contra_CS$Soma_Side == "R"

# Create sub-matrices
mat_ipsi_L <- mat_ipsi[, ipsi_L_cols, drop = FALSE]
mat_ipsi_R <- mat_ipsi[, ipsi_R_cols, drop = FALSE]
mat_contra_L <- mat_contra[, contra_L_cols, drop = FALSE]
mat_contra_R <- mat_contra[, contra_R_cols, drop = FALSE]

# Prepare combined matrices
mat_ipsi_combined <- cbind(mat_ipsi_L, mat_ipsi_R)
mat_contra_combined <- cbind(mat_contra_L, mat_contra_R)

# Store for later use
proj_df_ipsi_s <- proj_df_ipsi_CS
proj_df_contra_s <- proj_df_contra_CS
```

# ========== 距离矩阵加载与 Spearman 相关分析 ==========

```{r load_and_spearman}

# 1. 加载 FNT 距离矩阵
fnt_df <- read_excel("../../data_output/clustering/raw_matrix_神经元矩阵.xlsx")
fnt_matrix <- as.matrix(fnt_df)
rownames(fnt_matrix) <- colnames(fnt_matrix) <- colnames(fnt_df)

cat("FNT matrix:", nrow(fnt_matrix), "x", ncol(fnt_matrix), "\n")
cat("First IDs:", head(colnames(fnt_matrix)), "\n")

# 2. 找到共同神经元
proj_ids <- colnames(mat_ipsi)
common_ids <- sort(intersect(proj_ids, colnames(fnt_matrix)))

cat("Projection neurons:", length(proj_ids), "\n")
cat("FNT neurons:", ncol(fnt_matrix), "\n")
cat("Common neurons:", length(common_ids), "\n")

# 3. 截取共同部分
fnt_sub <- fnt_matrix[common_ids, common_ids]
proj_idx <- match(common_ids, proj_ids)
mat_ipsi_sub <- mat_ipsi[, proj_idx]

# 4. 计算投射距离矩阵（Bray-Curtis）
proj_dist_raw <- as.matrix(vegdist(t(mat_ipsi_sub), method = "bray"))
rownames(proj_dist_raw) <- colnames(proj_dist_raw) <- common_ids

# 5. ========== Spearman 相关分析 ==========
# 提取上三角向量
fnt_vec <- fnt_sub[upper.tri(fnt_sub)]
proj_vec <- proj_dist_raw[upper.tri(proj_dist_raw)]

# 计算 Spearman 相关（使用 cor 函数）
spearman_result <- cor.test(fnt_vec, proj_vec, method = "spearman")

cat("\n=== Spearman Correlation ===\n")
cat(sprintf("Correlation (rho): %.3f\n", spearman_result$estimate))
cat(sprintf("p-value: %.4f\n", spearman_result$p.value))

# 置换检验（Permutation test）
n_perm <- 999
perm_stats <- replicate(n_perm, {
  shuffled_proj <- sample(proj_vec)
  cor(fnt_vec, shuffled_proj, method = "spearman", use = "complete.obs")
})

perm_p <- mean(abs(perm_stats) >= abs(spearman_result$estimate))
cat(sprintf("Permutation p-value: %.4f\n", perm_p))

# 保存结果
fnt_sub <- fnt_matrix[common_ids, common_ids]
proj_dist_matrix <- proj_dist_raw

# Mantel：在两套“神经元对–神经元对”距离上检验整体相关性（与下面 Spearman
# 向量相关检验目标一致；Mantel 通过置换得到 p 值， ecologists 常用写法）
mantel_fnt_proj <- mantel(
  as.dist(fnt_sub),
  as.dist(proj_dist_matrix),
  method = "spearman",
  permutations = 999
)
cat("\n=== Mantel: FNT distance ~ projection distance ===\n")
print(mantel_fnt_proj)
```

---

# Level 1: 描述性探索

## 1.1 样本分布

```{r level1_distribution}
# Distribution by side and neuron type
p1 <- ggplot(proj_df_ipsi_s, aes(x = Soma_Side, fill = Soma_Side)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(title = "Neuron Count by Hemisphere", x = "Soma Side", y = "Count") +
  theme_minimal()

p2 <- ggplot(proj_df_ipsi_s, aes(x = Neuron_Type, fill = Soma_Side)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +geom_text(
  stat = "count",
  aes(label = after_stat(count)),
  position = position_dodge(0.9),
  vjust = -0.6
) +
  labs(title = "Neuron Type Distribution by Side") +
  theme_minimal()

p1 + p2
```

## 1.1b Step 1：类型构成、平均投射强度、投射 PCA

```{r step1_composition_mean_pca}
# 类型 × 半球：计数 + 半球内构成比例
tab_type_side <- xtabs(~ Neuron_Type + Soma_Side, data = proj_df_ipsi_s)
cat("\n=== Cell counts (Type × Side) ===\n")
print(tab_type_side)
cat("\n=== Proportions within each hemisphere (columns sum to 1) ===\n")
print(round(prop.table(tab_type_side, margin = 2), 3))

# “Projection length” here = matrix row-sum strength (ipsi / contra / both)，与热图矩阵一致
tot_ipsi <- colSums(mat_ipsi)
tot_contra <- colSums(mat_contra)
tot_both <- tot_ipsi + tot_contra
step1_df <- data.frame(
  NeuronID = colnames(mat_ipsi),
  Soma_Side = proj_df_ipsi_s$Soma_Side,
  Neuron_Type = proj_df_ipsi_s$Neuron_Type,
  Total_ipsi = as.numeric(tot_ipsi),
  Total_contra = as.numeric(tot_contra),
  Total_both = as.numeric(tot_both),
  stringsAsFactors = FALSE
)

mean_by_type_side <- step1_df %>%
  group_by(Neuron_Type, Soma_Side) %>%
  summarise(
    n = n(),
    mean_ipsi = mean(Total_ipsi),
    mean_contra = mean(Total_contra),
    mean_total = mean(Total_both),
    .groups = "drop"
  )
cat("\n=== Mean row-sum (projection strength) by Type × Side ===\n")
print(as.data.frame(mean_by_type_side))

ggplot(step1_df, aes(x = Neuron_Type, y = Total_both, fill = Soma_Side)) +
  geom_boxplot(outlier.size = 1) +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(
    title = "Total projection strength (ipsi + contra), by type and hemisphere",
    y = "Sum of matrix entries per neuron"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# PCA：神经元为行、脑区为列（与原始投射矩阵同形转置）；PCoA/PCA 都降维，度量不同
X_pca <- t(mat_ipsi)
pca_proj <- prcomp(log1p(X_pca), center = TRUE, scale. = TRUE)
ve <- summary(pca_proj)$importance[2, 1:2] * 100
pca_axes <- data.frame(
  NeuronID = colnames(mat_ipsi),
  PC1 = pca_proj$x[, 1],
  PC2 = pca_proj$x[, 2],
  Soma_Side = proj_df_ipsi_s$Soma_Side,
  Neuron_Type = proj_df_ipsi_s$Neuron_Type
)

ggplot(pca_axes, aes(x = PC1, y = PC2, color = Soma_Side)) +
  geom_point(alpha = 0.75, size = 2.5) +
  stat_ellipse(level = 0.95, linewidth = 0.4) +
  scale_color_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(
    title = "PCA of ipsilateral projection profiles",
    subtitle = "One point per neuron; features = brain regions (log1p + scaled)",
    x = sprintf("PC1 (%.1f%% variance)", ve[1]),
    y = sprintf("PC2 (%.1f%% variance)", ve[2])
  ) +
  theme_minimal()

ggplot(pca_axes, aes(x = PC1, y = PC2, color = Soma_Side)) +
  geom_point(alpha = 0.75, size = 2) +
  facet_wrap(~Neuron_Type) +
  scale_color_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(
    title = "PCA by neuron type",
    x = sprintf("PC1 (%.1f%%)", ve[1]),
    y = sprintf("PC2 (%.1f%%)", ve[2])
  ) +
  theme_minimal()
```

## 1.2 投射强度统计

```{r level1_projection_stats}
# Calculate total projection per neuron
total_proj_ipsi <- colSums(mat_ipsi)
total_proj_contra <- colSums(mat_contra)

proj_summary_df <- data.frame(
  NeuronID = colnames(mat_ipsi),
  Side = c(rep("L", sum(ipsi_L_cols)), rep("R", sum(ipsi_R_cols))),
  Neuron_Type = proj_df_ipsi_s$Neuron_Type,
  Cluster = proj_df_ipsi_s$Cluster,
  Ipsi_Total = total_proj_ipsi,
  Contra_Total = total_proj_contra,
  Total_Proj = total_proj_ipsi + total_proj_contra,
  Ipsi_Ratio = total_proj_ipsi / (total_proj_ipsi + total_proj_contra + 0.001)
)

# Summary by side
summary_stats <- proj_summary_df %>%
  group_by(Side) %>%
  summarise(
    n = n(),
    mean_total = mean(Total_Proj),
    median_total = median(Total_Proj),
    mean_ipsi_ratio = mean(Ipsi_Ratio),
    .groups = "drop"
  )

print(summary_stats)

# Distribution plots
p3 <- ggplot(proj_summary_df, aes(x = Total_Proj, fill = Side)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(title = "Total Projection Distribution") +
  theme_minimal()


p4_final <- proj_summary_df %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Side) == 2) %>%  # 过滤只有单侧的类型
  ungroup() %>%
  ggplot(aes(x = Neuron_Type, y = Total_Proj, fill = Side)) +
  geom_boxplot(outlier.size = 1) +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  stat_compare_means(
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE
  ) +
  labs(title = "Total Projection (L vs R)") +
  theme_minimal()


p3 + p4_final
```
``` {r }
# 加载包
library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)

# Calculate total projection per neuron
total_proj_ipsi <- colSums(mat_ipsi)
total_proj_contra <- colSums(mat_contra)

proj_summary_df <- data.frame(
  NeuronID = colnames(mat_ipsi),
  Side = c(rep("L", sum(ipsi_L_cols)), rep("R", sum(ipsi_R_cols))),
  Neuron_Type = proj_df_ipsi_s$Neuron_Type,
  Cluster = proj_df_ipsi_s$Cluster,
  Ipsi_Total = total_proj_ipsi,
  Contra_Total = total_proj_contra,
  Total_Proj = total_proj_ipsi + total_proj_contra,
  Ipsi_Ratio = total_proj_ipsi / (total_proj_ipsi + total_proj_contra + 0.001)
)

# Summary by side
summary_stats <- proj_summary_df %>%
  group_by(Side) %>%
  summarise(
    n = n(),
    mean_total = mean(Total_Proj),
    median_total = median(Total_Proj),
    mean_ipsi_ratio = mean(Ipsi_Ratio),
    .groups = "drop"
  )

print(summary_stats)

# Distribution plots
p3 <- ggplot(proj_summary_df, aes(x = Total_Proj, fill = Side)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(title = "Total Projection Distribution") +
  theme_minimal()

# ---------------- 核心：箱线图 + 显示样本数量 n ----------------
p4_final <- proj_summary_df %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Side) == 2) %>%  # 过滤只有单侧的类型，防止报错
  ungroup() %>%
  
  ggplot(aes(x = Neuron_Type, y = Total_Proj, fill = Side)) +
  geom_boxplot(outlier.size = 1, width = 0.6) +
  scale_fill_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  
  # 统计检验：只显示星号，隐藏ns
  stat_compare_means(
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = TRUE,
    vjust = 0.8
  ) +
  
  # ✅ 关键：显示每组样本数量 n=xxx
  stat_summary(
    aes(label = after_stat(paste("n =", ..count..))),
    fun = "mean",
    geom = "text",
    vjust = -2.5,  # 位置上下调整
    hjust = 0.5,
    size = 3.5,
    color = "black"
  ) +
  
  labs(
    title = "Total Projection by Side & Neuron Type",
    x = "Neuron Type",
    y = "Total Projection"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 出图
p4_final

```



---

# Level 2: 整体差异检验

## 2.1 投射距离的 PERMANOVA

```{r level2_permanova}
# PERMANOVA on projection distance profiles
dist_matrix <- as.dist(proj_dist_matrix)
side_factor <- factor(proj_df_ipsi_s$Soma_Side[match(common_ids, proj_df_ipsi_s$NeuronID)])

adonis_result <- adonis2(dist_matrix ~ side_factor, permutations = 999, by = "margin")
print(adonis_result)

# PCoA
pcoa_result <- cmdscale(dist_matrix, k = 2, eig = TRUE)
pcoa_df <- data.frame(
  PC1 = pcoa_result$points[, 1],
  PC2 = pcoa_result$points[, 2],
  Side = side_factor,
  Neuron_Type = proj_df_ipsi_s$Neuron_Type[match(common_ids, proj_df_ipsi_s$NeuronID)]
)

var_exp <- round(100 * pcoa_result$eig / sum(pcoa_result$eig), 1)

ggplot(pcoa_df, aes(x = PC1, y = PC2, color = Side)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("L" = "#E74C3C", "R" = "#3498DB")) +
  labs(title = "PCoA of Projection Distance",
       subtitle = sprintf("PERMANOVA: R²=%.3f, p=%.4f", 
                          adonis_result$R2[1], adonis_result$`Pr(>F)`[1]),
       x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC2 (%.1f%%)", var_exp[2])) +
  theme_minimal()
```

---

# Level 3: 形态 vs 投射 Spearman 相关（核心）

## 3.1 Spearman 相关结果

```{r spearman_results}
cat("=== Spearman Correlation: Morphology vs Projection ===\n")
cat(sprintf("rho = %.3f\n", spearman_result$estimate))
cat(sprintf("p-value = %.4f\n", spearman_result$p.value))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", spearman_result$conf.int[1], spearman_result$conf.int[2]))
cat(sprintf("Permutation p = %.4f\n", perm_p))

# 解释
if (perm_p < 0.05) {
  if (abs(spearman_result$estimate) > 0.5) {
    cat("\n>>> Strong correlation: Morphology strongly predicts projection!\n")
  } else if (abs(spearman_result$estimate) > 0.3) {
    cat("\n>>> Moderate correlation: Some relationship exists.\n")
  } else {
    cat("\n>>> Weak correlation.\n")
  }
} else {
  cat("\n>>> No significant correlation.\n")
}
```

## 3.2 Spearman 散点图（秩次）

```{r spearman_scatter}
# 计算秩次用于可视化
fnt_rank <- rank(fnt_vec)
proj_rank <- rank(proj_vec)

plot_df <- data.frame(
  Morphology_Rank = fnt_rank,
  Projection_Rank = proj_rank
)

# 散点图
p_scatter <- ggplot(plot_df, aes(x = Morphology_Rank, y = Projection_Rank)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c() +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Spearman Correlation: Morphology vs Projection",
       subtitle = sprintf("rho = %.3f, p = %.4f (n = %d pairs)", 
                          spearman_result$estimate, perm_p, length(fnt_vec)),
       x = "Morphology Distance (rank)",
       y = "Projection Distance (rank)") +
  theme_minimal()

p_scatter
```

## 3.3 原始尺度散点图（参考）

```{r raw_scatter}
plot_df_raw <- data.frame(
  Morphology = fnt_vec,
  Projection = proj_vec
)

p_raw <- ggplot(plot_df_raw, aes(x = Morphology, y = Projection)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Original Scale (for reference)",
       subtitle = sprintf("Spearman rho = %.3f", spearman_result$estimate),
       x = "FNT Distance (raw)",
       y = "Projection Distance (Bray-Curtis)") +
  theme_minimal()

p_raw
```

## 3.4 分面分析

```{r facet_analysis}
# 获取神经元类型和side
neuron_types <- proj_df_ipsi_s$Neuron_Type[match(common_ids, proj_df_ipsi_s$NeuronID)]
neuron_sides <- proj_df_ipsi_s$Soma_Side[match(common_ids, proj_df_ipsi_s$NeuronID)]

# 创建配对数据
n <- length(common_ids)
pair_data <- data.frame(
  Morphology = numeric(),
  Projection = numeric(),
  SameType = logical(),
  SameSide = logical()
)

for(i in 1:(n-1)) {
  for(j in (i+1):n) {
    pair_data <- rbind(pair_data, data.frame(
      Morphology = fnt_sub[i, j],
      Projection = proj_dist_matrix[i, j],
      SameType = neuron_types[i] == neuron_types[j],
      SameSide = neuron_sides[i] == neuron_sides[j]
    ))
  }
}

# 计算分层 Spearman 相关
cor_same <- cor(pair_data$Morphology[pair_data$SameType], 
                pair_data$Projection[pair_data$SameType], 
                method = "spearman")
cor_diff <- cor(pair_data$Morphology[!pair_data$SameType], 
                pair_data$Projection[!pair_data$SameType], 
                method = "spearman")

# 分面图
p_facet <- ggplot(pair_data, aes(x = Morphology, y = Projection)) +
  geom_hex(bins = 30) +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ SameType, 
             labeller = labeller(SameType = c("TRUE" = "Same Type", 
                                               "FALSE" = "Different Type"))) +
  labs(title = "Spearman Correlation by Cell Type",
       subtitle = sprintf("Same type: rho=%.3f | Diff type: rho=%.3f", cor_same, cor_diff),
       x = "Morphology Distance",
       y = "Projection Distance") +
  theme_minimal()

p_facet

cat("\nSpearman correlation (same type):", round(cor_same, 3), "\n")
cat("Spearman correlation (diff type):", round(cor_diff, 3), "\n")
```

---

# Level 4: 热图对比（相同排序才有可比性）

并排热图要在**同一行/列顺序**下阅读：各自 `cluster_rows = TRUE` 会得到两套不同排序，视觉上无法对齐比较。

做法：用两套距离的**归一化平均**做层次聚类，得到一个顺序 `ord`，两张图共用 `row_order = column_order = ord`。第三张为**逐元素差异**（各自在上三角上 z-score 后取绝对差），大值 = 两种距离最不「同变」的神经元对。

```{r heatmap_comparison}
library(ComplexHeatmap)
side_info <- neuron_sides
ut <- upper.tri(fnt_sub)
z_f <- (fnt_sub - mean(fnt_sub[ut])) / sd(fnt_sub[ut])
z_p <- (proj_dist_matrix - mean(proj_dist_matrix[ut])) / sd(proj_dist_matrix[ut])
diff_z <- abs(z_f - z_p)
diag(diff_z) <- NA

fnt_n <- fnt_sub / max(fnt_sub)
proj_n <- proj_dist_matrix / max(proj_dist_matrix)
combo <- (fnt_n + proj_n) / 2
ord <- hclust(as.dist(combo))$order

fnt_col <- colorRamp2(
  quantile(fnt_sub[ut], probs = c(0.05, 0.5, 0.95)),
  c("white", "orange", "red4")
)
proj_col <- colorRamp2(
  quantile(proj_dist_matrix[ut], probs = c(0.05, 0.5, 0.95)),
  c("white", "steelblue", "navy")
)
diff_col <- colorRamp2(
  quantile(diff_z, probs = c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("white", "tan", "brown4")
)

ha <- rowAnnotation(
  Side = side_info,
  col = list(Side = c("L" = "#E74C3C", "R" = "#3498DB"))
)

ht1 <- Heatmap(
  fnt_sub,
  name = "FNT",
  col = fnt_col,
  left_annotation = ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_order = ord,
  column_order = ord,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "FNT distance"
)
ht2 <- Heatmap(
  proj_dist_matrix,
  name = "Proj",
  col = proj_col,
  left_annotation = ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_order = ord,
  column_order = ord,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "Projection distance"
)
ht3 <- Heatmap(
  diff_z,
  name = "|Dz|",
  col = diff_col,
  left_annotation = ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_order = ord,
  column_order = ord,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "|z_FNT - z_Proj|"
)

ht1 + ht2 + ht3
```
</think>
Fixing a bad partial edit — reading the file and applying a clean replacement.

<｜tool▁calls▁begin｜><｜tool▁call▁begin｜>
Read

---

# Level 5: 细胞类型特异性

```{r celltype_analysis}
# PERMANOVA by neuron type
celltype_results <- data.frame(
  Neuron_Type = character(),
  R2 = numeric(),
  p_value = numeric(),
  n = integer()
)

for(ntype in unique(neuron_types)) {
  idx <- neuron_types == ntype
  if(sum(idx) < 3) next
  
  sub_dist <- as.dist(proj_dist_matrix[idx, idx])
  sub_side <- factor(neuron_sides[idx])
  
  if(length(unique(sub_side)) < 2) next
  
  res <- adonis2(sub_dist ~ sub_side, permutations = 999)
  
  celltype_results <- rbind(celltype_results, data.frame(
    Neuron_Type = ntype,
    R2 = res$R2[1],
    p_value = res$`Pr(>F)`[1],
    n = sum(idx)
  ))
}

print(celltype_results)

ggplot(celltype_results, aes(x = Neuron_Type, y = R2, fill = p_value < 0.05)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("n=%d", n)), vjust = -0.5) +
  scale_fill_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#95A5A6")) +
  labs(title = "Lateralization by Neuron Type",
       y = "PERMANOVA R²") +
  theme_minimal()
```

---

# 结果汇总

```{r summary}
summary_report <- data.frame(
  Level = c("Level 1", "Level 2", "Level 3", "Level 4"),
  Test = c("Descriptive", "PERMANOVA (Projection)", 
           "Spearman (Morph vs Proj)", "Cell-type"),
  Result = c(
    sprintf("L=%d, R=%d", sum(idx_L), sum(idx_R)),
    sprintf("R²=%.3f, p=%.4f", adonis_result$R2[1], adonis_result$`Pr(>F)`[1]),
    sprintf("rho=%.3f, p=%.4f", spearman_result$estimate, perm_p),
    sprintf("%d types tested", nrow(celltype_results))
  ),
  Interpretation = c(
    "Sample sizes",
    ifelse(adonis_result$`Pr(>F)`[1] < 0.05, "Significant side diff", "No side diff"),
    ifelse(perm_p < 0.05, 
           ifelse(abs(spearman_result$estimate) > 0.5, "Strong morph-proj link", "Weak link"),
           "No morph-proj link"),
    "Type-specific lateralization"
  )
)

print(summary_report)

# Save
write.xlsx(list(
  Summary = summary_report,
  Spearman_Test = data.frame(
    Rho = spearman_result$estimate,
    P_value = spearman_result$p.value,
    Permutation_p = perm_p,
    CI_lower = spearman_result$conf.int[1],
    CI_upper = spearman_result$conf.int[2]
  ),
  CellType_Tests = celltype_results
), file = "laterality_spearman_results.xlsx", overwrite = TRUE)

cat("\nResults saved to: laterality_spearman_results.xlsx\n")
```

---

**分析完成**: `r Sys.time()`

**方法**: Spearman 秩相关分析形态与投射距离的关系
