---
title: "laterlity"
output: html_document
---

---
title: "Insula laterality: analyses by scientific question"
subtitle: "Left vs right insula neurons (structured by question)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    fig_width: 12
    fig_height: 8
---

This report is organized by **scientific question**: for each question we state a one-line hypothesis, prepare data, plot, and run statistics. Edit all file paths in the `config` chunk below (paths are relative to this Rmd in `R_analysis/scripts/`).

```{r config, include=TRUE}
# --- Paths (relative to this Rmd: R_analysis/scripts/) ---
PATH_INS_XLSX    <- "../tables/251637_INS.xlsx"
PATH_CLUSTER_RDS <- "../../data_output/clustering/clustering_results.rds"
# FNT = feature / morphology distance matrix (neuron x neuron); Chinese filename from lab export
PATH_FNT_XLSX    <- c(
  "../../data_output/clustering/raw_matrix_神经元矩阵.xlsx",
  "../../raw_matrix_神经元矩阵.xlsx",
  "raw_matrix_神经元矩阵.xlsx"
)
PATH_FNT_XLSX <- PATH_FNT_XLSX[file.exists(PATH_FNT_XLSX)][1]

PERM_N <- 999L
set.seed(42)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = TRUE, fig.align = "center")
library(readxl)
library(dplyr)
library(ggplot2)
library(vegan)
library(ComplexHeatmap)
library(circlize)
library(ggpubr)
library(openxlsx)
library(scales)
library(patchwork)

# Automatic significance labeling used across SQ1/SQ2 tables
p_to_sig <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}
```

------------------------------------------------------------------------

# Preparation: align projection matrices with neuron metadata

```{r prep}
cluster_info <- readRDS(PATH_CLUSTER_RDS)
proj_ipsi    <- read_excel(PATH_INS_XLSX, sheet = "Projection_Strength_L3_ipsi")
proj_contra  <- read_excel(PATH_INS_XLSX, sheet = "Projection_Strength_L3_contra")
summary_sh   <- read_excel(PATH_INS_XLSX, sheet = "Summary")

# Hemisphere tag in atlas string (e.g. CL_Ia/Id, CR_Ial) — strip for clarity; L/R stays in Soma_Side
.strip_soma_region_prefix <- function(x) {
  v <- trimws(as.character(x))
  w <- sub("^(CL|CR)_", "", v, perl = TRUE)
  ifelse(nzchar(w), w, v)
}

meta_cols <- 6L

proj_ipsi <- proj_ipsi %>%
  merge(cluster_info$cluster_df %>% dplyr::select(-BioType), by = "NeuronID") %>%
  merge(summary_sh %>% dplyr::select(NeuronID, Soma_Region, Soma_Side), by = "NeuronID") %>%
  relocate(NeuronID, Soma_Region, Soma_Side, Neuron_Type, Cluster, DendroOrder) %>%
  mutate(Soma_Region = factor(.strip_soma_region_prefix(Soma_Region)))

proj_contra <- proj_contra %>%
  merge(cluster_info$cluster_df %>% dplyr::select(-BioType), by = "NeuronID") %>%
  merge(summary_sh %>% dplyr::select(NeuronID, Soma_Region, Soma_Side), by = "NeuronID") %>%
  relocate(NeuronID, Soma_Region, Soma_Side, Neuron_Type, Cluster, DendroOrder) %>%
  mutate(Soma_Region = factor(.strip_soma_region_prefix(Soma_Region)))

mat_ipsi <- t(as.matrix(proj_ipsi %>% dplyr::select(-(1:meta_cols))))
colnames(mat_ipsi) <- proj_ipsi$NeuronID

mat_contra <- t(as.matrix(proj_contra %>% dplyr::select(-(1:meta_cols))))
colnames(mat_contra) <- proj_contra$NeuronID

stopifnot(all.equal(proj_ipsi$NeuronID, proj_contra$NeuronID))

meta <- proj_ipsi %>%
  transmute(
    NeuronID = NeuronID,
    Soma_Region = droplevels(Soma_Region),
    Soma_Side = factor(Soma_Side, levels = c("L", "R")),
    Neuron_Type = factor(Neuron_Type),
    Cluster = Cluster
  )

cat(sprintf("Neurons: n=%d | L=%d R=%d\n",
            nrow(meta), sum(meta$Soma_Side == "L"), sum(meta$Soma_Side == "R")))
cat(sprintf("Soma regions (CL_/CR_ stripped in prep): %d levels\n", nlevels(meta$Soma_Region)))
```

### Matrix sparsity diagnostics

```{r matrix_sparsity_report}
matrix_sparsity <- function(m, name) {
  nz_per_neuron <- colSums(m > 0, na.rm = TRUE)
  nz_per_region <- rowSums(m > 0, na.rm = TRUE)
  data.frame(
    matrix = name,
    n_regions = nrow(m),
    n_neurons = ncol(m),
    zero_fraction = mean(m == 0, na.rm = TRUE),
    nonzero_fraction = mean(m > 0, na.rm = TRUE),
    mean_nonzero_targets_per_neuron = mean(nz_per_neuron, na.rm = TRUE),
    median_nonzero_targets_per_neuron = median(nz_per_neuron, na.rm = TRUE),
    mean_neurons_per_region = mean(nz_per_region, na.rm = TRUE),
    median_neurons_per_region = median(nz_per_region, na.rm = TRUE)
  )
}

sparsity_report <- bind_rows(
  matrix_sparsity(mat_ipsi, "ipsi_strength"),
  matrix_sparsity(mat_contra, "contra_strength"),
  {
    # Align by shared region and neuron IDs before summing.
    common_regions <- intersect(rownames(mat_ipsi), rownames(mat_contra))
    common_neurons <- intersect(colnames(mat_ipsi), colnames(mat_contra))
    if (length(common_regions) == 0L || length(common_neurons) == 0L) {
      warning("No overlap between ipsi and contra matrices; total_strength sparsity uses ipsi only.")
      matrix_sparsity(mat_ipsi, "total_strength_fallback_ipsi")
    } else {
      mat_ipsi_al <- mat_ipsi[common_regions, common_neurons, drop = FALSE]
      mat_contra_al <- mat_contra[common_regions, common_neurons, drop = FALSE]
      matrix_sparsity(mat_ipsi_al + mat_contra_al, "total_strength")
    }
  }
)

print(sparsity_report)
cat(sprintf("ipsi matrix sparsity: %.1f%% zeros\n", 100 * sparsity_report$zero_fraction[sparsity_report$matrix == "ipsi_strength"]))
cat(sprintf("contra matrix sparsity: %.1f%% zeros\n", 100 * sparsity_report$zero_fraction[sparsity_report$matrix == "contra_strength"]))
```

Whether **FNT (morphology / feature-neuron distance)** is available controls whether chunks SQ4–SQ5 evaluate (`has_fnt`):

```{r prep_fnt}
has_fnt <- !is.na(PATH_FNT_XLSX) && nzchar(PATH_FNT_XLSX)
if (has_fnt) {
  fnt_full <- readxl::read_excel(PATH_FNT_XLSX)
  fnt_mat <- as.matrix(fnt_full)
  rownames(fnt_mat) <- colnames(fnt_mat) <- colnames(fnt_full)
  common_ids <- sort(intersect(colnames(mat_ipsi), colnames(fnt_mat)))
  fnt_sub <- fnt_mat[common_ids, common_ids, drop = FALSE]
  mat_ipsi_fnt <- mat_ipsi[, common_ids, drop = FALSE]
  proj_dist <- as.matrix(vegdist(t(mat_ipsi_fnt), method = "bray"))
  rownames(proj_dist) <- colnames(proj_dist) <- common_ids
  meta_fnt <- meta %>% filter(NeuronID %in% common_ids)
  cat(sprintf("FNT ok: |common|=%d (of %d projection neurons)\n",
              length(common_ids), ncol(mat_ipsi)))
} else {
  common_ids <- character(0)
  fnt_sub <- proj_dist <- NULL
  meta_fnt <- meta
  cat("FNT file not found: skip SQ4–SQ5 or set PATH_FNT_XLSX in config.\n")
}
```

------------------------------------------------------------------------

# Part A: Count-based Laterality Analyses

This section is specialized for **count / composition statistics** (contingency tables, proportions, chi-square/Fisher).

## SQ1: Count-based composition tests block (SQ1a + SQ1b)

This unified SQ1 block includes both count-statistics components:
- **SQ1a (L/R):** `Neuron_Type × Soma_Side`
- **SQ1b (Region):** `Neuron_Type × Soma_Region`
- **SQ1c (Region vs Side count):** `Soma_Region × Soma_Side`

For both tables, the primary inference is Pearson $\chi^2$; Fisher's exact test is used as a fallback when expected counts are small and the table size is feasible.

**Composition display**: Plots use **proportions within each side** ($n_{\text{type,side}} / N_{\text{side}}$), so unequal total neuron counts between L and R do not distort the visual comparison of cell-type mix. **Atlas `Soma_Region` strings are stripped of `CL_` / `CR_` in prep** so labels match across hemispheres; side comes from **`Soma_Side`**. SQ1 includes **raw counts per `Soma_Region` dodged by L/R** for region-wise sampling clarity.

```{r sq1_composition, fig.width=14, fig.height=12.5}
tab <- xtabs(~ Neuron_Type + Soma_Side, data = meta)
print(tab)
# Proportion of each type within each hemisphere (columns sum to 1)
prop_side <- prop.table(tab, margin = 2)
print(round(prop_side, 4))

sq1_by_side <- meta %>%
  dplyr::count(Soma_Side, Neuron_Type, name = "n") %>%
  group_by(Soma_Side) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

chi <- suppressWarnings(chisq.test(tab))
print(chi)

# Rule of thumb: expected frequencies < 5 undermine chi-square asymptotics (Cochran's guideline).
exp_ok <- all(chi$expected >= 5, na.rm = TRUE)
if (!exp_ok) {
  message("SQ1: Some expected counts < 5; chi-square p-value may be inaccurate; running Fisher's exact test.")
}
fish_tab <- NULL
if (!exp_ok && is.matrix(tab) && prod(dim(tab)) <= 1e6) {
  fish_tab <- tryCatch(fisher.test(tab), error = function(e) NULL)
  if (!is.null(fish_tab)) print(fish_tab)
}

sq1_lr_summary <- data.frame(
  table_name = "Neuron_Type_x_Soma_Side",
  preferred_test = if (exp_ok) "chi-square" else if (!is.null(fish_tab)) "fisher" else "chi-square_with_small_expected",
  p_value = if (exp_ok || is.null(fish_tab)) chi$p.value else fish_tab$p.value
) %>%
  mutate(
    significant = p_value < 0.05,
    sig_label = p_to_sig(p_value)
  )
print(sq1_lr_summary)

side_cols <- c(L = "#E74C3C", R = "#3498DB")

# Region-level count tables used by SQ1b/SQ1c statistics and plots
meta_sq1_region <- meta %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region)))) %>%
  mutate(
    Soma_Region = droplevels(Soma_Region),
    Neuron_Type = droplevels(Neuron_Type)
  )

# Export helpers (kept for compatibility with export chunk names)
sq1b_tab <- NULL
sq1b_prop_region <- NULL
sq1b_chi <- NULL
sq1b_exp_ok <- NA
sq1b_fish_tab <- NULL
sq1_region_side_tab <- NULL
sq1_region_side_chi <- NULL
sq1_region_side_exp_ok <- NA
sq1_region_side_fisher <- NULL

if (nrow(meta_sq1_region) >= 4L && nlevels(meta_sq1_region$Soma_Region) >= 2L) {
  # SQ1c: Region vs Side (count table)
  sq1_region_side_tab <- xtabs(~ Soma_Region + Soma_Side, data = meta_sq1_region)
  print(sq1_region_side_tab)
  sq1_region_side_chi <- suppressWarnings(chisq.test(sq1_region_side_tab))
  print(sq1_region_side_chi)
  sq1_region_side_exp_ok <- all(sq1_region_side_chi$expected >= 5, na.rm = TRUE)
  if (!sq1_region_side_exp_ok && prod(dim(sq1_region_side_tab)) <= 1e6) {
    sq1_region_side_fisher <- tryCatch(fisher.test(sq1_region_side_tab), error = function(e) NULL)
    if (!is.null(sq1_region_side_fisher)) print(sq1_region_side_fisher)
  }
  sq1_region_side_summary <- data.frame(
    table_name = "Soma_Region_x_Soma_Side",
    preferred_test = if (sq1_region_side_exp_ok) "chi-square" else if (!is.null(sq1_region_side_fisher)) "fisher" else "chi-square_with_small_expected",
    p_value = if (sq1_region_side_exp_ok || is.null(sq1_region_side_fisher)) sq1_region_side_chi$p.value else sq1_region_side_fisher$p.value
  ) %>%
    mutate(
      significant = p_value < 0.05,
      sig_label = p_to_sig(p_value)
    )
  print(sq1_region_side_summary)

  # SQ1b: Region vs Type (count table)
  if (nlevels(meta_sq1_region$Neuron_Type) >= 2L) {
    sq1b_tab <- xtabs(~ Neuron_Type + Soma_Region, data = meta_sq1_region)
    print(sq1b_tab)
    print(addmargins(sq1b_tab))
    sq1b_prop_region <- prop.table(sq1b_tab, margin = 2)
    print(round(sq1b_prop_region, 4))

    sq1b_chi <- suppressWarnings(chisq.test(sq1b_tab))
    print(sq1b_chi)
    sq1b_exp_ok <- all(sq1b_chi$expected >= 5, na.rm = TRUE)
    if (!sq1b_exp_ok && prod(dim(sq1b_tab)) <= 1e6) {
      sq1b_fish_tab <- tryCatch(fisher.test(sq1b_tab), error = function(e) NULL)
      if (!is.null(sq1b_fish_tab)) print(sq1b_fish_tab)
    }
    sq1_region_type_summary <- data.frame(
      table_name = "Neuron_Type_x_Soma_Region",
      preferred_test = if (sq1b_exp_ok) "chi-square" else if (!is.null(sq1b_fish_tab)) "fisher" else "chi-square_with_small_expected",
      p_value = if (sq1b_exp_ok || is.null(sq1b_fish_tab)) sq1b_chi$p.value else sq1b_fish_tab$p.value
    ) %>%
      mutate(
        significant = p_value < 0.05,
        sig_label = p_to_sig(p_value)
      )
    print(sq1_region_type_summary)
  }
}

n_L <- sum(meta$Soma_Side == "L")
n_R <- sum(meta$Soma_Side == "R")

sq1_p1 <- ggplot(meta, aes(x = Soma_Side, fill = Soma_Side)) +
  geom_bar(width = 0.6) +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.35,
    size = 4
  ) +
  scale_fill_manual(values = side_cols) +
  labs(
    title = "SQ1: Total neurons per hemisphere (raw N)",
    subtitle = sprintf("N_L = %d, N_R = %d", n_L, n_R),
    x = "Soma side",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Within each side, proportion per type (sums to 100% for L and separately for R)
sq1_p2 <- ggplot(sq1_by_side, aes(x = Neuron_Type, y = prop, fill = Soma_Side)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.75) +
  geom_text(
    aes(label = sprintf("%d\n(%s)", n, percent(prop, accuracy = 0.1))),
    position = position_dodge(width = 0.85),
    vjust = -0.1,
    size = 2.85,
  ) +
  scale_fill_manual(values = side_cols, name = "Side") +
  scale_y_continuous(labels = label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "SQ1: Cell-type mix standardized by side (proportion of N_side)",
    subtitle = "Each bar = n_type,side / N_hemisphere; compare L vs R at equal total weight",
    x = "Neuron type",
    y = "Proportion within hemisphere"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

# 100% stacked: full composition per hemisphere on one axis
sq1_p3 <- ggplot(meta, aes(x = Soma_Side, fill = Neuron_Type)) +
  geom_bar(position = "fill", width = 0.55) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "SQ1: Composition (100% per side)",
    x = "Soma side",
    y = "Proportion within hemisphere"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

# Soma_Region hemisphere-neutral in prep; compare L vs R sample depth per label

n_drop_region <- nrow(meta) - nrow(meta_sq1_region)
sq1_p4_sub <- if (n_drop_region > 0L) {
  sprintf("Excluded %d neuron(s) with missing/blank Soma_Region.", n_drop_region)
} else {
  "All neurons have a non-blank Soma_Region."
}

sq1_p3b <- NULL
if (nrow(meta_sq1_region)) {
  sq1_p3b <- ggplot(meta_sq1_region, aes(x = Soma_Region, fill = Soma_Side)) +
    geom_bar(position = position_dodge(width = 0.85), width = 0.78) +
    geom_text(
      stat = "count",
      aes(label = after_stat(count), group = Soma_Side),
      position = position_dodge(width = 0.85),
      vjust = -0.25,
      size = 3.2
    ) +
    scale_fill_manual(values = side_cols, name = "Soma side") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.14))) +
    labs(
      title = "SQ1: Neuron count per soma region, by hemisphere",
      subtitle = paste0(
        "Dodged bars: L vs R within each Soma_Region. ",
        sq1_p4_sub
      ),
      x = "Soma_Region",
      y = "Count"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
}

sq1_top <- (sq1_p1 | sq1_p2) + plot_layout(widths = c(1, 2.25), guides = "collect")
if (!is.null(sq1_p3b)) {
  sq1_top / sq1_p3 / sq1_p3b +
    plot_layout(heights = c(1.05, 0.72, 0.92)) &
    theme(plot.title = element_text(size = 11.5))
} else {
  sq1_top / sq1_p3 +
    plot_layout(heights = c(1.05, 0.85)) &
    theme(plot.title = element_text(size = 11.5))
}
```

------------------------------------------------------------------------

# Part B: Projection-Strength Laterality Analyses

This section is specialized for **projection strength** (continuous outcomes; Wilcoxon / distributional comparisons / multivariate distance models).

## SQ2: Projection-strength counterparts to SQ1 (L/R and Soma_Region)

**SQ2a (L/R counterpart):** Within each `Neuron_Type`, compare total projection strength between L and R (`wilcox.test(total ~ Soma_Side)` with BH correction across types).\
**SQ2b (Region counterpart):** In **bilateral** regions (both L and R present), compare total strength by hemisphere overall within each region and within `Soma_Region × Neuron_Type` strata (BH across strata).

**Note**: `total` is a matrix-derived projection-strength summary, not necessarily physical axon length unless source columns are length-based.

```{r sq2_univariate_strength, fig.width=18, fig.height=10}
strength <- data.frame(
  NeuronID = meta$NeuronID,
  Soma_Side = meta$Soma_Side,
  Neuron_Type = meta$Neuron_Type,
  Soma_Region = meta$Soma_Region,
  ipsi = colSums(mat_ipsi),
  contra = colSums(mat_contra),
  total = colSums(mat_ipsi) + colSums(mat_contra),
  stringsAsFactors = FALSE
)

side_cols <- c(L = "#E74C3C", R = "#3498DB")

strength_cmp <- strength %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Soma_Side) == 2L) %>%
  ungroup()

# Left panel: only types with both L and R neurons (drop types present on one side only, e.g. all-ITc-on-one-side)
sq2_density <- ggplot(strength_cmp, aes(x = total, fill = Soma_Side)) +
  geom_density(alpha = 0.45, linewidth = 0.3) +
  facet_wrap(~Neuron_Type, scales = "free") +
  scale_fill_manual(values = side_cols, name = "Side") +
  labs(
    title = "SQ2: Distribution of total strength (ipsi + contra)",
    x = "Total projection strength",
    y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 11, face = "plain"),
    axis.text = element_text(size = 10)
  )

sq2_n_lab <- strength_cmp %>%
  group_by(Neuron_Type, Soma_Side) %>%
  summarise(
    n = n(),
    y = max(total, na.rm = TRUE) * 1.08,
    .groups = "drop"
  )

# Right panel: matches laterality_hypothesis_testing.Rmd style — one figure, x = neuron type
sq2_box <- ggplot(strength_cmp, aes(x = Neuron_Type, y = total, fill = Soma_Side)) +
  geom_boxplot(
    position = position_dodge(width = 0.85),
    width = 0.62,
    alpha = 0.88,
    outlier.size = 1
  ) +
  geom_text(
    data = sq2_n_lab,
    aes(x = Neuron_Type, y = y, label = paste0("n=", n), group = Soma_Side),
    position = position_dodge(width = 0.85),
    size = 3.2, 
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = side_cols, name = "Side") +
    # Two-sample Wilcoxon (Mann–Whitney); uncorrected p per type — see table for BH-adjusted q
    stat_compare_means(
    aes(group = Soma_Side),
    method = "wilcox.test",
    label = "p.signif",
    hide.ns = FALSE,
    tip.length = 0.02,
    bracket.size = 0.22,
    step.increase = 0.06
  ) +
  labs(
    title = "SQ2: Total projection by type & side (Wilcoxon L vs R within each type; plot p uncorrected)",
    x = "Neuron type",
    y = "Total projection strength"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sq2_row1 <- (sq2_density | sq2_box) +
  plot_layout(widths = c(1, 1), guides = "collect") &
  theme(plot.title = element_text(size = 12))

strength_reg <- strength %>%
  filter(!is.na(Soma_Region), nzchar(trimws(as.character(Soma_Region)))) %>%
  mutate(Soma_Region = droplevels(factor(Soma_Region)))

bilateral_region_levels <- strength_reg %>%
  group_by(Soma_Region) %>%
  filter(n_distinct(Soma_Side) == 2L) %>%
  group_keys() %>%
  pull(Soma_Region)

strength_bilateral_regions <- strength_reg %>%
  filter(Soma_Region %in% bilateral_region_levels) %>%
  mutate(Soma_Region = droplevels(Soma_Region))

sq2_region_density <- NULL
wt_region_overall <- NULL
wt_region_type <- NULL

if (nrow(strength_bilateral_regions) >= 4L && length(bilateral_region_levels)) {
  sq2_region_density <- ggplot(strength_bilateral_regions, aes(x = total, fill = Soma_Side)) +
    geom_density(alpha = 0.45, linewidth = 0.3) +
    facet_wrap(~ Soma_Region, nrow = 1L, scales = "free") +
    scale_fill_manual(values = side_cols, name = "Side") +
    labs(
      title = "SQ2 (regions): Total strength by hemisphere, bilateral Soma_Region only",
      subtitle = "All neuron types pooled within each region; facets = regions with both L and R",
      x = "Total projection strength",
      y = "Density"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text = element_text(size = 10, face = "plain"),
      legend.position = "bottom"
    )

  wt_region_overall <- strength_bilateral_regions %>%
    group_by(Soma_Region) %>%
    filter(n_distinct(Soma_Side) == 2L, n() >= 4L) %>%
    summarise(
      p_wilcox_total = tryCatch(
        wilcox.test(total ~ Soma_Side)$p.value,
        error = function(e) NA_real_
      ),
      n = n(),
      n_L = sum(Soma_Side == "L"),
      n_R = sum(Soma_Side == "R"),
      .groups = "drop"
    ) %>%
    mutate(
      p_wilcox_BH = p.adjust(p_wilcox_total, method = "BH"),
      significant = p_wilcox_BH < 0.05,
      sig_label = p_to_sig(p_wilcox_BH)
    )

  wt_region_type <- strength_bilateral_regions %>%
    group_by(Soma_Region, Neuron_Type) %>%
    filter(n_distinct(Soma_Side) == 2L, n() >= 4L) %>%
    summarise(
      p_wilcox_total = tryCatch(
        wilcox.test(total ~ Soma_Side)$p.value,
        error = function(e) NA_real_
      ),
      n = n(),
      n_L = sum(Soma_Side == "L"),
      n_R = sum(Soma_Side == "R"),
      .groups = "drop"
    ) %>%
    mutate(
      p_wilcox_BH = p.adjust(p_wilcox_total, method = "BH"),
      significant = p_wilcox_BH < 0.05,
      sig_label = p_to_sig(p_wilcox_BH)
    )

  cat("SQ2: Wilcoxon total ~ Side, bilateral regions (all types pooled)\n")
  print(wt_region_overall)
  cat("SQ2: Wilcoxon total ~ Side within Soma_Region × Neuron_Type (BH across strata)\n")
  print(wt_region_type)

  (sq2_row1) / sq2_region_density +
    plot_layout(heights = c(1, 0.68), guides = "collect") &
    theme(plot.title = element_text(size = 11))
} else {
  message("SQ2: No bilateral soma regions or too few neurons; skipping region-level strength plots.")
  sq2_row1
}

# Per-type Wilcoxon on total strength (no region stratification); BH across types
wt <- strength %>%
  group_by(Neuron_Type) %>%
  filter(n_distinct(Soma_Side) == 2, n() >= 4) %>%
  summarise(
    p_wilcox_total = tryCatch(
      wilcox.test(total ~ Soma_Side)$p.value,
      error = function(e) NA_real_
    ),
    n = n(),
    .groups = "drop"
  )
wt <- wt %>%
  mutate(
    p_wilcox_BH = p.adjust(p_wilcox_total, method = "BH"),
    significant = p_wilcox_BH < 0.05,
    sig_label = p_to_sig(p_wilcox_BH)
  )
print(wt)
```

------------------------------------------------------------------------

## SQ3: Do left and right differ in **multivariate projection pattern** (target-region vectors)?

**Hypothesis**: In the high-dimensional space of region × strength, the L vs R clouds differ in location and/or spread.\
**Why ipsi-only here?** We use `mat_ipsi` to isolate hemisphere-specific targeting and avoid a trivial side-coupling that can appear when ipsi + contra are pooled (because contra definitions are side-referenced). This keeps the distance model focused on within-hemisphere projection organization.\
**Steps**: (1) neuron × region matrix (`mat_ipsi`) → Bray–Curtis dissimilarities between neurons; (2) `adonis2` (PERMANOVA) on `Side`; (3) `betadisper` + `permutest` to test dispersion homogeneity (with euclidification for Bray–Curtis in this step: `sqrt.dist=TRUE`, `add=\"lingoes\"`); (4) PCoA for ordination. **Caveat**: PERMANOVA assumes **exchangeability** of residuals under permutation; heterogeneity of dispersion can inflate false positives, hence the explicit `betadisper` check. $R^2$ is **partial variance explained** by `Side` in the pseudo-F framework, not OLS $R^2$.

```{r sq3_permanova, fig.width=14, fig.height=13}
ipsi_nr <- t(mat_ipsi) # neuron x region
side <- meta$Soma_Side[match(rownames(ipsi_nr), meta$NeuronID)]

# Profile-normalized matrix (composition-like view)
ipsi_tot <- rowSums(ipsi_nr)
ipsi_prop <- ipsi_nr
ok_tot <- ipsi_tot > 0
ipsi_prop[ok_tot, ] <- ipsi_nr[ok_tot, , drop = FALSE] / ipsi_tot[ok_tot]

# Presence/absence matrix for Jaccard
ipsi_pa <- ipsi_nr > 0

# Aitchison via CLR + Euclidean with a small pseudocount
pos_vals <- ipsi_nr[ipsi_nr > 0]
pseudocount <- if (length(pos_vals)) min(pos_vals) / 2 else 1e-6
ipsi_clr <- log(ipsi_nr + pseudocount) - rowMeans(log(ipsi_nr + pseudocount))

dist_list <- list(
  bray_raw = vegdist(ipsi_nr, method = "bray"),
  bray_prop = vegdist(ipsi_prop, method = "bray"),
  jaccard_pa = vegdist(ipsi_pa, method = "jaccard", binary = TRUE),
  aitchison_clr = dist(ipsi_clr)
)

sq3_multi_tab <- lapply(names(dist_list), function(nm) {
  d <- dist_list[[nm]]
  side_d <- side[match(labels(d), rownames(ipsi_nr))]
  ado_i <- adonis2(d ~ side_d, permutations = PERM_N, by = "margin")

  # Euclidify for dispersion check when distance is non-Euclidean
  bd_i <- betadisper(d, side_d, type = "centroid", sqrt.dist = TRUE, add = "lingoes")
  bd_perm_i <- permutest(bd_i, permutations = PERM_N)
  p_bd <- suppressWarnings(as.numeric(bd_perm_i$tab[1, "Pr(>F)"]))

  data.frame(
    distance = nm,
    R2 = as.numeric(ado_i$R2[1]),
    p_permanova = as.numeric(ado_i$`Pr(>F)`[1]),
    p_betadisper = p_bd
  )
}) %>%
  bind_rows() %>%
  mutate(
    significant_permanova = p_permanova < 0.05,
    sig_permanova = p_to_sig(p_permanova),
    significant_betadisper = p_betadisper < 0.05,
    sig_betadisper = p_to_sig(p_betadisper)
  )
print(sq3_multi_tab)

# SQ3-specific sparsity diagnostics on the matrices used in this section.
sq3_sparsity <- bind_rows(
  matrix_sparsity(t(ipsi_nr), "sq3_ipsi_raw_region_by_neuron"),
  matrix_sparsity(t(ipsi_prop), "sq3_ipsi_profile_normalized"),
  matrix_sparsity(t(ipsi_pa), "sq3_ipsi_presence_absence")
) %>%
  mutate(zero_percent = 100 * zero_fraction)
print(sq3_sparsity)

# Keep legacy export variables based on primary metric (bray_raw)
sq3_summary <- sq3_multi_tab %>%
  filter(distance == "bray_raw") %>%
  transmute(
    term = "side",
    R2 = R2,
    p = p_permanova,
    significant = significant_permanova,
    sig_label = sig_permanova
  )
sq3_betadisper <- sq3_multi_tab %>%
  filter(distance == "bray_raw") %>%
  transmute(
    p = p_betadisper,
    significant = significant_betadisper,
    sig_label = sig_betadisper
  )

ord_plot <- function(d, title, subtitle = NULL) {
  md <- cmdscale(d, k = 2, eig = TRUE)
  eig_pos <- pmax(0, md$eig)
  pct <- if (sum(eig_pos) > 0) eig_pos / sum(eig_pos) * 100 else c(NA_real_, NA_real_)
  odf <- data.frame(
    NeuronID = labels(d),
    Axis1 = md$points[, 1],
    Axis2 = md$points[, 2]
  ) %>%
    left_join(meta, by = "NeuronID")
  ggplot(odf, aes(Axis1, Axis2, color = Soma_Side)) +
    geom_point(alpha = 0.8) +
    stat_ellipse(level = 0.95, linewidth = 0.35) +
    scale_color_manual(values = c(L = "#E74C3C", R = "#3498DB")) +
    labs(
      title = title,
      subtitle = subtitle,
      x = sprintf("Axis 1 (%.1f%%)", pct[1]),
      y = sprintf("Axis 2 (%.1f%%)", pct[2])
    ) +
    theme_minimal(base_size = 10)
}

row_br <- sq3_multi_tab %>% filter(distance == "bray_raw")
p1 <- ord_plot(
  dist_list$bray_raw,
  "SQ3: PCoA (Bray raw)",
  sprintf("PERMANOVA p=%.4f, betadisper p=%.4f", row_br$p_permanova[1], row_br$p_betadisper[1])
)
p2 <- ord_plot(dist_list$bray_prop, "SQ3: PCoA (Bray on normalized profile)")
p3 <- ord_plot(dist_list$jaccard_pa, "SQ3: PCoA (Jaccard presence/absence)")
p4 <- ord_plot(dist_list$aitchison_clr, "SQ3: PCoA (Aitchison CLR-Euclidean)")
(p1 | p2) / (p3 | p4)
```

**Extension**: If SQ1 suggests different type composition by side, add Covariates: e.g. `adonis2(... ~ Neuron_Type + Soma_Side)` (order matters for sequential SS interpretation) or **stratify by type** as in SQ6.

------------------------------------------------------------------------

## SQ4: Are **pairwise FNT distances** concordant with **pairwise projection distances**?

**Hypothesis**: Pairs closer in morphology tend to be closer in projection space (direction and strength of association are empirical).\
**Method**: On `common_ids`, **Mantel test** correlates the two distance matrices with a permutation null (proper inference for matrix-valued dependence). A raw **Spearman correlation** on upper-triangle pairs duplicates entries and **does not yield a valid independent-sample p-value**; we still report **rho** descriptively and use **Mantel's permuted p** for significance.

```{r sq4_mantel, eval=has_fnt}
ut <- upper.tri(fnt_sub)
fv <- fnt_sub[ut]
pv <- proj_dist[ut]

# Descriptive association only — pairs are not independent; do not interpret cor.test p-value
sp_rho <- suppressWarnings(cor(fv, pv, method = "spearman", use = "complete.obs"))
cat(sprintf("Spearman rho (upper triangle, descriptive) = %.3f\n", sp_rho))

mt <- mantel(as.dist(fnt_sub), as.dist(proj_dist), method = "spearman", permutations = PERM_N)
print(mt)
sq4_summary <- data.frame(
  statistic = unname(mt$statistic),
  p = unname(mt$signif)
) %>%
  mutate(
    significant = p < 0.05,
    sig_label = p_to_sig(p)
  )
print(sq4_summary)

ggplot(data.frame(FNT = fv, Proj = pv), aes(FNT, Proj)) +
  geom_hex(bins = 60) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  labs(title = "SQ4: Pairwise FNT vs pairwise projection distance (upper triangle)",
       subtitle = sprintf("Spearman rho=%.3f (descriptive); inference: Mantel p", sp_rho)) +
  theme_minimal()
```

------------------------------------------------------------------------

## SQ5: How to compare the two distance matrices side by side in heatmaps?

**Key idea**: Both heatmaps must use the **same row/column order** (here from a fused clustering signal). A third panel shows **\|**$z_{\text{FNT}} - z_{\text{Proj}}$\| per pair after z-scaling upper-triangle moments to highlight discrepancies.

```{r sq5_heatmaps, eval=has_fnt, fig.height=5, fig.width=14}
ut_h <- upper.tri(fnt_sub)
side_lab <- meta_fnt$Soma_Side[match(rownames(fnt_sub), meta_fnt$NeuronID)]
ha <- rowAnnotation(
  Side = side_lab,
  col = list(Side = c(L = "#E74C3C", R = "#3498DB"))
)

z_f <- (fnt_sub - mean(fnt_sub[ut_h])) / sd(fnt_sub[ut_h])
z_p <- (proj_dist - mean(proj_dist[ut_h])) / sd(proj_dist[ut_h])
diff_z <- abs(z_f - z_p)
diag(diff_z) <- NA

combo <- (fnt_sub / max(fnt_sub) + proj_dist / max(proj_dist)) / 2
ord <- hclust(as.dist(combo))$order

rcols <- function(x, utm) {
  colorRamp2(quantile(x[utm], c(0.05, 0.5, 0.95)), c("white", "orange", "red4"))
}

Heatmap(fnt_sub, name = "FNT", col = rcols(fnt_sub, ut_h),
        left_annotation = ha, cluster_rows = FALSE, cluster_columns = FALSE,
        row_order = ord, column_order = ord, show_row_names = FALSE, show_column_names = FALSE) +
Heatmap(proj_dist, name = "Proj", col = rcols(proj_dist, ut_h),
        left_annotation = ha, cluster_rows = FALSE, cluster_columns = FALSE,
        row_order = ord, column_order = ord, show_row_names = FALSE, show_column_names = FALSE) +
Heatmap(diff_z, name = "|Dz|",
        col = colorRamp2(quantile(diff_z, c(0.05, 0.5, 0.95), na.rm = TRUE), c("white", "tan", "brown4")),
        left_annotation = ha, cluster_rows = FALSE, cluster_columns = FALSE,
        row_order = ord, column_order = ord, show_row_names = FALSE, show_column_names = FALSE,
        column_title = "|z_FNT - z_Proj|")
```

------------------------------------------------------------------------

## SQ6: **Does laterality depend on neuron type?** (stratified PERMANOVA)

**Hypothesis**: Only some types show L vs R separation in multivariate projection space.

**Multiple testing**: $p$-values across types are corrected with **Benjamini–Hochberg FDR** in the printed table (`p_BH`).

```{r sq6_stratify}
res_type <- lapply(split(meta, meta$Neuron_Type), function(m) {
  if (nrow(m) < 6L || n_distinct(m$Soma_Side) < 2L) return(NULL)
  ix <- colnames(mat_ipsi) %in% m$NeuronID
  d <- vegdist(t(mat_ipsi[, ix, drop = FALSE]), method = "bray")
  s <- m$Soma_Side[match(labels(d), m$NeuronID)]
  a <- adonis2(d ~ s, permutations = PERM_N)
  data.frame(
    Neuron_Type = unique(m$Neuron_Type),
    n = nrow(m),
    R2 = a$R2[1],
    p = a$`Pr(>F)`[1]
  )
})
sq6_tab <- bind_rows(Filter(Negate(is.null), res_type))
if (nrow(sq6_tab)) {
  sq6_tab <- sq6_tab %>%
    mutate(
      p_BH = p.adjust(p, method = "BH"),
      significant_raw = p < 0.05,
      sig_label_raw = p_to_sig(p),
      significant_BH = p_BH < 0.05,
      sig_label_BH = p_to_sig(p_BH)
    )
}
print(sq6_tab)

if (nrow(sq6_tab)) {
  sq6_plot <- sq6_tab %>%
    mutate(sig_fdr = ifelse(p_BH < 0.05, "BH q < 0.05", "BH q >= 0.05"))
  ggplot(sq6_plot, aes(reorder(Neuron_Type, R2), R2, fill = sig_fdr)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
      values = c("BH q < 0.05" = "#E74C3C", "BH q >= 0.05" = "#bdc3c7"),
      name = NULL
    ) +
    labs(
      title = "SQ6: PERMANOVA R² for Side within each neuron type",
      subtitle = "Bar color: Benjamini–Hochberg FDR across neuron types",
      x = NULL
    ) +
    theme_minimal()
}
```

------------------------------------------------------------------------

# Export results

```{r export}
out <- list(
  MATRIX_sparsity = sparsity_report,
  SQ1_table = as.data.frame.matrix(tab),
  SQ1_prop_within_side = as.data.frame.matrix(prop_side),
  SQ1_chisq = data.frame(
    statistic = unname(chi$statistic),
    p = chi$p.value,
    expected_counts_ok = exp_ok,
    significant = chi$p.value < 0.05,
    sig_label = p_to_sig(chi$p.value)
  ),
  SQ2_wilcox = wt,
  SQ3_PERMANOVA = as.data.frame(ado),
  SQ6_STRATIFIED = sq6_tab
)
if (exists("wt_region_overall", inherits = FALSE) && !is.null(wt_region_overall) && nrow(wt_region_overall)) {
  out$SQ2_wilcox_bilateral_region_overall <- wt_region_overall
}
if (exists("wt_region_type", inherits = FALSE) && !is.null(wt_region_type) && nrow(wt_region_type)) {
  out$SQ2_wilcox_bilateral_region_by_type <- wt_region_type
}
if (!is.null(fish_tab)) {
  out$SQ1_fisher <- data.frame(
    p = unname(fish_tab$p.value),
    significant = unname(fish_tab$p.value) < 0.05,
    sig_label = p_to_sig(unname(fish_tab$p.value))
  )
}
if (!is.null(sq1_region_side_tab)) {
  out$SQ1_region_side_table <- as.data.frame.matrix(sq1_region_side_tab)
}
if (!is.null(sq1_region_side_chi)) {
  out$SQ1_region_side_chisq <- data.frame(
    statistic = unname(sq1_region_side_chi$statistic),
    df = unname(sq1_region_side_chi$parameter),
    p = sq1_region_side_chi$p.value,
    expected_counts_ok = sq1_region_side_exp_ok,
    significant = sq1_region_side_chi$p.value < 0.05,
    sig_label = p_to_sig(sq1_region_side_chi$p.value)
  )
}
if (!is.null(sq1_region_side_fisher)) {
  out$SQ1_region_side_fisher <- data.frame(
    p = unname(sq1_region_side_fisher$p.value),
    significant = unname(sq1_region_side_fisher$p.value) < 0.05,
    sig_label = p_to_sig(unname(sq1_region_side_fisher$p.value))
  )
}
if (!is.null(sq1b_tab)) {
  out$SQ1b_table <- as.data.frame.matrix(sq1b_tab)
  out$SQ1b_prop_within_region <- as.data.frame.matrix(sq1b_prop_region)
}
if (!is.null(sq1b_chi)) {
  out$SQ1b_chisq <- data.frame(
    statistic = unname(sq1b_chi$statistic),
    df = unname(sq1b_chi$parameter),
    p = sq1b_chi$p.value,
    expected_counts_ok = sq1b_exp_ok,
    significant = sq1b_chi$p.value < 0.05,
    sig_label = p_to_sig(sq1b_chi$p.value)
  )
}
if (!is.null(sq1b_fish_tab)) {
  out$SQ1b_fisher <- data.frame(
    p = unname(sq1b_fish_tab$p.value),
    significant = unname(sq1b_fish_tab$p.value) < 0.05,
    sig_label = p_to_sig(unname(sq1b_fish_tab$p.value))
  )
}
if (exists("sq3_summary", inherits = FALSE) && !is.null(sq3_summary) && nrow(sq3_summary)) {
  out$SQ3_summary <- sq3_summary
}
if (exists("sq3_multi_tab", inherits = FALSE) && !is.null(sq3_multi_tab) && nrow(sq3_multi_tab)) {
  out$SQ3_multi_distance <- sq3_multi_tab
}
if (exists("sq3_sparsity", inherits = FALSE) && !is.null(sq3_sparsity) && nrow(sq3_sparsity)) {
  out$SQ3_sparsity <- sq3_sparsity
}
if (exists("sq3_betadisper", inherits = FALSE) && !is.null(sq3_betadisper) && nrow(sq3_betadisper)) {
  out$SQ3_betadisper <- sq3_betadisper
}
if (exists("mt", inherits = FALSE)) {
  out$SQ4_Mantel <- data.frame(
    r = unname(mt$statistic),
    p = unname(mt$signif),
    significant = unname(mt$signif) < 0.05,
    sig_label = p_to_sig(unname(mt$signif))
  )
}
write.xlsx(out, "laterality_by_questions_results.xlsx", overwrite = TRUE)
```
