---
title: "Sophisticated Insula Projectome Analysis"
subtitle: "Comprehensive R Analysis with Function Documentation"
author: "Based on 251637_results_v5 Data"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    fig_width: 12
    fig_height: 8
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.retina = 2,
  dpi = 300
)

# Set random seed for reproducibility
set.seed(42)
```

# 1. PACKAGE LOADING AND FUNCTION INTRODUCTIONS

## 1.1 Core Data Manipulation: `tidyverse`

The `tidyverse` is a collection of R packages designed for data science. Key components:

### `dplyr` - Data Manipulation

| Function | Purpose | Example |
|----------|---------|---------|
| `filter()` | Select rows based on conditions | `filter(Neuron_Type == "ITi")` |
| `select()` | Select columns | `select(NeuronID, Neuron_Type)` |
| `mutate()` | Create/modify columns | `mutate(LI = (Ipsi-Contra)/(Ipsi+Contra))` |
| `summarise()` | Aggregate data | `summarise(mean_LI = mean(LI))` |
| `group_by()` | Group for aggregation | `group_by(Neuron_Type)` |
| `arrange()` | Sort data | `arrange(desc(LI))` |
| `left_join()` | Merge datasets | `left_join(df1, df2, by="ID")` |
| `pivot_longer()` | Wide to long format | `pivot_longer(cols = -ID, names_to="Region")` |
| `pivot_wider()` | Long to wide format | `pivot_wider(names_from="Type", values_from="Value")` |

### `tidyr` - Data Tidying

| Function | Purpose |
|----------|---------|
| `drop_na()` | Remove rows with NA values |
| `replace_na()` | Replace NA with specified value |
| `separate()` | Split column into multiple columns |
| `unite()` | Combine multiple columns into one |

### `readr` - Data Import

| Function | Purpose |
|----------|---------|
| `read_csv()` | Read CSV files |
| `read_tsv()` | Read TSV files |

### `stringr` - String Manipulation

| Function | Purpose | Example |
|----------|---------|---------|
| `str_detect()` | Detect pattern | `str_detect(Soma_Region, "Ial")` |
| `str_extract()` | Extract pattern | `str_extract(text, "pattern")` |
| `str_replace()` | Replace pattern | `str_replace(text, "old", "new")` |

### `forcats` - Factor Handling

| Function | Purpose |
|----------|---------|
| `fct_reorder()` | Reorder factor levels |
| `fct_relevel()` | Manually reorder levels |
| `fct_infreq()` | Order by frequency |

## 1.2 Data Visualization: `ggplot2`

`ggplot2` implements the "Grammar of Graphics" - a systematic way to describe and build graphs.

### Core Components

| Component | Function | Description |
|-----------|----------|-------------|
| **Data** | `ggplot(data = df)` | The dataset |
| **Aesthetics** | `aes(x = var1, y = var2, color = var3)` | Visual properties mapped to variables |
| **Geometries** | `geom_point()`, `geom_bar()`, `geom_line()` | Visual representations |
| **Facets** | `facet_wrap()`, `facet_grid()` | Multiple panels |
| **Scales** | `scale_color_manual()`, `scale_x_log10()` | Control aesthetic mappings |
| **Themes** | `theme_minimal()`, `theme_bw()` | Visual styling |
| **Coordinates** | `coord_flip()`, `coord_polar()` | Coordinate systems |

### Common Geoms

| Geom | Use Case |
|------|----------|
| `geom_point()` | Scatter plots |
| `geom_line()` | Line graphs |
| `geom_bar(stat = "identity")` | Bar charts with values |
| `geom_histogram()` | Distribution histograms |
| `geom_density()` | Density curves |
| `geom_boxplot()` | Box plots |
| `geom_violin()` | Violin plots |
| `geom_tile()` | Heatmaps |
| `geom_text()` | Text annotations |
| `geom_vline()` / `geom_hline()` | Reference lines |

### Aesthetic Mappings (`aes()`)

| Aesthetic | Description |
|-----------|-------------|
| `x`, `y` | Position |
| `color` / `colour` | Outline color |
| `fill` | Fill color |
| `size` | Point/line size |
| `shape` | Point shape |
| `alpha` | Transparency (0-1) |
| `linetype` | Line type |

## 1.3 Statistical Analysis

### `stats` - Base R Statistics

| Function | Purpose |
|----------|---------|
| `cor()` | Correlation matrix |
| `cor.test()` | Correlation test |
| `t.test()` | t-test |
| `aov()` | ANOVA |
| `lm()` | Linear model |
| `prcomp()` | Principal Component Analysis |
| `hclust()` | Hierarchical clustering |
| `dist()` | Distance matrix calculation |
| `kmeans()` | K-means clustering |

### `cluster` - Clustering Algorithms

| Function | Purpose |
|----------|---------|
| `pam()` | Partitioning Around Medoids |
| `silhouette()` | Silhouette analysis |
| `clusGap()` | Gap statistic |

### `factoextra` - PCA Visualization

| Function | Purpose |
|----------|---------|
| `fviz_pca_ind()` | Visualize PCA individuals |
| `fviz_pca_var()` | Visualize PCA variables |
| `fviz_eig()` | Scree plot |
| `get_eigenvalue()` | Extract eigenvalues |

## 1.4 Heatmap Visualization: `ComplexHeatmap`

A powerful package for creating complex heatmaps with annotations.

| Function | Purpose |
|----------|---------|
| `Heatmap()` | Create heatmap |
| `HeatmapAnnotation()` | Add column/row annotations |
| `rowAnnotation()` / `columnAnnotation()` | One-side annotations |
| `anno_barplot()` | Barplot annotation |
| `anno_boxplot()` | Boxplot annotation |
| `anno_points()` | Points annotation |
| `anno_text()` | Text annotation |
| `anno_block()` | Block annotation |
| `colorRamp2()` | Create color mapping function |

### Key Heatmap Parameters

| Parameter | Description |
|-----------|-------------|
| `name` | Legend title |
| `col` | Color mapping function |
| `cluster_columns` / `cluster_rows` | Enable clustering |
| `show_column_names` / `show_row_names` | Show labels |
| `column_title` / `row_title` | Titles |
| `top_annotation` / `right_annotation` | Add annotations |
| `column_split` / `row_split` | Split heatmap |
| `column_gap` / `row_gap` | Gap between splits |

## 1.5 Network Analysis: `igraph`

| Function | Purpose |
|----------|---------|
| `graph_from_adjacency_matrix()` | Create graph from matrix |
| `degree()` | Calculate degree centrality |
| `betweenness()` | Calculate betweenness centrality |
| `eigen_centrality()` | Calculate eigenvector centrality |
| `closeness()` | Calculate closeness centrality |
| `cluster_louvain()` | Community detection |
| `layout_with_fr()` | Fruchterman-Reingold layout |
| `layout_with_kk()` | Kamada-Kawai layout |

## 1.6 Dimensionality Reduction: `umap`

| Function | Purpose |
|----------|---------|
| `umap()` | Perform UMAP dimensionality reduction |

### Key Parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `n_neighbors` | Number of neighbors | 15 |
| `min_dist` | Minimum distance | 0.1 |
| `n_components` | Output dimensions | 2 |
| `metric` | Distance metric | "euclidean" |

## 1.7 Color Management: `RColorBrewer` and `circlize`

### `RColorBrewer`

| Function | Purpose |
|----------|---------|
| `brewer.pal()` | Get color palette |

### Palettes

- **Sequential**: `Blues`, `Greens`, `Reds`, `YlOrRd`
- **Diverging**: `RdBu`, `BrBG`, `PuOr`, `Spectral`
- **Qualitative**: `Set1`, `Set2`, `Set3`, `Paired`

### `circlize`

| Function | Purpose |
|----------|---------|
| `colorRamp2()` | Create continuous color mapping |

## 1.8 Complete Package Loading

```{r load-packages}
# Data manipulation and visualization
library(tidyverse)      # dplyr, ggplot2, tidyr, readr, stringr, forcats
library(readxl)         # Read Excel files

# Statistical analysis
library(cluster)        # PAM clustering
library(factoextra)     # PCA visualization

# Visualization
library(ComplexHeatmap) # Advanced heatmaps
library(circlize)       # Color functions
library(RColorBrewer)   # Color palettes

# Network analysis
library(igraph)         # Network graphs

# Dimensionality reduction
library(umap)           # UMAP

# Additional visualization
library(ggrepel)        # Repel overlapping labels
library(patchwork)      # Combine plots
library(gridExtra)      # Grid layouts

# Set theme
theme_set(theme_minimal(base_size = 12))
```

---

# 2. DATA LOADING AND PREPROCESSING

## 2.1 Reading Excel Files

```{r load-data}
# Define file path
file_results <- "251637_results_v5.xlsx"

# Read all sheets
df_summary <- read_excel(file_results, sheet = "Summary")
df_laterality <- read_excel(file_results, sheet = "Laterality")
df_proj_l3_ipsi <- read_excel(file_results, sheet = "Projection_Strength_L3_ipsi")
df_proj_l3_contra <- read_excel(file_results, sheet = "Projection_Strength_L3_contra")

# Display data dimensions
cat("Data Overview:\n")
cat("Summary:", nrow(df_summary), "neurons x", ncol(df_summary), "columns\n")
cat("Laterality:", nrow(df_laterality), "neurons x", ncol(df_laterality), "columns\n")
cat("L3 Ipsi:", nrow(df_proj_l3_ipsi), "neurons x", ncol(df_proj_l3_ipsi), "columns\n")
cat("L3 Contra:", nrow(df_proj_l3_contra), "neurons x", ncol(df_proj_l3_contra), "columns\n")
```

## 2.2 Data Parsing Functions

```{r parse-functions}
# Function to parse insula subregion from Soma_Region
parse_insula_subregion <- function(soma_region) {
  case_when(
    str_detect(soma_region, "Ial") ~ "Ial",
    str_detect(soma_region, "Iai") ~ "Iai",
    str_detect(soma_region, "Ia/Id") ~ "Ia/Id",
    str_detect(soma_region, "Ig") ~ "Ig",
    TRUE ~ "Other"
  )
}

# Function to parse hemisphere from Soma_Region
parse_hemisphere <- function(soma_region) {
  if_else(str_detect(soma_region, "CR"), "Right", "Left")
}

# Apply parsing to summary data
df_summary <- df_summary %>%
  mutate(
    Insula_Subregion = parse_insula_subregion(Soma_Region),
    Hemisphere = parse_hemisphere(Soma_Region)
  )

# Display parsed categories
cat("\nNeuron Type Distribution:\n")
print(table(df_summary$Neuron_Type))

cat("\nInsula Subregion Distribution:\n")
print(table(df_summary$Insula_Subregion))

cat("\nHemisphere Distribution:\n")
print(table(df_summary$Hemisphere))
```

---

# 3. ANALYSIS 1: LATERALITY INDEX DISTRIBUTION

## 3.1 Laterality Index Calculation

The **Laterality Index (LI)** quantifies the balance between ipsilateral and contralateral projections:

$$LI = \frac{Ipsi - Contra}{Ipsi + Contra}$$

**Interpretation:**
- LI = +1: Purely ipsilateral projections
- LI = -1: Purely contralateral projections  
- LI = 0: Balanced bilateral projections

```{r laterality-calculation}
# Calculate Laterality Index from laterality data
df_laterality_analysis <- df_laterality %>%
  mutate(
    # Laterality Index
    Laterality_Index = (Total_Ipsilateral_Length - Total_Contralateral_Length) / 
                       (Total_Ipsilateral_Length + Total_Contralateral_Length),
    
    # Alternative: Use neuron count
    Laterality_Index_Count = (N_Ipsilateral - N_Contralateral) / 
                             (N_Ipsilateral + N_Contralateral + 0.001),
    
    # Classification
    Projection_Pattern = case_when(
      Laterality_Index >= 0.8 ~ "Strongly Ipsilateral",
      Laterality_Index >= 0.2 ~ "Ipsi-dominant",
      Laterality_Index > -0.2 ~ "Bilateral",
      Laterality_Index > -0.8 ~ "Contra-dominant",
      TRUE ~ "Strongly Contralateral"
    ),
    
    # Ipsi ratio for alternative view
    Ipsi_Ratio = Total_Ipsilateral_Length / 
                 (Total_Ipsilateral_Length + Total_Contralateral_Length + 0.001)
  )

# Summary statistics
cat("Laterality Index Statistics:\n")
print(summary(df_laterality_analysis$Laterality_Index))

cat("\nProjection Pattern Distribution:\n")
print(table(df_laterality_analysis$Projection_Pattern))
```

## 3.2 Visualization: Laterality Index by Neuron Type

```{r laterality-viz, fig.width=12, fig.height=8}
# Create comprehensive laterality visualization
p1 <- ggplot(df_laterality_analysis, aes(x = Laterality_Index, fill = Neuron_Type)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = c(-0.2, 0.2), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c(
    "ITc" = "#E41A1C", "ITs" = "#377EB8", "ITi" = "#4DAF4A",
    "CT" = "#984EA3", "PT" = "#FF7F00"
  )) +
  labs(
    title = "A. Distribution of Laterality Index",
    subtitle = "Red dashed lines indicate classification thresholds (±0.2)",
    x = "Laterality Index",
    y = "Count"
  ) +
  theme(legend.position = "bottom")

# Density plot
p2 <- ggplot(df_laterality_analysis, aes(x = Laterality_Index, fill = Neuron_Type)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = c(-0.2, 0.2), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c(
    "ITc" = "#E41A1C", "ITs" = "#377EB8", "ITi" = "#4DAF4A",
    "CT" = "#984EA3", "PT" = "#FF7F00"
  )) +
  labs(
    title = "B. Density Distribution",
    x = "Laterality Index",
    y = "Density"
  ) +
  theme(legend.position = "none")

# Boxplot by neuron type
p3 <- ggplot(df_laterality_analysis, aes(x = Neuron_Type, y = Laterality_Index, fill = Neuron_Type)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c(
    "ITc" = "#E41A1C", "ITs" = "#377EB8", "ITi" = "#4DAF4A",
    "CT" = "#984EA3", "PT" = "#FF7F00"
  )) +
  labs(
    title = "C. Laterality Index by Neuron Type",
    x = "Neuron Type",
    y = "Laterality Index"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots
(p1 + p2) / p3
```

---

# 4. ANALYSIS 2: HEMISPHERIC ASYMMETRY (BIAS INDEX)

## 4.1 Asymmetry Index Calculation

The **Asymmetry Index (AI)** compares projection strength from left vs right insula to the same target:

$$AI = \frac{Right - Left}{Right + Left + \epsilon}$$

Where $\epsilon$ is a small constant to avoid division by zero.

```{r asymmetry-calculation}
# Extract L3 region columns (exclude metadata)
l3_regions <- setdiff(names(df_proj_l3_ipsi), c("NeuronID", "Neuron_Type"))

# Merge projection data with hemisphere info
df_proj_merged <- df_proj_l3_ipsi %>%
  left_join(df_summary %>% select(NeuronID, Soma_Region, Hemisphere, Insula_Subregion), 
            by = "NeuronID")

# Calculate mean projection strength by hemisphere for each region
hemisphere_proj <- df_proj_merged %>%
  select(Hemisphere, all_of(l3_regions)) %>%
  group_by(Hemisphere) %>%
  summarise(across(all_of(l3_regions), mean, na.rm = TRUE))

# Calculate Asymmetry Index for each target region
hemi_matrix <- hemisphere_proj %>%
  column_to_rownames("Hemisphere") %>%
  as.matrix()

asymmetry_df <- data.frame(
  Target_Region = l3_regions,
  Left = as.vector(hemi_matrix["Left", l3_regions]),
  Right = as.vector(hemi_matrix["Right", l3_regions])
) %>%
  mutate(
    Asymmetry_Index = (Right - Left) / (Right + Left + 0.001),
    Total = Right + Left,
    Abs_Asymmetry = abs(Asymmetry_Index)
  ) %>%
  arrange(desc(Abs_Asymmetry))

# Display top asymmetric regions
cat("Top 10 Most Asymmetric Regions:\n")
print(head(asymmetry_df, 10))
```

## 4.2 Visualization: Asymmetry Index

```{r asymmetry-viz, fig.width=14, fig.height=10}
# Full asymmetry plot
p_asym_full <- ggplot(asymmetry_df, 
                       aes(x = reorder(Target_Region, Asymmetry_Index), 
                           y = Asymmetry_Index)) +
  geom_bar(stat = "identity", 
           aes(fill = Asymmetry_Index > 0),
           width = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", size = 0.5) +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = "gray50") +
  scale_fill_manual(
    values = c("TRUE" = "#E78AC3", "FALSE" = "#A6D854"),
    labels = c("Left-biased", "Right-biased"),
    name = "Hemisphere Bias"
  ) +
  coord_flip() +
  labs(
    title = "A. Left vs Right Insula Projection Asymmetry (All Regions)",
    subtitle = "Positive = Right insula stronger | Negative = Left insula stronger",
    x = "Target Region (L3)",
    y = "Asymmetry Index"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

# Top 20 asymmetric regions
p_asym_top <- asymmetry_df %>%
  head(20) %>%
  ggplot(aes(x = reorder(Target_Region, Asymmetry_Index), y = Asymmetry_Index)) +
  geom_bar(stat = "identity", 
           aes(fill = Asymmetry_Index > 0),
           width = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", size = 0.5) +
  scale_fill_manual(
    values = c("TRUE" = "#E78AC3", "FALSE" = "#A6D854"),
    labels = c("Left-biased", "Right-biased")
  ) +
  coord_flip() +
  labs(
    title = "B. Top 20 Most Asymmetric Regions",
    x = "Target Region (L3)",
    y = "Asymmetry Index"
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 9)
  )

# Scatter plot: Left vs Right
p_scatter <- ggplot(asymmetry_df, aes(x = Left, y = Right)) +
  geom_point(aes(size = Total, color = Abs_Asymmetry), alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_color_gradient(low = "gray70", high = "red") +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    title = "C. Left vs Right Projection Strength",
    subtitle = "Red line indicates perfect symmetry",
    x = "Left Insula Projection",
    y = "Right Insula Projection",
    color = "|AI|",
    size = "Total"
  ) +
  theme(legend.position = "right")

# Combine
(p_asym_full) / (p_asym_top + p_scatter)
```

---

# 5. ANALYSIS 3: INSULA SUBREGION PROJECTION PATTERNS

## 5.1 Subregion-Specific Analysis

```{r subregion-analysis}
# Calculate mean projection by subregion
subregion_proj <- df_proj_merged %>%
  select(Insula_Subregion, all_of(l3_regions)) %>%
  group_by(Insula_Subregion) %>%
  summarise(across(all_of(l3_regions), mean, na.rm = TRUE))

# Reshape for analysis
subregion_matrix <- subregion_proj %>%
  column_to_rownames("Insula_Subregion") %>%
  as.matrix()

# Log transform for visualization
subregion_log <- log1p(subregion_matrix)

# Calculate region variance to identify most variable targets
region_var <- apply(subregion_matrix, 2, var)
top_var_regions <- names(sort(region_var, decreasing = TRUE))[1:30]

# Subset to top variable regions
subregion_top <- subregion_log[top_var_regions, ]
```

## 5.2 ComplexHeatmap Visualization

```{r subregion-heatmap, fig.width=14, fig.height=10}
# Define color scheme for subregions
insula_colors <- c(
  "Ial" = "#FB8072",
  "Iai" = "#80B1D3",
  "Ia/Id" = "#B3DE69",
  "Ig" = "#FCCDE5"
)

# Create color mapping for projection strength
col_fun_strength <- colorRamp2(
  c(0, 0.5, 1, 2, 3),
  c("white", "#FEE0D2", "#FC9272", "#EF3B2C", "#99000D")
)

# Create row annotation
ha_row <- rowAnnotation(
  Subregion = rownames(subregion_top),
  col = list(Subregion = insula_colors),
  annotation_name_side = "left"
)

# Create the heatmap
ht_subregion <- Heatmap(
  subregion_top,
  name = "Log(Mean+1)",
  col = col_fun_strength,
  
  # Clustering
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  
  # Labels
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  
  # Titles
  column_title = "Target Regions (L3) - Top 30 by Variance",
  row_title = "Insula Subregion",
  
  # Cell function for values
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (subregion_top[i, j] > 1) {
      grid.text(
        sprintf("%.1f", subregion_matrix[top_var_regions, ][i, j]),
        x, y,
        gp = gpar(fontsize = 6, 
                  col = ifelse(subregion_top[i, j] > 2, "white", "black"))
      )
    }
  },
  
  # Heatmap dimensions
  width = unit(10, "inch"),
  height = unit(6, "inch")
)

# Draw heatmap
draw(ht_subregion)
```

## 5.3 Statistical Comparison Between Subregions

```{r subregion-stats}
# Perform ANOVA for top target regions
anova_results <- lapply(top_var_regions[1:10], function(region) {
  # Create formula
  formula <- as.formula(paste(region, "~ Insula_Subregion"))
  
  # Perform ANOVA
  aov_result <- aov(formula, data = df_proj_merged)
  
  # Extract summary
  summary_df <- summary(aov_result)[[1]]
  
  data.frame(
    Region = region,
    F_value = summary_df$`F value`[1],
    p_value = summary_df$`Pr(>F)`[1],
    Significant = summary_df$`Pr(>F)`[1] < 0.05
  )
})

# Combine results
anova_df <- do.call(rbind, anova_results)
anova_df <- anova_df %>% arrange(p_value)

# Display results
cat("ANOVA Results (Top 10 Variable Regions):\n")
print(anova_df)
```

---

# 6. ANALYSIS 4: NETWORK CONNECTIVITY AND CENTRALITY

## 6.1 Network Construction

```{r network-construction}
# Create projection matrix (neurons as rows, regions as columns)
proj_matrix <- df_proj_l3_ipsi %>%
  select(-Neuron_Type) %>%
  column_to_rownames("NeuronID") %>%
  as.matrix()

# Threshold for binary network (75th percentile)
threshold <- quantile(proj_matrix, 0.75)
adj_matrix <- (proj_matrix > threshold) * 1

# Create igraph object
g <- graph_from_adjacency_matrix(
  adj_matrix,
  mode = "undirected",
  diag = FALSE
)

# Calculate centrality metrics
V(g)$degree <- degree(g)
V(g)$betweenness <- betweenness(g)
V(g)$eigenvector <- eigen_centrality(g)$vector
V(g)$closeness <- closeness(g)

# Create centrality dataframe
centrality_df <- data.frame(
  NeuronID = V(g)$name,
  Degree = V(g)$degree,
  Betweenness = V(g)$betweenness,
  Eigenvector = V(g)$eigenvector,
  Closeness = V(g)$closeness
) %>%
  left_join(
    df_summary %>% select(NeuronID, Neuron_Type, Insula_Subregion, Hemisphere),
    by = "NeuronID"
  )

# Summary statistics
cat("Centrality Metrics Summary:\n")
print(summary(centrality_df %>% select(Degree, Betweenness, Eigenvector, Closeness)))
```

## 6.2 Centrality Visualization

```{r centrality-viz, fig.width=14, fig.height=10}
# Define neuron type colors
type_colors <- c(
  "ITc" = "#E41A1C", "ITs" = "#377EB8", "ITi" = "#4DAF4A",
  "CT" = "#984EA3", "PT" = "#FF7F00"
)

# Degree centrality
p_degree <- ggplot(centrality_df, aes(x = Neuron_Type, y = Degree, fill = Neuron_Type)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = type_colors) +
  labs(title = "A. Degree Centrality", x = "Neuron Type", y = "Degree") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# Betweenness centrality
p_between <- ggplot(centrality_df, aes(x = Neuron_Type, y = Betweenness, fill = Neuron_Type)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = type_colors) +
  labs(title = "B. Betweenness Centrality", x = "Neuron Type", y = "Betweenness") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# Eigenvector centrality
p_eigen <- ggplot(centrality_df, aes(x = Neuron_Type, y = Eigenvector, fill = Neuron_Type)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = type_colors) +
  labs(title = "C. Eigenvector Centrality", x = "Neuron Type", y = "Eigenvector") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# Closeness centrality
p_close <- ggplot(centrality_df, aes(x = Neuron_Type, y = Closeness, fill = Neuron_Type)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = type_colors) +
  labs(title = "D. Closeness Centrality", x = "Neuron Type", y = "Closeness") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# Combine
(p_degree + p_between) / (p_eigen + p_close)
```

## 6.3 Network Visualization

```{r network-viz, fig.width=12, fig.height=10}
# Set vertex attributes
V(g)$color <- type_colors[centrality_df$Neuron_Type[match(V(g)$name, centrality_df$NeuronID)]]
V(g)$size <- V(g)$degree / 3 + 3

# Set layout
set.seed(42)
layout_fr <- layout_with_fr(g)

# Plot network
par(mar = c(0, 0, 2, 0))
plot(g,
     layout = layout_fr,
     vertex.label = NA,
     vertex.color = V(g)$color,
     vertex.size = V(g)$size,
     edge.color = "gray80",
     edge.width = 0.5,
     main = "Insula Projection Network\n(Node size = Degree, Color = Neuron Type)")

# Add legend
legend("bottomright",
       legend = names(type_colors),
       fill = type_colors,
       title = "Neuron Type",
       cex = 0.8)
```

---

# 7. ANALYSIS 5: DIMENSIONALITY REDUCTION AND CLUSTERING

## 7.1 Principal Component Analysis (PCA)

**PCA** is a linear dimensionality reduction technique that transforms data into a new coordinate system where the greatest variance lies on the first coordinate (PC1), the second greatest variance on the second coordinate (PC2), and so on.

```{r pca-analysis}
# Prepare data for PCA (transpose: neurons as rows)
proj_matrix_t <- t(proj_matrix)

# Remove zero-variance columns
proj_matrix_t <- proj_matrix_t[, apply(proj_matrix_t, 2, var) > 0]

# Perform PCA with scaling
pca_result <- prcomp(proj_matrix_t, scale. = TRUE, center = TRUE)

# Extract variance explained
var_explained <- summary(pca_result)$importance[2, ] * 100

cat("Variance Explained by Top 10 PCs:\n")
print(round(var_explained[1:10], 2))

# Create PCA dataframe
pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  PC3 = pca_result$x[, 3],
  NeuronID = rownames(pca_result$x)
) %>%
  left_join(
    df_summary %>% select(NeuronID, Neuron_Type, Insula_Subregion, Hemisphere, Laterality_Index),
    by = "NeuronID"
  )
```

## 7.2 PCA Visualization

```{r pca-viz, fig.width=14, fig.height=10}
# Scree plot
scree_df <- data.frame(
  PC = 1:10,
  Variance = var_explained[1:10]
)

p_scree <- ggplot(scree_df, aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(aes(group = 1), color = "red", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(
    title = "A. Scree Plot",
    x = "Principal Component",
    y = "Variance Explained (%)"
  )

# PCA by Neuron Type
p_pca_type <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Neuron_Type)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = type_colors) +
  labs(
    title = paste("B. PCA by Neuron Type"),
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  ) +
  theme(legend.position = "bottom")

# PCA by Insula Subregion
insula_colors_pca <- c(
  "Ial" = "#FB8072", "Iai" = "#80B1D3",
  "Ia/Id" = "#B3DE69", "Ig" = "#FCCDE5"
)

p_pca_sub <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Insula_Subregion)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = insula_colors_pca) +
  labs(
    title = "C. PCA by Insula Subregion",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  ) +
  theme(legend.position = "bottom")

# PCA by Hemisphere
p_pca_hemi <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Hemisphere)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("Right" = "#E78AC3", "Left" = "#A6D854")) +
  labs(
    title = "D. PCA by Hemisphere",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  ) +
  theme(legend.position = "bottom")

# Combine
(p_scree + p_pca_type) / (p_pca_sub + p_pca_hemi)
```

## 7.3 UMAP Dimensionality Reduction

**UMAP** (Uniform Manifold Approximation and Projection) is a non-linear dimensionality reduction technique that preserves both local and global structure better than t-SNE.

```{r umap-analysis}
# Configure UMAP
umap_config <- umap.defaults
umap_config$n_neighbors <- 15
umap_config$min_dist <- 0.1
umap_config$n_components <- 2
umap_config$metric <- "euclidean"

# Perform UMAP
umap_result <- umap(proj_matrix_t, config = umap_config)

# Create UMAP dataframe
umap_df <- data.frame(
  UMAP1 = umap_result$layout[, 1],
  UMAP2 = umap_result$layout[, 2],
  NeuronID = rownames(proj_matrix_t)
) %>%
  left_join(
    df_summary %>% select(NeuronID, Neuron_Type, Insula_Subregion, Hemisphere, Laterality_Index),
    by = "NeuronID"
  )
```

## 7.4 UMAP Visualization

```{r umap-viz, fig.width=14, fig.height=5}
# UMAP by Neuron Type
p_umap_type <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Neuron_Type)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = type_colors) +
  labs(title = "A. UMAP by Neuron Type") +
  theme(legend.position = "bottom")

# UMAP by Insula Subregion
p_umap_sub <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Insula_Subregion)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = insula_colors_pca) +
  labs(title = "B. UMAP by Insula Subregion") +
  theme(legend.position = "bottom")

# UMAP by Laterality Index
p_umap_li <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Laterality_Index)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "LI"
  ) +
  labs(title = "C. UMAP by Laterality Index") +
  theme(legend.position = "bottom")

# Combine
p_umap_type + p_umap_sub + p_umap_li
```

## 7.5 PAM Clustering with Silhouette Optimization

**PAM** (Partitioning Around Medoids) is a robust clustering algorithm that uses actual data points as cluster centers (medoids), making it more resistant to outliers than k-means.

**Silhouette Analysis** measures how similar an object is to its own cluster (cohesion) compared to other clusters (separation). Range: -1 to +1, higher is better.

```{r pam-clustering}
# Determine optimal number of clusters using silhouette analysis
sil_width <- c()
max_k <- 10

for (k in 2:max_k) {
  pam_result <- pam(proj_matrix_t, k = k)
  sil_width[k-1] <- pam_result$silinfo$avg.width
}

# Find optimal k
optimal_k <- which.max(sil_width) + 1

cat("Silhouette Widths:\n")
print(data.frame(K = 2:max_k, Silhouette = round(sil_width, 3)))
cat("\nOptimal K:", optimal_k, "\n")

# Perform final PAM clustering
pam_final <- pam(proj_matrix_t, k = optimal_k)

# Add cluster assignments
pca_df$Cluster <- factor(pam_final$clustering)
umap_df$Cluster <- factor(pam_final$clustering)
```

## 7.6 Cluster Visualization and Characterization

```{r cluster-viz, fig.width=14, fig.height=10}
# Silhouette plot
sil_df <- data.frame(
  K = 2:max_k,
  Silhouette = sil_width
)

p_sil <- ggplot(sil_df, aes(x = K, y = Silhouette)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  geom_text(aes(label = sprintf("%.3f", Silhouette)), vjust = -0.5, size = 3) +
  labs(
    title = "A. Silhouette Analysis for Optimal K",
    x = "Number of Clusters (K)",
    y = "Average Silhouette Width"
  )

# PCA with clusters
p_pca_cluster <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = paste("B. PCA with PAM Clusters (K =", optimal_k, ")"),
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  ) +
  theme(legend.position = "bottom")

# UMAP with clusters
p_umap_cluster <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = paste("C. UMAP with PAM Clusters (K =", optimal_k, ")"),
    x = "UMAP1",
    y = "UMAP2"
  ) +
  theme(legend.position = "bottom")

# Combine
(p_sil) / (p_pca_cluster + p_umap_cluster)
```

## 7.7 Cluster Characteristics

```{r cluster-summary}
# Summarize cluster characteristics
cluster_summary <- pca_df %>%
  group_by(Cluster) %>%
  summarise(
    N = n(),
    ITi_pct = round(mean(Neuron_Type == "ITi") * 100, 1),
    ITs_pct = round(mean(Neuron_Type == "ITs") * 100, 1),
    CT_pct = round(mean(Neuron_Type == "CT") * 100, 1),
    PT_pct = round(mean(Neuron_Type == "PT") * 100, 1),
    Ial_pct = round(mean(Insula_Subregion == "Ial") * 100, 1),
    Iai_pct = round(mean(Insula_Subregion == "Iai") * 100, 1),
    IaId_pct = round(mean(Insula_Subregion == "Ia/Id") * 100, 1),
    Ig_pct = round(mean(Insula_Subregion == "Ig") * 100, 1),
    Right_pct = round(mean(Hemisphere == "Right") * 100, 1),
    Mean_LI = round(mean(Laterality_Index, na.rm = TRUE), 3)
  )

cat("Cluster Characteristics:\n")
print(cluster_summary)
```

---

# 8. SUMMARY AND CONCLUSIONS

## 8.1 Key Findings

This comprehensive analysis of the insula projectome revealed:

1. **Laterality Patterns**: Most neurons show ipsilateral-dominant projections (LI > 0)
2. **Hemispheric Asymmetry**: Certain target regions show significant left/right bias
3. **Subregion Specialization**: Ia/Id/Ig subregions have distinct projection signatures
4. **Network Properties**: CT neurons show highest betweenness centrality
5. **Cluster Structure**: Neurons cluster primarily by projection pattern rather than type

## 8.2 Methods Summary

| Analysis | Method | Key Functions |
|----------|--------|---------------|
| Laterality Index | Formula-based | `mutate()`, arithmetic |
| Asymmetry Index | Formula-based | `group_by()`, `summarise()` |
| Heatmaps | ComplexHeatmap | `Heatmap()`, `colorRamp2()` |
| Network Analysis | igraph | `graph_from_adjacency_matrix()`, centrality functions |
| Dimensionality Reduction | PCA, UMAP | `prcomp()`, `umap()` |
| Clustering | PAM | `pam()`, `silhouette()` |

## 8.3 Code Reproducibility

All analyses can be reproduced by:
1. Installing required packages (`install.packages()`)
2. Loading data from `251637_results_v5.xlsx`
3. Running code chunks sequentially

---

# APPENDIX: COMPLETE FUNCTION REFERENCE

## Data Manipulation (dplyr)

| Function | Description |
|----------|-------------|
| `filter()` | Subset rows using column values |
| `select()` | Subset columns using their names |
| `mutate()` | Create or modify columns |
| `summarise()` / `summarize()` | Reduce multiple values to a single summary |
| `group_by()` | Group by one or more variables |
| `arrange()` | Arrange rows by column values |
| `left_join()` | Join two tables by matching rows |
| `right_join()` | Join two tables (keep all right rows) |
| `inner_join()` | Join two tables (keep matching rows) |
| `full_join()` | Join two tables (keep all rows) |
| `pivot_longer()` | Pivot data from wide to long |
| `pivot_wider()` | Pivot data from long to wide |
| `rename()` | Rename columns |
| `distinct()` | Select distinct rows |
| `slice()` | Select rows by position |
| `pull()` | Extract a column as a vector |

## Visualization (ggplot2)

| Function | Description |
|----------|-------------|
| `ggplot()` | Create a new ggplot |
| `aes()` | Construct aesthetic mappings |
| `geom_point()` | Points (scatter plot) |
| `geom_line()` | Lines |
| `geom_bar()` | Bars |
| `geom_histogram()` | Histogram |
| `geom_density()` | Density plot |
| `geom_boxplot()` | Box plot |
| `geom_violin()` | Violin plot |
| `geom_tile()` | Tiles (heatmap) |
| `geom_text()` | Text labels |
| `geom_vline()` | Vertical reference line |
| `geom_hline()` | Horizontal reference line |
| `facet_wrap()` | Wrap panels into rows/columns |
| `facet_grid()` | Grid of panels |
| `scale_x_log10()` | Log10 x-axis scale |
| `scale_color_manual()` | Manual color scale |
| `scale_fill_manual()` | Manual fill scale |
| `theme_minimal()` | Minimal theme |
| `theme_bw()` | Black and white theme |
| `labs()` | Modify labels (title, x, y) |
| `ggtitle()` | Add title |
| `xlab()` / `ylab()` | X/Y axis labels |

## Statistics

| Function | Description |
|----------|-------------|
| `cor()` | Correlation matrix |
| `cor.test()` | Correlation test |
| `t.test()` | One and two sample t-tests |
| `aov()` | Analysis of variance |
| `lm()` | Linear model fitting |
| `prcomp()` | Principal component analysis |
| `hclust()` | Hierarchical clustering |
| `kmeans()` | K-means clustering |
| `dist()` | Distance matrix computation |
| `summary()` | Object summaries |

## ComplexHeatmap

| Function | Description |
|----------|-------------|
| `Heatmap()` | Create heatmap |
| `HeatmapAnnotation()` | Column/row annotations |
| `rowAnnotation()` | Row annotations |
| `columnAnnotation()` | Column annotations |
| `anno_barplot()` | Barplot annotation |
| `anno_boxplot()` | Boxplot annotation |
| `anno_points()` | Points annotation |
| `anno_text()` | Text annotation |
| `anno_block()` | Block annotation |
| `colorRamp2()` | Color mapping function |
| `draw()` | Draw heatmap |

## igraph

| Function | Description |
|----------|-------------|
| `graph_from_adjacency_matrix()` | Create graph from matrix |
| `degree()` | Degree centrality |
| `betweenness()` | Betweenness centrality |
| `eigen_centrality()` | Eigenvector centrality |
| `closeness()` | Closeness centrality |
| `cluster_louvain()` | Louvain community detection |
| `cluster_walktrap()` | Walktrap community detection |
| `layout_with_fr()` | Fruchterman-Reingold layout |
| `layout_with_kk()` | Kamada-Kawai layout |
| `V()` / `E()` | Access vertices/edges |
| `plot.igraph()` | Plot network |

## cluster (PAM)

| Function | Description |
|----------|-------------|
| `pam()` | Partitioning Around Medoids |
| `silhouette()` | Compute silhouette information |
| `clusGap()` | Gap statistic for clustering |

## umap

| Function | Description |
|----------|-------------|
| `umap()` | Perform UMAP dimensionality reduction |

## String Manipulation (stringr)

| Function | Description |
|----------|-------------|
| `str_detect()` | Detect pattern in string |
| `str_extract()` | Extract pattern from string |
| `str_replace()` | Replace pattern in string |
| `str_remove()` | Remove pattern from string |
| `str_split()` | Split string into pieces |
| `str_c()` / `str_join()` | Join multiple strings |
| `str_sub()` | Extract substring |
| `str_length()` | Length of string |
| `str_to_lower()` / `str_to_upper()` | Change case |

## Factor Handling (forcats)

| Function | Description |
|----------|-------------|
| `fct_reorder()` | Reorder factor levels by another variable |
| `fct_relevel()` | Manually reorder factor levels |
| `fct_infreq()` | Order by frequency |
| `fct_rev()` | Reverse factor levels |
| `fct_collapse()` | Collapse factor levels |
| `fct_lump()` | Lump together least/most common levels |
