---
title: "Single-Neuron Projection Analysis - 251637 Dataset"
subtitle: "Based on Gao et al., 2023 Neuron methodology"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    fig_width: 10
    fig_height: 8
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 8
)

# Load required packages
packages <- c("readxl", "dplyr", "tidyr", "ggplot2", "ggpubr", "ggExtra",
              "ComplexHeatmap", "circlize", "RColorBrewer", "viridis",
              "grid", "igraph", "pheatmap")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org/")
    library(pkg, character.only = TRUE)
  }
}

# ComplexHeatmap from Bioconductor
if (!require("ComplexHeatmap", quietly = TRUE)) {
  if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("ComplexHeatmap")
  library(ComplexHeatmap)
}

# Set color schemes
type_colors <- c(
  "ITc" = "#4DAF4A",  # Green
  "ITs" = "#984EA3",  # Purple
  "CT"  = "#FF7F00",  # Orange
  "PT"  = "#E41A1C",  # Red
  "ITi" = "#377EB8"   # Blue
)

col_fun_strength <- colorRamp2(
  c(0, 2, 4, 6, 8),
  c("white", "lightblue", "blue", "darkblue", "navy")
)
```

# 1. Data Loading and Overview

```{r load-data}
# File path - modify if needed
file_results <- "251637_results_v5.xlsx"

# Check file exists
if (!file.exists(file_results)) {
  stop("File not found: ", file_results, 
       "\nPlease ensure the file is in the working directory.")
}

# Load all sheets
df_summary <- read_excel(file_results, sheet = "Summary")
df_hierarchy <- read_excel(file_results, sheet = "Soma_Hierarchy")
df_terminals <- read_excel(file_results, sheet = "Terminal_Sites")
df_laterality <- read_excel(file_results, sheet = "Laterality")

# Load L3 projection data (hierarchy level 3)
df_proj_l3_ipsi <- read_excel(file_results, sheet = "Projection_Strength_L3_ipsi")
df_proj_l3_contra <- read_excel(file_results, sheet = "Projection_Strength_L3_contra")
df_proj_length_l3_ipsi <- read_excel(file_results, sheet = "Projection_Length_L3_ipsi")

# Print data overview
cat("Data Overview:\n")
cat("================\n")
cat("Summary:", nrow(df_summary), "neurons\n")
cat("L3 ipsilateral regions:", ncol(df_proj_l3_ipsi) - 2, "\n")
cat("L3 contralateral regions:", ncol(df_proj_l3_contra) - 2, "\n")
cat("Terminal sites:", nrow(df_terminals), "\n")
```

## 1.1 Neuron Type Distribution

```{r neuron-type-dist}
# Create type distribution plot
ggplot(df_summary, aes(x = Neuron_Type, fill = Neuron_Type)) +
  geom_bar() +
  scale_fill_manual(values = type_colors) +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5, size = 5) +
  labs(
    title = "Neuron Type Distribution",
    subtitle = paste("n =", nrow(df_summary), "neurons from CR_Ial"),
    x = "Neuron Type",
    y = "Count"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")
```

# 2. Morphological Characteristics

## 2.1 Total Axon Length by Type

```{r axon-length}
ggplot(df_summary, aes(x = Neuron_Type, y = Total_Length, fill = Neuron_Type)) +
  geom_boxplot(alpha = 0.8) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = type_colors) +
  labs(
    title = "Total Axon Length by Neuron Type",
    x = "Neuron Type",
    y = "Total Length (μm)"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none") +
  stat_compare_means(method = "anova", label.y = max(df_summary$Total_Length) * 1.1)
```

## 2.2 Axon Length vs Terminal Count

```{r length-vs-terminal}
p_scatter <- ggplot(df_summary, aes(x = Total_Length, y = Terminal_Count, color = Neuron_Type)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = type_colors) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  labs(
    title = "Axon Length vs Terminal Count",
    x = "Total Axon Length (μm)",
    y = "Terminal Count"
  ) +
  theme_classic(base_size = 14)

# Add marginal distributions
ggExtra::ggMarginal(p_scatter, groupColour = TRUE, groupFill = TRUE, margins = "both")
```

## 2.3 Statistical Summary by Type

```{r type-stats}
# Calculate statistics
type_stats <- df_summary %>%
  group_by(Neuron_Type) %>%
  summarise(
    n = n(),
    mean_length = mean(Total_Length),
    sd_length = sd(Total_Length),
    median_length = median(Total_Length),
    mean_terminals = mean(Terminal_Count),
    sd_terminals = sd(Terminal_Count),
    mean_laterality = mean(Laterality_Index),
    .groups = "drop"
  )

knitr::kable(type_stats, digits = 2, 
             caption = "Morphological Characteristics by Neuron Type")
```

# 3. Projection Heatmap Analysis

## 3.1 Prepare Projection Matrix

```{r prepare-matrix}
# Extract projection matrix for ipsilateral projections
proj_matrix_l3 <- df_proj_l3_ipsi %>%
  select(-NeuronID, -Neuron_Type) %>%
  as.matrix()
rownames(proj_matrix_l3) <- df_proj_l3_ipsi$NeuronID

# Transpose: rows = brain regions, columns = neurons
proj_matrix_t <- t(proj_matrix_l3)

# Clean data (remove NA/Inf)
good_rows <- rowSums(!is.finite(proj_matrix_t)) == 0
mat_clean <- proj_matrix_t[good_rows, ]

# Get neuron type information
neuron_types <- df_proj_l3_ipsi$Neuron_Type[match(colnames(mat_clean), df_proj_l3_ipsi$NeuronID)]

# Get summary info
neuron_info <- df_summary %>%
  filter(NeuronID %in% colnames(mat_clean)) %>%
  arrange(match(NeuronID, colnames(mat_clean)))

cat("Projection matrix:", nrow(mat_clean), "regions x", ncol(mat_clean), "neurons\n")
```

## 3.2 Basic Projection Heatmap

```{r basic-heatmap, fig.width=12, fig.height=14}
# Create type annotation
ha_top_type <- HeatmapAnnotation(
  Type = neuron_types,
  col = list(Type = type_colors),
  annotation_name_side = "left"
)

# Basic heatmap
ht_basic <- Heatmap(
  log1p(mat_clean),
  name = "Log Strength",
  col = col_fun_strength,
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  top_annotation = ha_top_type,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 8),
  column_title = "Neurons (grouped by type)",
  row_title = "Brain Regions (L3)",
  width = unit(10, "cm"),
  height = unit(12, "cm")
)

draw(ht_basic)
```

## 3.3 Heatmap with Rich Annotations

```{r rich-heatmap, fig.width=14, fig.height=16}
# Create rich annotation
ha_top_rich <- HeatmapAnnotation(
  Type = neuron_types,
  TotalLength = anno_barplot(
    neuron_info$Total_Length,
    gp = gpar(fill = "steelblue"),
    height = unit(0.8, "cm")
  ),
  TerminalCount = anno_barplot(
    neuron_info$Terminal_Count,
    gp = gpar(fill = "orange"),
    height = unit(0.8, "cm")
  ),
  col = list(Type = type_colors),
  annotation_name_side = "left"
)

# Create type factor for splitting
type_factor <- factor(neuron_types, levels = c("ITi", "ITs", "CT", "PT", "ITc"))

# Rich heatmap
ht_rich <- Heatmap(
  log1p(mat_clean),
  name = "Log Strength",
  col = col_fun_strength,
  cluster_columns = TRUE,
  column_split = type_factor,
  column_gap = unit(1, "mm"),
  top_annotation = ha_top_rich,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 7),
  column_title = "Neurons (k=212, grouped by type)",
  row_title = "Brain Regions (L3 hierarchy)",
  width = unit(12, "cm"),
  height = unit(14, "cm"),
  border = TRUE
)

draw(ht_rich)
```

# 4. Functional Analysis - Projection Patterns

## 4.1 Type-to-Region Projection Profile

```{r type-region-profile}
# Calculate mean projection for each type to each region
type_region_profile <- df_proj_l3_ipsi %>%
  group_by(Neuron_Type) %>%
  summarise(across(-NeuronID, mean, na.rm = TRUE), .groups = "drop") %>%
  column_to_rownames("Neuron_Type")

# Normalize by row (each type)
type_region_norm <- type_region_profile / rowMaxs(as.matrix(type_region_profile))

# Plot heatmap
ht_type_region <- Heatmap(
  as.matrix(type_region_norm),
  name = "Normalized\nStrength",
  col = colorRamp2(c(0, 0.5, 1), c("white", "yellow", "red")),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_names_side = "left",
  column_names_side = "top",
  row_names_gp = gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = gpar(fontsize = 6),
  column_names_rot = 45,
  width = unit(14, "cm"),
  height = unit(5, "cm"),
  border = TRUE
)

draw(ht_type_region)
```

## 4.2 Projection Strength Distribution

```{r strength-distribution}
# Convert to long format
df_proj_long <- df_proj_l3_ipsi %>%
  pivot_longer(cols = -(1:2), names_to = "Region", values_to = "Strength")

# Plot violin
ggplot(df_proj_long, aes(x = Neuron_Type, y = Strength, fill = Neuron_Type)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.8) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  scale_fill_manual(values = type_colors) +
  labs(
    title = "Projection Strength Distribution by Neuron Type",
    x = "Neuron Type",
    y = "Projection Strength (log)"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none") +
  stat_compare_means()
```

## 4.3 Number of Projection Targets

```{r n-targets}
# Calculate number of targets per neuron
df_n_targets <- df_proj_l3_ipsi %>%
  mutate(n_targets = rowSums(select(., -(1:2)) > 0)) %>%
  select(NeuronID, Neuron_Type, n_targets)

ggplot(df_n_targets, aes(x = Neuron_Type, y = n_targets, fill = Neuron_Type)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.8) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  scale_fill_manual(values = type_colors) +
  labs(
    title = "Number of Projection Targets",
    x = "Neuron Type",
    y = "Number of Target Regions"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none") +
  stat_compare_means()
```

# 5. Laterality Analysis

## 5.1 Laterality Index Distribution

```{r laterality-index}
ggplot(df_summary, aes(x = Neuron_Type, y = Laterality_Index, fill = Neuron_Type)) +
  geom_boxplot(alpha = 0.8) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = type_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Laterality Index by Neuron Type",
    subtitle = "1 = Purely Ipsilateral, -1 = Purely Contralateral",
    x = "Neuron Type",
    y = "Laterality Index"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")
```

## 5.2 Ipsilateral vs Contralateral Comparison

```{r ipsi-contra}
# Prepare data
df_lateral <- df_summary %>%
  select(NeuronID, Neuron_Type, 
         Ipsilateral = Total_Ipsilateral_Length,
         Contralateral = Total_Contralateral_Length) %>%
  pivot_longer(cols = c(Ipsilateral, Contralateral),
               names_to = "Hemisphere", values_to = "Length")

ggplot(df_lateral, aes(x = Hemisphere, y = Length, fill = Hemisphere)) +
  geom_boxplot(alpha = 0.8) +
  facet_wrap(~Neuron_Type, scales = "free_y") +
  scale_fill_manual(values = c("Ipsilateral" = "#FDB462", "Contralateral" = "#80B1D3")) +
  labs(
    title = "Ipsilateral vs Contralateral Projection Length",
    y = "Total Length (μm)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

# 6. Terminal Site Analysis

## 6.1 Terminal Distribution by Brain Lobe

```{r terminal-l1}
# Count terminals by L1 (if available)
terminal_l1 <- df_terminals %>%
  filter(!is.na(Terminal_L1)) %>%
  group_by(Terminal_L1) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n))

if (nrow(terminal_l1) > 0) {
  ggplot(terminal_l1, aes(x = reorder(Terminal_L1, -n), y = n)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = "Terminal Distribution by Brain Lobe (L1)",
      x = "Brain Lobe",
      y = "Number of Terminal Sites"
    ) +
    theme_classic(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("No L1 data available in Terminal_Sites sheet\n")
}
```

## 6.2 Terminal Distribution by L3 Region

```{r terminal-l3}
# Count terminals by L3 (if available)
terminal_l3 <- df_terminals %>%
  filter(!is.na(Terminal_L3)) %>%
  group_by(Terminal_L3) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  head(20)

if (nrow(terminal_l3) > 0) {
  ggplot(terminal_l3, aes(x = reorder(Terminal_L3, -n), y = n)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = "Top 20 Terminal Regions (L3)",
      x = "Brain Region (L3)",
      y = "Number of Terminal Sites"
    ) +
    theme_classic(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("No L3 data available in Terminal_Sites sheet\n")
}
```

# 7. Network Analysis

## 7.1 Type-to-Type Connection Matrix

```{r type-connection}
# Calculate connection strength between types
# For each type pair, calculate mean projection from type A to regions containing type B name

types <- unique(df_proj_l3_ipsi$Neuron_Type)
n_types <- length(types)
conn_matrix <- matrix(0, nrow = n_types, ncol = n_types)
rownames(conn_matrix) <- types
colnames(conn_matrix) <- types

for (i in 1:n_types) {
  for (j in 1:n_types) {
    source_neurons <- df_proj_l3_ipsi$NeuronID[df_proj_l3_ipsi$Neuron_Type == types[i]]
    target_regions <- grep(types[j], colnames(df_proj_l3_ipsi), value = TRUE, ignore.case = TRUE)
    
    if (length(source_neurons) > 0 && length(target_regions) > 0) {
      values <- as.matrix(df_proj_l3_ipsi[df_proj_l3_ipsi$NeuronID %in% source_neurons, target_regions])
      conn_matrix[i, j] <- mean(values, na.rm = TRUE)
    }
  }
}

# Plot connection matrix
ht_conn <- Heatmap(
  conn_matrix,
  name = "Connection\nStrength",
  col = colorRamp2(c(0, max(conn_matrix)/2, max(conn_matrix)), 
                   c("white", "orange", "red")),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  column_names_side = "top",
  row_names_gp = gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = gpar(fontsize = 12, fontface = "bold"),
  width = unit(6, "cm"),
  height = unit(6, "cm"),
  border = TRUE,
  rect_gp = gpar(col = "gray", lwd = 1)
)

draw(ht_conn)
```

## 7.2 Network Visualization

```{r network-viz}
# Create network graph
threshold <- quantile(conn_matrix, 0.3)
adj_matrix <- conn_matrix > threshold

net <- graph_from_adjacency_matrix(
  adj_matrix,
  mode = "directed",
  weighted = TRUE,
  diag = TRUE
)

# Set node colors
V(net)$color <- type_colors[V(net)$name]
V(net)$size <- 30

# Plot
par(mar = c(0, 0, 0, 0))
plot(net,
     vertex.label.cex = 1.2,
     vertex.label.font = 2,
     edge.width = E(net)$weight * 2,
     edge.arrow.size = 0.6,
     edge.curved = 0.2,
     main = "Type-to-Type Connection Network")
```

# 8. Soma Hierarchy Analysis

## 8.1 Soma Location Distribution

```{r soma-hierarchy}
# Analyze soma hierarchy levels
if ("Soma_L1" %in% colnames(df_hierarchy)) {
  soma_l1 <- df_hierarchy %>%
    group_by(Soma_L1) %>%
    summarise(n = n(), .groups = "drop")
  
  ggplot(soma_l1, aes(x = reorder(Soma_L1, -n), y = n)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = "Soma Distribution by Brain Lobe (L1)",
      x = "Brain Lobe",
      y = "Number of Neurons"
    ) +
    theme_classic(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("Soma hierarchy data structure:\n")
  print(colnames(df_hierarchy))
}
```

# 9. Summary Statistics

```{r summary-stats}
# Overall summary
cat("=" ,rep("=", 60), "\n", sep="")
cat("SUMMARY STATISTICS\n")
cat("=" ,rep("=", 60), "\n\n", sep="")

cat("Total neurons:", nrow(df_summary), "\n")
cat("Neuron types:", paste(unique(df_summary$Neuron_Type), collapse = ", "), "\n")
cat("L3 regions (ipsilateral):", ncol(df_proj_l3_ipsi) - 2, "\n")
cat("Terminal sites:", nrow(df_terminals), "\n\n")

cat("Morphological characteristics:\n")
cat("  Mean axon length:", round(mean(df_summary$Total_Length), 2), "μm\n")
cat("  Mean terminal count:", round(mean(df_summary$Terminal_Count), 2), "\n")
cat("  Mean laterality index:", round(mean(df_summary$Laterality_Index), 4), "\n")
```

# 10. Save Results

```{r save-results, eval=FALSE}
# Create output directory
output_dir <- "analysis_output"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# Save type statistics
write.csv(type_stats, file.path(output_dir, "type_statistics.csv"), row.names = FALSE)

# Save type-region profile
write.csv(type_region_profile, file.path(output_dir, "type_region_profile.csv"))

# Save type-type connection
write.csv(conn_matrix, file.path(output_dir, "type_type_connection.csv"))

# Save neuron info
write.csv(neuron_info, file.path(output_dir, "neuron_info.csv"), row.names = FALSE)

cat("Results saved to:", output_dir, "\n")
```

---

**Analysis completed!** 

This pipeline provides comprehensive single-neuron projection analysis following Gao et al., 2023 Neuron methodology.
